---
title: "Human input and approvals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Human input and approvals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Some decisions belong to a person. Deputy lets the model ask questions while it
works, and gives you three ways to hold a tool call until someone approves it:

- a hook that asks straight away and waits for the answer;
- a durable approval that stops the run, writes it to disk, and resumes later,
  even in a new R session;
- a Shiny module that shows the pending call and lets the reviewer edit its
  inputs before approving.

## Let the model ask questions

`tools_interactive()` returns an `ask_user` tool. The model sends one to four
multiple-choice questions, and your handler returns the answers as a named list
that maps each question to the chosen label:

```{r}
library(deputy)

agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  tools = c(
    tools_file(),
    tools_interactive(
      callback = function(questions, context) {
        answers <- lapply(questions, function(q) q$options[[1]]$label)
        setNames(answers, vapply(questions, `[[`, "", "question"))
      }
    )
  )
)
```

That handler always picks the first option, which is only useful for testing.
At the R console, leave out `callback` and the tool asks with `readline()`,
where the person can also type a free answer instead of picking an option.

Each question has `question`, a short `header`, two to four `options` (each
with a `label` and `description`), and `multiSelect`. For a multi-select
question, join the chosen labels with `", "`.

A handler can answer in three ways:

- return the answers directly, as above;
- return a promise that resolves to the answers, when it has to wait for a
  person without blocking R, as in Shiny;
- return `AskUserDeferred()`, which tells the model the questions are on
  screen and it should end its turn. The answers then arrive as the person's
  next message, like any other chat message.

A subagent's handler can return answers or a promise, but not
`AskUserDeferred()`, because the answers have to arrive before the subagent's
run ends.

The `context` argument of `tools_interactive()` passes routing values to the
handler, such as the IDs your app uses to find the right browser session. Each
call to `tools_interactive()` creates a separate tool, so agents in different
Shiny sessions never share a handler. (`set_ask_user_callback()` sets one
handler for the whole R process; it exists for single-agent scripts and
shouldn't be used in apps.)

## Ask before a tool runs

Deputy ships a small recipe, `approval-gates.R`, that turns a `PreToolUse`
hook into an approval prompt. Load it:

```{r approval-recipe-load, eval = TRUE}
source(system.file("examples", "approval-gates.R", package = "deputy"))
```

`approval_gate()` asks, through an `ask_user` handler, whether to run each
matching tool call (by default `run_bash` and `write_file`). The question shows
the tool name and every argument with its declared type. Only the answer
`"Approve"` lets the call run; any other answer, a cancelled prompt, or an
error in the handler denies it.

Here it is with a handler that always declines:

```{r approval-simple-demo, eval = TRUE}
decline <- function(questions, context) {
  setNames(list("Deny"), questions[[1]]$question)
}
gate <- approval_gate(callback = decline)
S7::prop(gate, "callback")(
  "write_file",
  list(path = "report.txt", content = "Draft"),
  context = list(run_id = "example-run")
)$permission
```

Add the gate to an agent whose permissions already allow the tools it guards.
A hook can deny a call the policy allows, but it can't allow one the policy
denies:

```{r}
agent$add_hook(approval_gate(callback = my_approval_handler))
```

Add approval hooks before any other hook for the same event, including the
built-in ones. The first hook to return a result decides, so an earlier hook
that returns a result would skip the approval. The gate returns `NULL` after
an approval, so hooks added after it still run.

The recipe also has `approval_after_install()`, which asks before
`push_changes` only if an `install_dependency` tool succeeded earlier in the
same run. It shows how hooks can share state keyed by `context$run_id`:

```{r approval-stateful-demo, eval = TRUE}
hooks <- approval_after_install(callback = decline)
run <- list(run_id = "installation-run")
# PostToolUse: record a successful installation.
S7::prop(hooks[[1]], "callback")(
  "install_dependency", list(installed = TRUE), NULL, run
)
# PreToolUse: this run installed something, so pushing needs approval.
S7::prop(hooks[[2]], "callback")("push_changes", list(remote = "origin"), run)$permission
# Another run installed nothing, so the hook has no opinion.
S7::prop(hooks[[2]], "callback")(
  "push_changes", list(remote = "origin"), list(run_id = "another-run")
)
```

Add all three hooks with a loop:
`for (hook in approval_after_install(my_handler)) agent$add_hook(hook)`.

A gate hook blocks the run while it waits for an answer. That is fine at the
console. When the answer may take hours, or the app can't block, use a durable
approval instead.

## Pause a run until someone decides

A durable approval stops the run before the tool executes and saves everything
needed to continue. Nothing is left waiting in memory: the approval can be
decided later, by another process, after R has restarted.

It needs three things:

- an existing, private directory passed to `Agent$new(approval_dir = )`;
- a permission callback that returns `PermissionResultPending(reason)` for the
  calls that need a decision;
- tools registered with `convert = FALSE`, so they receive the raw JSON
  arguments and can check them. The reviewer may edit those arguments, so the
  tool must validate its own inputs.

```{r}
approval_dir <- file.path(tempdir(), "approvals")
dir.create(approval_dir, showWarnings = FALSE)

export_report <- ellmer::tool(
  function(name) {
    if (!is.character(name) || length(name) != 1L || is.na(name)) {
      cli::cli_abort("{.arg name} must be a single string.")
    }
    paste("Exported", name)
  },
  name = "export_report",
  description = "Export a named report.",
  arguments = list(name = ellmer::type_string()),
  convert = FALSE,
  annotations = ellmer::tool_annotations(
    read_only_hint = FALSE,
    destructive_hint = FALSE,
    open_world_hint = FALSE
  )
)

make_agent <- function() {
  Agent$new(
    chat = ellmer::chat("openai/gpt-6-luna"),
    tools = list(export_report),
    permissions = Permissions(
      can_use_tool = function(tool_name, tool_input, context) {
        PermissionResultPending("A person must approve every export.")
      }
    ),
    approval_dir = approval_dir,
    working_dir = approval_dir,
    session_id = "report-session",
    agent_id = "report-agent"
  )
}

agent <- make_agent()
result <- agent$run_sync("Export the annual report.")
result$stop_reason
#> [1] "approval_pending"
```

The run stops with `"approval_pending"`. `agent$pending_approval()` describes
the call, and its `source$path` is what you store to find it again:

```{r}
pending <- agent$pending_approval()
pending$request$tool_name
pending$request$tool_input
approval_path <- pending$source$path
```

`approval_read(approval_path)` returns the same description without a chat or
an agent, so a review screen or another process can show it. To decide, build
an agent with the same tools, permission callback, IDs, working directory and
approval directory, and resume:

```{r}
resumed <- make_agent()
result <- resumed$resume_approval(
  approval_path,
  "approve",
  tool_input = list(name = "annual report (reviewed)")
)
# or: resumed$resume_approval(approval_path, "deny")
```

`tool_input` is optional; leave it out to approve the call exactly as the model
proposed it. When a mismatched agent tries to resume, for example with a
different tool definition or working directory, Deputy refuses. The permission
callback runs again on resume, and the approval can't be used to run a tool the
policy doesn't allow.

### What to expect

If the model asked for several tools at once and one of them needs a decision,
the calls before it keep their results and the calls after it don't run. On
resume, the model is told those later calls weren't executed, so it can ask for
them again. Calls that already ran are never repeated.

With `approval_dir` set, a usage limit reached when the model asks for a tool
also pauses the run as a pending approval, instead of ending it. To let it
continue with a bigger budget, pass
`usage_limits = UsageLimits(...)` to `resume_approval()`. The new limits can't
exceed the agent's own limits, and usage from before the pause still counts.

Each approval can be decided once. If R crashes while an approved tool is
running, the record stays in the `"executing"` state and can't be resumed,
because Deputy can't tell whether the tool finished. Check the tool's own
records (did the report get exported?) and decide what to do. Approvals are
not transactions, and a power cut can lose the latest state.

The approval directory holds the conversation and tool arguments, so keep it
private, and link each approval to its owner in your own database. Each saved
state is limited to 25 MiB. Subagents can't pause for durable approval; see
[Trusted mini-agents](trusted-mini-agents.html) for a pattern that works
around this.

## Review a pending call in Shiny

`approval_review_ui()` and `approval_review_server()` show a pending approval
as a form: one row per argument, with its declared type and description, and
the value the model proposed. The reviewer can edit simple values (text,
numbers, logicals and choices from a list) and then approve or deny.

```{r}
ui <- bslib::page_fluid(approval_review_ui("review"))

server <- function(input, output, session) {
  agent <- make_agent()
  approval_review_server("review", agent)
}
```

By default the module resumes the approval in the Shiny process, which blocks
it while the model continues. In a deployed app, pass `decide`, a function that
runs the continuation elsewhere (for example with mirai) and returns a promise.

`tool_input_review()` builds the same table as a data frame, for use in your
own UI, a hook, or a permission callback:

```{r, eval = TRUE}
arguments <- ellmer::type_object(
  city = ellmer::type_string("City to forecast"),
  days = ellmer::type_integer("Number of days")
)
tool_input_review(list(city = "Oslo", days = 3L), arguments)
```

Permission callbacks and hooks receive a tool's declared arguments as
`context$tool_arguments`, so a review can be built wherever a call is
intercepted. [Trusted mini-agents](trusted-mini-agents.html) puts these pieces
together in a complete app.
