---
title: "Claude Agent SDK Compatibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Claude Agent SDK Compatibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

deputy keeps a provider-agnostic R runtime at its core and layers an opt-in
Anthropic-compatible facade on top. This vignette shows what is covered by the
compatibility layer, how the Anthropic-style entrypoints map to deputy, and
where the current boundary remains intentionally narrower than Anthropic's full
ecosystem surface.

## Compatibility Matrix

| Surface | Status | deputy mapping |
|--------|--------|----------------|
| `claude_sdk_query()` / `query()` | Covered | `claude_sdk_query()` and `ClaudeSDKClient$query()` |
| Session ids, resume, fork | Covered | Persisted snapshots plus `resume()` / CLI flags |
| Permission modes | Covered | `default`, `acceptEdits`, `readonly`, `plan`, `bypassPermissions` |
| Hook events | Covered | Existing hook system plus `Notification` |
| Claude settings | Covered | `CLAUDE.md`, `.claude/skills`, `.claude/commands`, `.claude/agents`, tool policy keys |
| Anthropic-style built-in tool names | Covered | `Read`, `Write`, `Edit`, `MultiEdit`, `Glob`, `Grep`, `LS`, `TodoRead`, `TodoWrite`, `WebFetch`, `WebSearch`, `Task` |
| Native deputy runtime | Canonical | `Agent`, `LeadAgent`, `Permissions`, `HookMatcher` |
| Plugin discovery / marketplace execution | Not in this wave | Track separately from the core compat facade |
| MCP resource subscriptions | Not in this wave | deputy exposes MCP tools, not Anthropic-specific resource flows |

## Entry Points

Use `claude_sdk_options()` to express Anthropic-shaped options and translate
them into deputy internals:

```{r, eval = FALSE}
library(deputy)

options <- claude_sdk_options(
  chat = ellmer::chat_anthropic(model = "claude-sonnet-4-5-20250929"),
  setting_sources = c("project", "user"),
  permission_mode = "plan",
  allowed_tools = c("Read", "Grep", "LS"),
  max_turns = 15
)

result <- claude_sdk_query("Summarize the package structure", options = options)
result$session_id
```

For a stateful client:

```{r, eval = FALSE}
client <- ClaudeSDKClient$new(options)
client$query("Inspect the R/ directory")
```

## Sessions, Resume, and Fork

The compatibility layer persists snapshots to:

```{r, eval = FALSE}
tools::R_user_dir("deputy", "cache")
```

Snapshots are enabled by default for the Claude-compatible APIs. Each completed
turn writes a new snapshot, and the latest snapshot path is attached to
`AgentResult$snapshot_path`.

```{r, eval = FALSE}
result <- client$query("Create a high-level summary")

# Restore the latest snapshot for a session id
client$resume(result$session_id)

# Restore the newest snapshot at or before a point in time
client$resume(result$session_id, at = "2026-03-07 10:30:00")

# Clone the restored state into a new session id
client$resume(result$session_id, fork = TRUE)
```

The CLI exposes the same behavior:

```bash
deputy --permission-mode plan --persist-session
deputy --resume-session-id abc123
deputy --resume-session-id abc123 --resume-session-at "2026-03-07 10:30:00"
deputy --resume-session-id abc123 --fork-session
```

## Permissions and Hooks

Anthropic's planning behavior maps to deputy's `plan` mode:

```{r, eval = FALSE}
permissions_plan()
```

`plan` allows only tools annotated with `read_only_hint = TRUE` plus the
approval tool named by `permission_prompt_tool_name` (default:
`"AskUserQuestion"`). Write, execution, and unannotated tools are denied by
default.

Informational runtime events now use the existing hook system via
`Notification`:

```{r, eval = FALSE}
hook <- HookMatcher$new(
  event = "Notification",
  callback = function(message, context) {
    cli::cli_alert_info("[{context$code}] {message}")
    NULL
  }
)
```

deputy emits `Notification` for cases such as session restore or fork notices,
permission-denied guidance, compaction fallbacks, and cost warnings.

## Settings, Agents, and Tool Aliases

`claude_settings_load()` and `claude_settings_apply()` now understand:

- `CLAUDE.md` memory blocks
- `.claude/skills`
- `.claude/commands`
- `.claude/agents`
- Tool policy keys such as `allowedTools`, `disallowedTools`, and `permissionPromptToolName`

Custom agents from `.claude/agents` are converted to `AgentDefinition`
objects and registered automatically when you use `LeadAgent` or
`ClaudeSDKClient`.

Anthropic-style tool names resolve through the compatibility layer rather than
changing deputy's native tool names:

```{r, eval = FALSE}
compat_tools <- claude_sdk_options(
  custom_tools = list(),
  allowed_tools = c("Read", "Edit", "TodoWrite")
)
```

This keeps `read_file`, `edit_file`, `todo_write`, and the rest of deputy's
snake_case tool surface intact for native R users.

## Design Boundary

deputy aims for behavioral compatibility, not a literal language-port of the
Node or Python SDKs. The runtime model is intentionally R-native:

- R6 objects instead of Python or TypeScript classes
- `coro` generators for streaming
- Base lists and data frames for snapshots and session indexes
- Existing deputy hooks and permissions reused by the compat facade

That boundary keeps the Anthropic surface available without making the rest of
the package Anthropic-specific.
