---
title: "Running R code"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running R code}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Letting the model write and run R code is the most flexible tool you can give
an agent, and the most dangerous. Deputy offers four ways to do it, which
differ in two respects: whether R state survives between calls, and whether
the code is confined by an OS sandbox.

| Option | State between calls | Confinement |
|---|---|---|
| `tool_run_r_code` | none; each call starts a new R process | your user account's access |
| `RSession` | variables, packages and plots persist | your user account's access |
| mcp-repl, through `mcp_repl_connection()` or `tools_mcp_repl()` | persists | OS sandbox |
| MCP Console, through `mcp_console_connection()` | persists; R, Python and SQL share data | OS sandbox |

Deputy calls the first two *trusted-code tools*. A separate process protects
your R session from crashes and runaway loops, but the code can still read,
write and send anything your account can. Use them when you trust the model,
the prompt and every input the agent reads, for example your own analysis on
your own data. Otherwise use a sandbox.

## One-off R code

`tool_run_r_code` runs each snippet in a fresh R process started with callr,
with a 30-second timeout, and returns the printed output and result. Enable
it with `r_code = TRUE`:

```{r}
library(deputy)

agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  tools = c(tools_data(), list(tool_run_r_code)),
  permissions = Permissions(r_code = TRUE, file_write = FALSE)
)
agent$run_sync("How many rows of mtcars have more than 4 cylinders?")
```

`tool_run_bash` does the same for shell commands and needs `bash = TRUE`.
Nothing carries over between calls, so the model has to reload data each time.

## A persistent R session

`RSession` gives one agent its own R worker process. Variables and loaded
packages persist between calls, so the model can load data once and work with
it over many steps. Plots come back to the model as images.

```{r}
agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  permissions = Permissions(r_code = TRUE, file_write = FALSE)
)
worker <- RSession$new(agent)
agent$register_tools(worker$tools())

agent$run_sync("Load mtcars, fit mpg ~ wt, and plot the residuals.")
agent$run_sync("Now add hp to the model. Did the fit improve?")

worker$close()
```

The worker starts on the first call. Calls run one at a time in order, each
with a 30-second timeout by default, and each starts in the agent's working
directory. `$cancel()`, or a timeout, stops the worker and discards its
variables; the next call starts a fresh worker and says so. Call `$close()`
when the conversation ends, for example from `session$onSessionEnded()` in
Shiny.

A saved conversation keeps the code, output and plots, but not the worker's
variables. After a restart the model has to recreate them.

Static plots from base graphics, ggplot2, grid and patchwork are captured.
htmlwidgets and other interactive output are reported as unsupported.

### Call your tools from R code

An R session can let the model's code call some of the agent's other tools, so
it can fetch data with a tool and then analyse the result in R:

```{r}
worker <- RSession$new(agent, tools = "fetch_measurements")
agent$register_tools(worker$tools())
```

Inside the session, the code calls `tools$fetch_measurements(...)` and gets
the tool's return value as an ordinary R object. Each of these calls goes
through the agent's permissions, hooks and usage limits like any other tool
call. The selected tools must be registered with `convert = FALSE` and check
their own arguments, and they can't pause for a durable approval.

## Sandboxed R with mcp-repl

[mcp-repl](https://github.com/posit-dev/mcp-repl) runs an R interpreter as an
MCP server inside an OS sandbox, so model-written code can only read and write
what the sandbox allows. Configure the server in your mcptools config with an
explicit `--sandbox` mode:

```json
{
  "mcpServers": {
    "r": {
      "command": "/absolute/path/to/mcp-repl",
      "args": ["--sandbox", "workspace-write", "--interpreter", "r"]
    }
  }
}
```

Then connect an agent to its own REPL:

```{r}
agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  permissions = Permissions(web = FALSE)
)
repl <- mcp_repl_connection(
  config = "~/.config/mcptools/config.json",
  agent = agent,
  server = "r",
  sandbox = "workspace-write"
)
agent$register_tools(repl$tools())
# ...
repl$close()
```

Deputy checks that the server's final `--sandbox` argument matches the mode you
ask for, and refuses to start otherwise. It accepts `"read-only"` and
`"workspace-write"`; a configuration that leaves the mode out or uses any other
mode is rejected. mcp-repl enforces the sandbox and refuses to start
on a system where it can't. Its
[sandbox documentation](https://github.com/posit-dev/mcp-repl/blob/main/docs/sandbox.md)
describes what each mode allows on each platform.

Each connection has its own REPL and belongs to one agent. Long computations
return a "busy" result after about three seconds; the model then polls with an
empty call to collect the output. `mcp_repl_control(repl, "interrupt")`
interrupts the running code, and `mcp_repl_control(repl, "reset")` starts a
fresh interpreter. Both return promises; check the result to see whether the
REPL accepted the request.

`tools_mcp_repl()` loads the same tools without a connection object, for simple
scripts:

```{r}
agent$register_tools(tools_mcp_repl(server = "r", sandbox = "workspace-write"))
```

Deputy works with mcptools 1.0.2 and 1.0.3 and refuses other versions until
they have been tested. It has been tested with mcp-repl 0.3.0.

## R, Python and SQL with MCP Console

[MCP Console](https://github.com/t-kalinowski/mcp-console) is a sandboxed
workbench that grew out of mcp-repl. Its single `send` tool runs R, Python and
DuckDB SQL cells in one worker, so a data frame made in R can be queried in SQL
and read from Python in the same conversation.

```{r}
agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  permissions = Permissions(bash = TRUE, web = TRUE)
)
console <- mcp_console_connection(
  agent,
  command = Sys.getenv("DEPUTY_MCP_CONSOLE_BIN"),
  dependencies = "allow"
)
agent$register_tools(console$tools())
# ...
console$close()
```

Install MCP Console 0.0.4 yourself (for example with
`uv tool install mcp-console==0.0.4`) and pass the path to the executable.
Deputy checks the version before starting it.

In its sandbox, code can read files but can't reach the network, and it can
write only to private temporary storage and to paths you allow. Deputy keeps it
that way: it refuses options that would turn the sandbox off, widen file
access, add a proxy, or run the code somewhere else. It accepts
`--writable-root PATH` and the `-c extends=:workspace` (write in the agent's
working directory), `-c extends=:read-only` and
`-c sandbox.network=restricted` presets. A project config file,
`.agents/console/config.yaml`, can change the sandbox policy, so Deputy won't
start when one exists unless you have reviewed it and pass
`project_config = TRUE`. After starting, Deputy checks the sandbox the server
reports and disconnects unless it is the native sandbox with restricted
networking.

`send` runs code, so Deputy treats it like a shell command: the agent needs
`bash = TRUE`, plus `web = TRUE` because the server doesn't annotate its tools.
Read-only and plan modes deny it.

MCP Console can also install the packages that code asks for, and that happens
outside the sandbox. It does this when a call lists `requirements`, and also on
its own when code loads a missing package. Since Deputy can't switch the second
case off, the default `dependencies = "deny"` refuses to connect to a server
that offers installation. With `dependencies = "allow"`, a call that lists
`requirements` also needs `install_packages = TRUE` in the agent's permissions.
Installers use their usual caches in your home directory; set variables such
as `UV_CACHE_DIR` through `env` to move them.

Calls wait up to 2.5 seconds. Longer work keeps running and returns
`[running; poll with an empty send]`, and the model collects the output with an
empty `send`. `mcp_console_control(console, "interrupt")` interrupts a cell and
keeps the session's state; `mcp_console_control(console, "restart")` discards
all R, Python and SQL state. If the restarted worker isn't ready within 2.5
seconds, Deputy closes the connection and you need to start a new one.

MCP Console records every call, with its output and plots, under
`.agents/console/sessions/` in the agent's working directory. Recordings aren't
redacted and aren't deleted automatically; `console$status()$execution` gives
the path.
