---
title: "Conversations and context"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Conversations and context}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

An agent keeps its conversation between runs, like an ellmer chat, so a
follow-up task can build on the last one. Long conversations eventually outgrow
the model's context window, and big tool results fill it faster. This article
covers what the agent remembers, how Deputy keeps the context small, and how to
save a conversation and pick it up later.

## Two views of the conversation

Deputy keeps two lists of turns:

- `agent$get_turns()` is the whole conversation, every message and tool call.
  This is what you show a user or store.
- `agent$get_context_turns()` is what the model receives on its next request.
  After compaction it is shorter: older turns are replaced by a summary.

Until the first compaction the two are the same. Code that measures the
model's context should use `get_context_turns()`. `set_turns()` replaces the
conversation with the turns you give it and discards any summary.

## Compaction

Every agent has a `ContextPolicy`. By default, before a request whose estimated
size exceeds 32,000 tokens, Deputy asks the model to summarise the older turns,
adds the summary to the system prompt, and keeps only the recent turns in the
model's context. The check runs at the start of each run and between rounds of
tool calls, after all pending tool results are in.

```{r}
library(deputy)

agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  tools = tools_file(),
  context_policy = ContextPolicy(max_tokens = 64000)
)
```

`ContextPolicy(max_tokens = NULL)` turns automatic compaction off. The summary
request is an ordinary model request: it counts against the run's limits and
can be interrupted with the run. Because the summary lives in the system
prompt, replacing the prompt with `set_system_prompt()` after a compaction
drops the summary, unless your new prompt includes it.

If the summary request fails, the run stops by default and the context is left
as it was. `fallback = "text"` instead falls back to a
truncated copy of the old turns and emits a `Notification` with code
`"compact_fallback"`. To send summaries to a second provider when the first is
down, list it in `summary_fallback_chats`:

```{r}
policy <- ContextPolicy(
  max_tokens = 64000,
  summary_fallback_chats = list(ellmer::chat("anthropic/claude-sonnet-5"))
)
```

Summary fallbacks never take over the task itself, and they receive only the
turns to summarise, without tools. [Fallbacks, tracing and
evaluation](runtime-integration.html) covers fallbacks for the task.

### Watch or control compaction

`agent$last_compaction()` reports the most recent compaction: how the summary
was made (`"llm"`, `"text"`, `"hook"` or `"custom"`), how many turns were
replaced, what it cost, and each attempt.

Two hook events surround it. A `PreCompact` hook can cancel the compaction or
supply its own summary with `HookResultPreCompact()`; a `PostCompact` hook sees
the result:

```{r}
agent$add_hook(HookMatcher(
  event = "PostCompact",
  callback = function(result, context) {
    cli::cli_inform("Summarised {result$turns_compacted} turns ({result$method}).")
    NULL
  }
))
```

To compact on demand, call `agent$compact()`. It runs straight away, outside
any run. `keep_last` sets how many recent turns to keep, and `summary` lets you
write the summary yourself.

### Clear old tool results

Tool results are often the bulk of a long conversation, and old ones are rarely
needed word for word. `agent$microcompact()` replaces every tool result before
the last few turns (two by default) with a short marker, without a model
request:

```{r}
agent$microcompact(keep_last = 6, keep_tools = "read_csv")
```

`keep_tools` names tools whose results are never cleared. Like compaction, this
only changes the model's context: `get_turns()` and saved conversations keep
the original results.

## Large tool results

A single tool result can be bigger than is sensible to send to the model. When
one exceeds 64 KiB, Deputy saves it to disk and gives the model a preview and a
`deputy://tool-result/...` reference instead. Deputy also gives the model a
`deputy_read_tool_result` tool, so it can read the stored result a chunk at a
time if it needs more.

Your code gets the full value back with `resolve_tool_result()`:

```{r}
value <- agent$resolve_tool_result("deputy://tool-result/...")
```

`ContextPolicy()` sets the size limit (`max_tool_result_bytes`), limits on
images returned by tools, and where results are stored (`offload_dir`). By
default they go in Deputy's user cache directory.

## Save and restore a conversation

`save_session()` writes the conversation to an RDS file, and `load_session()`
restores it into an agent:

```{r}
agent$save_session("review.rds")

later <- Agent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  tools = tools_file(),
  permissions = permissions_readonly()
)
later$load_session("review.rds")
later$run_sync("Pick up where we left off.")
```

The file holds both views of the conversation, the system prompt and any
compaction summary, copies of stored tool results, the run context, and file
checkpoint history if checkpointing is on. It holds no tools, permissions or
hooks: the receiving agent's own configuration applies, so loading a
conversation never grants an agent more than it was created with.

A saved conversation is a snapshot, not a database. Where you keep it, who may
load it, and how you handle branches are up to your application.

## In a Shiny app

shinychat's conversation history works with compaction. It stores and
restores the whole conversation (`get_turns()`), while the model keeps
receiving the compacted context. See [Shiny chat](example-shiny-chat.html) for
an app that tells users when older messages have been summarised.
