---
title: "Briefing and inspecting subagents"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Briefing and inspecting subagents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

[Subagents](multi-agent.html) covers defining subagents and letting a lead
delegate to them. This article is about everything around a delegation: what
a subagent is given, what it hands back, and how to show its work to the people
using your app without showing them anything they shouldn't see.

## Write a structured brief

A plain string is a fine task. When the task needs more structure,
`DelegationInput()` separates the instruction from its constraints, the
evidence to use, the expected deliverable and when to stop:

```{r}
library(deputy)

brief <- DelegationInput(
  "Assess the reported concentration.",
  constraints = "Keep the reported units.",
  evidence = list(list(source_id = "assay-17", revision = "r3")),
  deliverable = "A conclusion that cites the source revision.",
  stop_conditions = "State the uncertainty if the method is missing."
)
```

`stop_conditions` are instructions to the model. They don't set limits;
`UsageLimits` and permissions do that.

The `evidence` entries refer to documents that your application supplies when
it creates the lead, as `delegation_sources`. Each source is a text record
with an ID, a revision, the owner and conversation it belongs to, and
optionally the subagents allowed to see it:

```{r}
lead <- LeadAgent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  sub_agents = list(
    agent_definition("reviewer", "Reviews evidence", "Keep units explicit.")
  ),
  delegation_scope = list(owner_id = "team-a", conversation_id = "review-17"),
  delegation_sources = list(list(
    source_id = "assay-17",
    revision = "r3",
    owner_id = "team-a",
    conversation_id = "review-17",
    text = "Reported concentration: 12 mg/L.",
    allowed_agents = "reviewer"
  ))
)

lead$parallel_delegate(list(reviewer = brief))
```

Deputy puts the text of each cited source into the subagent's first message.
A reference must match the lead's scope and the exact revision, and the source
must allow that subagent; otherwise the delegation is rejected before any model
request, and `list_subagents()$input_error` says why. Leave out
`allowed_agents` to allow every subagent; `character()` allows none.

The model can write the same kind of brief when it calls `delegate_to_agent`
itself, and can pick sources by ID and revision, but it can't supply source
text or change the scope. Your application decides what the sources contain
and who may use them. They are a snapshot: to update a source, create a new
lead.

A delegation whose prepared prompt, first message or manifest is larger than
64 KiB is rejected before any request. `LeadAgent$new(delegation_max_bytes = )`
changes the limit.

### See what a subagent was given

`lead$get_subagent_contexts()` shows the prepared input for each delegation: a
`DelegationManifest` with the system prompt, the first message, the resolved
sources and their digests, the model, the tool names and the sizes. With
`view = "current"` it shows the subagent's working context instead.

```{r}
id <- lead$list_subagents()$delegation_id[[1]]
lead$get_subagent_contexts(id)
lead$get_subagent_contexts(id, view = "current")
lead$get_subagent_contexts(id, redact = TRUE) # metadata only, no text
```

## Give each subagent its own resources

A subagent's tools are the objects in its definition, so every delegation to
that definition shares them, along with anything their closures hold, such as
a database connection. `DelegationPolicy()` controls this with
`resource_mode`:

- `"shared"`, the default, lets every delegation use the definition's tools as
  they are. Deputy never closes them.
- `"exclusive"` also shares them, but while one delegation holds the resource
  named by `resource_key`, another delegation that needs it fails instead of
  sharing it. This only coordinates delegations within one R process.
- `"owned"` builds fresh tools for every delegation with a factory function,
  and cleans them up afterwards.

For example, to give each delegation its own R session:

```{r}
policy <- DelegationPolicy(
  resource_mode = "owned",
  resources = function(agent, definition, context) {
    worker <- RSession$new(agent)
    DelegationResources(worker$tools(), cleanup = worker$close)
  }
)

lead <- LeadAgent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  permissions = permissions_full(), # the R session runs with your account's access
  sub_agents = list(
    agent_definition("analyst", "Analyses data", "Check the data before modelling.")
  ),
  delegation_policy = policy
)
```

The factory gets the new subagent before its first request. It should only
build tools, not run anything. If it fails, it must clean up whatever it
created; once it returns, Deputy calls `cleanup` exactly once, after the
delegation finishes or fails. Cleanup errors appear in
`list_subagents()$cleanup_error`. In owned mode, definitions can't carry tools
of their own, and a definition that names `mcp_servers` needs a factory to
create those connections.

Lead hooks for `PreToolUse`, `PostToolUse` and `PostToolUseFailure` always run
for subagent tool calls. List other events in `observers` to forward those too.

### Subagents that ask questions

A subagent that should be able to ask the user something needs an `ask_user`
tool (include `tools_interactive()` in its definition) and a handler on the
policy. The handler receives the subagent's identifiers, so your app can route
the question to the right place:

```{r}
policy <- DelegationPolicy(
  human_input = function(questions, context) {
    ask_in_app(
      questions,
      owner = context$scope$owner_id,
      delegation_id = context$delegation_id
    )
  }
)
```

`ask_in_app()` stands in for your own code. Without a handler, a subagent with
`ask_user` fails before it starts. An answer from a person is just an answer:
it doesn't approve any tool call.

## What a delegation returns

When a subagent finishes, the `delegate_to_agent` tool returns a compact JSON
summary, a `DelegationOutcome`, rather than the subagent's whole conversation.
Its `runtime` part holds facts Deputy recorded: status, stop reason and
identifiers. Its `answer` and optional `claims` (missing evidence, unresolved
work) are what the model wrote, and nobody has checked them. `parallel_delegate()`
returns the same outcomes as `batch$outcomes`.

Answers are capped at 8 KiB. A longer answer is stored like a [large tool
result](conversations.html#large-tool-results), and the outcome includes a
reference to the full text.

## Show subagent work to users

The subagent's full conversation stays out of the lead's context, but you may
want to show it to the user. `list_subagents()`, `get_subagent_results()` and
`get_subagent_messages()` return everything to your code without any checks.
The inspection methods below go through a `DelegationDisclosure`, which your
app configures to decide who may see what:

```{r}
disclosure <- DelegationDisclosure(
  authorize = function(requester, scope) {
    identical(requester$user_id, scope$owner_id)
  },
  redact = function(view, requester) {
    view # remove whatever this user shouldn't see
  }
)

lead <- LeadAgent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  sub_agents = definitions,
  delegation_scope = list(owner_id = "user-1", conversation_id = "chat-1"),
  delegation_disclosure = disclosure
)

views <- lead$inspect_subagents(current_user, transcript = TRUE)
views[[1]]$outcome$runtime
views[[1]]$turns
```

`authorize` must return exactly `TRUE` to allow access, and runs before Deputy
looks anything up, so a denied user can't tell whether a delegation exists.
`redact` sees each view before it's returned. The default disclosure denies
everything. `requester` is whatever your app uses to identify the signed-in
user; Deputy doesn't authenticate anyone.

`views[[1]]$turns` holds the subagent's conversation as ellmer turns, ready to
render. Treat the text as untrusted when you display it.

### Save and replay subagent history

`export_subagents()` returns finished subagent histories as a plain list you
can store, and `delegation_history()` turns a stored list back into views,
checking access again with the disclosure and scope you pass in:

```{r}
history <- lead$export_subagents(current_user)
saveRDS(history, "subagent-history.rds")

restored <- delegation_history(
  readRDS("subagent-history.rds"),
  current_user,
  disclosure,
  scope = scope_from_your_database
)
```

Take `scope` from your own records, not from the saved file. Replaying history
only displays it: nothing is run, resumed or approved.
`read_subagent_result()` reads an answer that was stored separately because it
was too long.

## Watch subagents live

`observe_subagents()` returns a subscription that reads subagent activity as it
happens, without slowing the subagents down:

```{r}
reader <- lead$observe_subagents(current_user)
reader$snapshot(transcript = FALSE)

promise <- lead$parallel_delegate_async(c(analyst = "Analyse the evidence."))

update <- reader$poll()
update$events
update$gaps
reader$close()
```

`poll()` returns the events since the last read. Events are kept in a buffer of
256 events or 1 MiB, whichever fills first. A reader that falls behind gets
`gaps` for the events it missed and can call `snapshot()` to catch up. To
reconnect later, pass the last `update$cursor` as `after`. Closing a reader
never stops a subagent. Every read checks the disclosure again, and `redact`
also receives each event (as `list(kind = "event", event = ...)`). Adjust the
buffer with `DelegationObservation()`.

To stop one subagent, call `lead$interrupt_subagent(delegation_id)`. That is a
control action, separate from viewing, so check that the user may cancel
before you call it.

## A subagent panel for Shiny

`subagent_chat_ui()` and `subagent_chat_server()` add a panel next to your
lead chat: a strip of activity cards, one per delegation, and the selected
subagent's conversation, rendered with shinychat's tool cards. The panel is
read-only; it has no input box and can't resume anything.

```{r}
ui <- bslib::page_fluid(
  bslib::layout_columns(
    shinychat::chat_ui("chat"),
    subagent_chat_ui("subagents")
  )
)

server <- function(input, output, session) {
  subagent_chat_server(
    "subagents",
    lead,
    requester = function() current_user(session),
    on_cancel = function(delegation_id, requester) {
      if (may_cancel(requester, delegation_id)) {
        lead$interrupt_subagent(delegation_id, "user_cancelled")
      }
    }
  )
}
```

Every poll checks the disclosure again, and denied access clears the panel. A
cancel button appears only if you pass `on_cancel`. To show saved history
instead of a live lead, pass `history`, together with `disclosure` and `scope`
from your own records. The panel needs shinychat 0.5.0 or later, bslib,
commonmark and xml2.

The package includes a demo that needs no API key. It runs two specialists
against a local test server and shows tool output, a plot, a failure,
cancellation, revoked access and replayed history:

```{r}
shiny::runApp(system.file("examples", "subagent-chats", package = "deputy"))
```
