---
title: "Example: Structured Data Extraction Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example: Structured Data Extraction Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(deputy)

docs_init <- getFromNamespace("deputy_docs_init", "deputy")
docs <- docs_init("example-extraction-pipeline", "verified-replay")
project <- docs$make_project()

setwd(project$root)
knitr::opts_knit$set(root.dir = project$root)

extract_json <- jsonlite::toJSON(
  list(
    name = "demoPkg",
    title = "Demo Package",
    version = "0.0.1",
    authors = list(list(name = "Demo Author", role = "aut, cre")),
    dependencies = c("cli", "rlang"),
    exported_functions = list(
      list(name = "say_hello", file = "R/hello.R")
    )
  ),
  auto_unbox = TRUE
)

enrich_json <- jsonlite::toJSON(
  list(
    name = "demoPkg",
    title = "Demo Package",
    version = "0.0.1",
    category_tags = c("infrastructure", "cli"),
    complexity = "simple",
    functions = list(
      list(
        name = "say_hello",
        summary = "Returns a greeting string for a supplied name."
      )
    )
  ),
  auto_unbox = TRUE
)

make_extract_chat <- function() docs$create_chat(list(docs$step_text(extract_json)))
make_enrich_chat <- function() docs$create_chat(list(docs$step_text(enrich_json)))
make_report_chat <- function() {
  docs$create_chat(list(
    docs$step_text(paste(
      "## Package Summary",
      "",
      "- **Name:** demoPkg",
      "- **Version:** 0.0.1",
      "- **Key function:** `say_hello()` returns a greeting string.",
      sep = "\n"
    ))
  ))
}
```

```{r, echo = FALSE, results = 'asis'}
docs$status_block()
```

This example builds a pipeline that extracts structured metadata from an R
package's source files. Each step is a separate agent call, with the output
of one step feeding into the next.

This follows the **Prompt Chaining** pattern: sequential agents where each
step has a focused task and produces structured output for the next.

## When to Use Prompt Chaining

Prompt chaining works well when:

- The task has distinct phases (extract, transform, report)
- Each phase benefits from a different system prompt or focus
- You need validated structured data between steps
- A single monolithic prompt would be too complex or unreliable

For open-ended exploration, an autonomous agent is better
(see `vignette("example-data-analysis")`).

## The Pipeline

Our pipeline has three steps:

1. **Extract** -- Read package files and extract raw metadata as JSON
2. **Enrich** -- Categorise and annotate the extracted metadata
3. **Report** -- Produce a human-readable summary

Each step uses `output_format` to constrain the LLM to a JSON schema.

## Step 1: Extract Package Metadata

The first agent reads the DESCRIPTION file and key R source files, then
extracts structured information:

```{r step1-schema, eval = docs$should_eval}
extract_schema <- list(
  type = "object",
  properties = list(
    name = list(type = "string"),
    title = list(type = "string"),
    version = list(type = "string"),
    authors = list(
      type = "array",
      items = list(
        type = "object",
        properties = list(
          name = list(type = "string"),
          role = list(type = "string")
        ),
        required = c("name", "role")
      )
    ),
    dependencies = list(
      type = "array",
      items = list(type = "string")
    ),
    exported_functions = list(
      type = "array",
      items = list(
        type = "object",
        properties = list(
          name = list(type = "string"),
          file = list(type = "string")
        ),
        required = c("name")
      )
    )
  ),
  required = c("name", "title", "version", "authors", "dependencies",
    "exported_functions")
)
```

```{r step1-run, eval = docs$should_eval}
chat <- make_extract_chat()

extractor <- Agent$new(
  chat = chat,
  tools = tools_file(),
  permissions = permissions_readonly(),
  system_prompt = "You are a metadata extractor. Read the package
    DESCRIPTION file and scan R/ source files for exported functions
    (look for @export roxygen tags). Return structured JSON only."
)

step1 <- extractor$run_sync(
  "Extract metadata from this R package. Read DESCRIPTION and scan
   the R/ directory for exported functions.",
  output_format = list(type = "json_schema", schema = extract_schema)
)

# Validate the output before proceeding
stopifnot(step1$is_success())
stopifnot(isTRUE(step1$structured_output$valid) ||
  is.na(step1$structured_output$valid))

pkg_metadata <- step1$structured_output$parsed
pkg_metadata$name
pkg_metadata$version
length(pkg_metadata$exported_functions)
```

## Step 2: Enrich with Categorisation

The second agent takes the extracted JSON and adds categorisation and
summaries:

```{r step2-schema, eval = docs$should_eval}
enrich_schema <- list(
  type = "object",
  properties = list(
    name = list(type = "string"),
    title = list(type = "string"),
    version = list(type = "string"),
    category_tags = list(
      type = "array",
      items = list(
        type = "string",
        enum = c("data", "modeling", "visualization", "infrastructure",
          "testing", "io", "web", "cli")
      )
    ),
    complexity = list(
      type = "string",
      enum = c("simple", "moderate", "complex")
    ),
    functions = list(
      type = "array",
      items = list(
        type = "object",
        properties = list(
          name = list(type = "string"),
          summary = list(type = "string")
        ),
        required = c("name", "summary")
      )
    )
  ),
  required = c("name", "title", "version", "category_tags",
    "complexity", "functions")
)
```

```{r step2-run, eval = docs$should_eval}
chat2 <- make_enrich_chat()

enricher <- Agent$new(
  chat = chat2,
  tools = tools_file(),
  permissions = permissions_readonly(),
  system_prompt = "You are a package analyst. Given package metadata,
    categorise the package, assess its complexity, and write a one-line
    summary of each exported function. You may read source files to
    understand what functions do. Return structured JSON only."
)

# Pass the extracted metadata as context
step2 <- enricher$run_sync(
  paste(
    "Enrich this package metadata with categorisation and function summaries.",
    "Read the source files if needed to understand what functions do.",
    "",
    "Package metadata:",
    jsonlite::toJSON(pkg_metadata, auto_unbox = TRUE, pretty = TRUE)
  ),
  output_format = list(type = "json_schema", schema = enrich_schema)
)

stopifnot(step2$is_success())
enriched <- step2$structured_output$parsed
enriched$category_tags
enriched$complexity
```

## Step 3: Generate Report

The third agent takes the enriched data and produces a readable report:

```{r step3-run, eval = docs$should_eval}
chat3 <- make_report_chat()

reporter <- Agent$new(
  chat = chat3,
  system_prompt = "You are a technical writer. Given enriched package
    metadata, produce a clear, concise package summary report in
    markdown format. Include sections for overview, key functions,
    dependencies, and recommendations."
)

step3 <- reporter$run_sync(
  paste(
    "Write a package summary report from this enriched metadata:",
    jsonlite::toJSON(enriched, auto_unbox = TRUE, pretty = TRUE)
  )
)

cat(step3$response)
```

## Putting It Together

Wrap the pipeline in a reusable function:

```{r pipeline-function, eval = docs$should_eval}
extraction_pipeline <- function(package_dir = ".") {
  # Step 1: Extract
  chat1 <- make_extract_chat()
  extractor <- Agent$new(
    chat = chat1,
    tools = tools_file(),
    permissions = permissions_readonly(),
    working_dir = package_dir,
    system_prompt = "Extract package metadata. Read DESCRIPTION and
      scan R/ for exported functions. Return structured JSON only."
  )

  step1 <- extractor$run_sync(
    "Extract metadata from this R package.",
    output_format = list(type = "json_schema", schema = extract_schema)
  )

  if (!step1$is_success()) {
    stop("Extraction failed: ", step1$stop_reason)
  }

  # Step 2: Enrich
  chat2 <- make_enrich_chat()
  enricher <- Agent$new(
    chat = chat2,
    tools = tools_file(),
    permissions = permissions_readonly(),
    working_dir = package_dir,
    system_prompt = "Categorise and summarise package functions.
      Return structured JSON only."
  )

  step2 <- enricher$run_sync(
    paste("Enrich this metadata:", jsonlite::toJSON(
      step1$structured_output$parsed,
      auto_unbox = TRUE
    )),
    output_format = list(type = "json_schema", schema = enrich_schema)
  )

  if (!step2$is_success()) {
    stop("Enrichment failed: ", step2$stop_reason)
  }

  # Step 3: Report
  chat3 <- make_report_chat()
  reporter <- Agent$new(chat = chat3)

  step3 <- reporter$run_sync(
    paste("Write a package summary:", jsonlite::toJSON(
      step2$structured_output$parsed,
      auto_unbox = TRUE, pretty = TRUE
    ))
  )

  list(
    metadata = step1$structured_output$parsed,
    enriched = step2$structured_output$parsed,
    report = step3$response,
    total_cost = step1$cost$total + step2$cost$total + step3$cost$total
  )
}

# Usage
result <- extraction_pipeline(".")
cat(result$report)
```

## Error Handling Between Steps

Each step validates its output before passing it to the next. Check both
`is_success()` (the agent completed normally) and `structured_output$valid`
(the JSON matches the schema):

```{r error-handling, eval = docs$should_eval}
step <- extractor$run_sync(
  "Extract metadata.",
  output_format = list(type = "json_schema", schema = extract_schema)
)

if (!step$is_success()) {
  cli::cli_abort("Agent stopped early: {step$stop_reason}")
}

if (identical(step$structured_output$valid, FALSE)) {
  cli::cli_abort(c(
    "Output doesn't match schema",
    "x" = step$structured_output$errors
  ))
}

# Safe to use the parsed output
parsed <- step$structured_output$parsed
```

## Next Steps

- `vignette("structured-output")` -- JSON schema output and validation
- `vignette("permissions")` -- Permission presets and read-only mode
- `vignette("example-code-review")` -- Multi-agent orchestration with
  LeadAgent
