---
title: "Example: Shiny Chat with shinychat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example: Shiny Chat with shinychat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This example shows how to use deputy with
[shinychat](https://posit-dev.github.io/shinychat/) to build an interactive
chat application with permissions, hooks, and tool call limits.

## The Problem

shinychat's `chat_append()` expects an ellmer content stream from
`chat$stream_async()`. Deputy's `run()` yields `AgentEvent` objects, while
`run_sync()` returns an `AgentResult`. Calling `agent$chat$stream_async()`
directly works but bypasses Deputy's permissions, hooks, and run limits from
`UsageLimits()`.

`run_shiny()` bridges this gap: it returns a content stream that
shinychat understands while still enforcing deputy's permissions, hooks,
and limits.

`run_shiny()` is deliberately strict about file tools: every supplied file
path must be absolute and resolve within the Agent's immutable `working_dir`.
Relative paths, omitted path defaults, and paths outside that root are rejected.
Bind a normalized workspace and include its exact path in the model's system
prompt, as both examples below do.

Following shinychat's recommended pattern, pass that stream directly to
`chat_append()`. `chat_append()` returns the completion promise, so make
it the last expression in your observer if you want shinychat to surface
streaming errors in the chat UI.

## What `run_shiny()` Enforces

| Feature | Enforced? | How |
|---|---|---|
| Permissions (file_read, bash, etc.) | Yes | `on_tool_request` callback |
| PreToolUse / PostToolUse hooks | Yes | `on_tool_request` / `on_tool_result` callbacks |
| File path confinement | Yes | Every recognized file path must be absolute and inside `working_dir` |
| Tool call limit | Yes | Checked at each tool request; an overage cancels the active stream |
| Request, token, and cost limits | Yes | Checked when usage is observed; an overage cancels the active stream |
| SessionStart / SessionEnd hooks | Yes | Fired before/after the stream |
| Stall detection | No | Requires deputy's own loop |
| `output_format` (structured output) | No | Requires deputy's own loop |

## Basic Setup

```{r}
library(shiny)
library(deputy)
library(shinychat)

workspace <- normalizePath(getwd(), mustWork = TRUE, winslash = "/")

ui <- bslib::page_fluid(
  chat_ui("chat", fill = TRUE)
)

server <- function(input, output, session) {
  chat <- ellmer::chat_openai(
    model = "gpt-4o-mini",
    system_prompt = paste(
      "You are a helpful assistant. Be concise.",
      sprintf(
        "For every file tool, use an absolute path inside: %s",
        workspace
      )
    )
  )

  agent <- Agent$new(
    chat = chat,
    tools = tools_file(),
    permissions = permissions_standard(workspace),
    working_dir = workspace
  )

  observeEvent(input$chat_user_input, {
    req(input$chat_user_input)

    chat_append(
      "chat",
      agent$run_shiny(input$chat_user_input)
    )
  })
}

shinyApp(ui, server)
```

## With Permissions and Hooks

Deputy's permissions and hooks fire on every tool call, even though
shinychat drives the streaming loop:

```{r}
workspace <- normalizePath(getwd(), mustWork = TRUE, winslash = "/")

server <- function(input, output, session) {
  chat <- ellmer::chat_anthropic(
    model = "claude-sonnet-4-5-20250929",
    system_prompt = paste(
      "You are a data analyst. Be concise.",
      sprintf(
        "For every file tool, use an absolute path inside: %s",
        workspace
      )
    )
  )

  agent <- Agent$new(
    chat = chat,
    tools = c(tools_file(), tools_data()),
    permissions = Permissions$new(
      file_read = TRUE,
      file_write = FALSE,
      r_code = FALSE,
      bash = FALSE
    ),
    usage_limits = UsageLimits(max_requests = 10, max_cost_usd = 0.50),
    working_dir = workspace
  )

  # Hooks still fire normally
  agent$add_hook(hook_log_tools(verbose = TRUE))

  observeEvent(input$chat_user_input, {
    req(input$chat_user_input)

    # max_tool_calls limits how many tool calls the agent can make
    chat_append(
      "chat",
      agent$run_shiny(
        input$chat_user_input,
        max_tool_calls = 10
      )
    )
  })
}
```

## How Limits Work

`run_shiny()` starts with the Agent's run limits and lets `max_tool_calls`
override that one field. When Deputy observes a request, tool, token, or cost
overage, it marks the run stopped and asks ellmer's stream controller to cancel
active generation. An overage found at a tool boundary also rejects that tool
call. `Stop` and `SessionEnd` hooks receive the corresponding limit reason.

Provider token and cost usage is not observable until a response arrives. The
initial response can therefore cross one of those limits before cancellation;
it may overshoot the configured threshold by one response. Cancellation stops
further work but cannot retract content already emitted. Treat token and cost
limits as observed stop conditions, not provider-side generation ceilings.

The `max_tool_calls` parameter counts individual tool call requests, not model
requests. One model request can include multiple parallel tool calls (for
example, reading three files at once), each counting separately.

## Running the Example

A complete example app is included in the package:

```{r}
shiny::runApp(system.file("examples/shiny-chat", package = "deputy"))
```
