---
title: "Example: Shiny Chat with shinychat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example: Shiny Chat with shinychat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Deputy Agents implement the ellmer chat methods used by
[shinychat](https://posit-dev.github.io/shinychat/). Pass an Agent directly to
`chat_server()`; there is no separate Shiny bridge and no path that bypasses
Deputy's governance.

```text
shinychat input -> Agent$stream_async() -> one governed run kernel -> ellmer
                                      \-> AgentResult + hooks + usage
```

The stream keeps ellmer's native chunks, so shinychat can render incremental
text, tool activity, and attachment-enabled input. After it finishes,
`agent$last_run()` exposes the corresponding `AgentResult`.

## Basic Setup

```{r}
library(shiny)
library(deputy)
library(shinychat)

workspace <- normalizePath(getwd(), mustWork = TRUE, winslash = "/")

ui <- bslib::page_fluid(
  chat_ui("chat", fill = TRUE, allow_attachments = TRUE)
)

server <- function(input, output, session) {
  chat <- ellmer::chat_openai(
    model = "gpt-5.6-luna",
    system_prompt = "You are a concise data assistant."
  )

  agent <- Agent$new(
    chat = chat,
    tools = c(tools_file(), tools_data()),
    permissions = permissions_readonly(),
    usage_limits = UsageLimits(
      max_requests = 10,
      max_tool_calls = 12,
      max_cost_usd = 0.50
    ),
    context_policy = ContextPolicy(
      max_tokens = 32000,
      fallback = "error"
    ),
    working_dir = workspace
  )

  chat_server("chat", agent)
}

shinyApp(ui, server)
```

`chat_server()` passes both plain text and attachment-enabled ellmer content to
the Agent unchanged. It also supplies shinychat history and cancellation around
Deputy's governed stream.

## What the Agent Adds

Every public chat and run method uses the same kernel:

| Method | Native return | Typical host |
|---|---|---|
| `chat()` | final text | scripts and consoles |
| `chat_async()` | promise of final text | async applications |
| `chat_structured()` / `chat_structured_async()` | typed data | structured-output hosts |
| `stream()` | synchronous ellmer stream | terminal hosts |
| `stream_async()` | asynchronous ellmer stream | shinychat |
| `run_sync()` | `AgentResult` | inspected workflows |
| `run_async()` | promise of `AgentResult` | subagents |

All interfaces enforce the Agent's permissions, hooks, `UsageLimits`, workspace path
resolution, file checkpoints, automatic context compaction, large-result
offloading, and run accounting. Native relative file paths resolve against the
Agent's immutable `working_dir`; Deputy never changes the process working
directory. Custom hosts can pass `run_context` to any chat, stream, or run
method to correlate a request with product-owned session and stage identities.

`ContextPolicy()` checks the estimated complete context after run initialization
and again at model-request boundaries between tool rounds, compacting whenever
it crosses the token threshold. Summary streaming is asynchronous and shares the
run's cancellation controller and usage limits. The default fails closed if LLM summary generation fails. Set
`fallback = "text"` only when a deterministic degraded summary is acceptable.
`summary_fallback_chats` explicitly configures separate summary recovery
destinations. `agent$last_compaction()` reports the method, run ID, attempt
conditions, and compaction usage.

Large tool results are stored as integrity-checked envelopes and replaced in
model context with a preview and `deputy://tool-result/...` reference. The
Agent registers `deputy_read_tool_result()` so the model can page through the
stored result in bounded chunks. Applications can retrieve the original R
value with `agent$resolve_tool_result(reference)`.

## Hooks and Cancellation

Hooks attach to the Agent as usual:

```{r}
agent$add_hook(hook_log_tools(verbose = TRUE))

agent$add_hook(HookMatcher$new(
  event = "PostCompact",
  callback = function(result, context) {
    cli::cli_inform("Compacted with: {result$method}")
  }
))
```

`chat_append()` owns the completion promise and surfaces stream errors in the
chat UI. shinychat cancellation propagates through ellmer's stream controller;
Deputy still finalizes checkpoints and fires `Stop` and `SessionEnd`.

## Subagents

`LeadAgent` delegation uses `run_async()` internally, so a subagent does not
block the Shiny R process while the parent chat is streaming. A custom async
tool can use the same contract:

```{r}
tool_scout <- ellmer::tool(
  fun = coro::async(function(prompt) {
    result <- coro::await(scout$run_async(prompt))
    result$response
  }),
  name = "agent_scout",
  description = "Delegate a literature search.",
  arguments = list(prompt = ellmer::type_string("Self-contained task"))
)
agent$register_tool(tool_scout)
```

## Running the Example

```{r}
shiny::runApp(system.file("examples/shiny-chat", package = "deputy"))
```
