---
title: "Data analysis agent"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data analysis agent}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This recipe gives an agent a dataset and R, and lets it decide what to compute
next from what it has found so far. It suits open-ended exploration. When the
steps are known in advance, a fixed [extraction
pipeline](example-extraction-pipeline.html) is cheaper and easier to check.

The agent runs R code with your user account's access to files and the
network. That's reasonable for your own data and prompts. For untrusted input,
run the code in a sandbox instead; see [Running R code](code-execution.html).

## Set up the agent

The agent gets the data-reading tools, `run_r_code`, and the bundled
`data_analysis` skill, which adds two summary tools and some analysis
guidance. The skill needs dplyr and ggplot2.

```{r setup}
library(deputy)

agent <- Agent$new(
  chat = ellmer::chat("anthropic/claude-sonnet-5"),
  tools = c(tools_data(), list(tool_run_r_code)),
  permissions = Permissions(r_code = TRUE, file_write = FALSE),
  usage_limits = UsageLimits(max_requests = 15, max_cost_usd = 0.50),
  system_prompt = "You are a data scientist. When you explore a dataset:
    1. Start with its structure and summary statistics.
    2. Check for missing values and data quality problems.
    3. Look at the distributions of the key variables.
    4. Look for relationships between variables.
    5. Summarise what you found.
    Use R code for every calculation, and say why you run each step."
)

agent$load_skill(system.file("skills", "data_analysis", package = "deputy"))
```

The skill's tools, `eda_summary` and `describe_column`, give the model quick
overviews, so it can save `run_r_code` for the questions they don't answer.

## Run the analysis

```{r run-analysis}
result <- agent$run_sync(
  "Explore the airquality dataset that comes with R. Tell me about data
   quality problems, the distributions of the variables, and any patterns."
)

cat(result$response)
```

`airquality` has missing values in `Ozone` and `Solar.R`, and daily readings
over five months, so a good answer will mention both. Each run can take a
different path, because the model chooses its next step from the results so
far.

The run stops after 15 model requests or $0.50, whichever comes first, and
`result$stop_reason` says which.

## Watch it work

`run()` returns a generator of events, so you can print progress as the agent
works:

```{r streaming}
events <- agent$run("Summarise the mtcars dataset.")

repeat {
  event <- events()
  if (coro::is_exhausted(event)) {
    break
  }
  switch(event$type,
    text = cat(event$text),
    tool_start = cli::cli_alert_info("Calling {event$tool_name}"),
    tool_end = cli::cli_alert_success("{event$tool_name} done"),
    stop = cli::cli_alert("Stopped: {event$reason}")
  )
}
```

For a simpler log, add `hook_log_tools()` before calling `run_sync()`.

## Check the result

```{r inspect-results}
result_is_success(result)
result$stop_reason

result_n_turns(result)
length(result_tool_calls(result))

result$duration
result$usage
```

`result_tool_calls()` also shows the code the model ran, in each call's
`tool_input`, which is the quickest way to check how it reached a number.

## Next steps

- [Tools](tools.html): skills and custom tools.
- [Running R code](code-execution.html): keep R state between calls, or
  sandbox the code.
- [Extraction pipeline](example-extraction-pipeline.html): fixed steps with
  structured output.
