---
title: "Shiny chat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shiny chat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

An agent works anywhere [shinychat](https://posit-dev.github.io/shinychat/)
expects an ellmer chat. Pass it to `chat_server()` and the app streams the
model's replies and tool calls, while Deputy applies the agent's permissions,
hooks and limits to every message.

## A minimal app

```{r}
library(shiny)
library(shinychat)
library(deputy)

workspace <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)

ui <- bslib::page_fluid(
  chat_ui("chat", fill = TRUE, allow_attachments = TRUE)
)

server <- function(input, output, session) {
  agent <- Agent$new(
    chat = ellmer::chat(
      "openai/gpt-6-luna",
      system_prompt = "You are a concise data assistant."
    ),
    tools = tools_data(),
    permissions = permissions_readonly(),
    usage_limits = UsageLimits(max_requests = 10, max_tool_calls = 12),
    working_dir = workspace
  )

  chat_server("chat", agent)
}

shinyApp(ui, server)
```

Create the agent inside `server()`, so that each browser session gets its own
conversation, usage and hooks. shinychat calls the agent's `stream_async()`
method, passing text and attachments through unchanged, and handles the
input box, cancellation and conversation history. After each reply,
`agent$last_run()` returns the run's `AgentResult`.

The limits apply to each message the user sends. The agent also keeps its
[context policy](conversations.html): long conversations are compacted
automatically, and large tool results are stored outside the model's context.

## Tell users when older messages are summarised

When compaction runs, the model sees a summary instead of the oldest messages,
but the chat still shows the whole conversation. The example app in
`inst/examples/shiny-chat/` shows a small notice above the chat while that
happens ("Summarizing earlier messages…") and afterwards ("Earlier messages
summarized for the assistant. Your full conversation is still available."),
with a "View summary" link that shows the summary text.

The notice is driven by hooks: `PreCompact` sets it to busy, `PostCompact`
records the summary, and `Stop` clears the busy state if a run fails or is
cancelled. If you call `agent$compact()` yourself, clear the notice when that
call returns, because a manual compaction doesn't fire `Stop`. The notice
belongs to the page, not the conversation: Deputy never adds messages to the
chat to report compaction.

shinychat's conversation history, where available, stores the whole
conversation (`agent$get_turns()`), so restoring a conversation brings back
every message while the model continues from the summary.

## Hooks in an app

Hooks are added to the agent as usual, inside `server()`:

```{r}
agent$add_hook(hook_log_tools())

agent$add_hook(HookMatcher(
  event = "PostCompact",
  callback = function(result, context) {
    showNotification("Earlier messages were summarised.")
    NULL
  }
))
```

When the user cancels a reply, the run stops at the next point where ellmer can
stop it, and Deputy still fires the `Stop` and `SessionEnd` hooks and saves any
file checkpoints.

To ask the user questions from a tool, give each session its own handler with
`tools_interactive()`; see [Human input and approvals](approvals.html).

## Long-running tools

Delegating to a subagent doesn't block the app: `LeadAgent` runs subagents with
`run_async()`, so other sessions keep working while the lead waits. Your own
tools can do the same by returning a promise:

```{r}
ask_scout <- ellmer::tool(
  coro::async(function(prompt) {
    result <- coro::await(scout$run_async(prompt))
    result$response
  }),
  name = "ask_scout",
  description = "Ask the literature scout to search for papers.",
  arguments = list(prompt = ellmer::type_string("A self-contained task"))
)
agent$register_tool(ask_scout)
```

Here `scout` is another `Agent`. A tool that calls another agent like this
should also have annotations, like any other tool; see [Tools](tools.html).

## Run the example

The package includes the app with the compaction notice:

```{r}
shiny::runApp(system.file("examples", "shiny-chat", package = "deputy"))
```

For apps that show subagent conversations beside the main chat, see [Briefing
and inspecting subagents](delegation.html#a-subagent-panel-for-shiny).
