---
title: "Multi-agent code review"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-agent code review}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This recipe reviews an R package with three specialist subagents, one each for
bugs, style and documentation, and a lead agent that combines their findings
into one report. Each reviewer works in its own conversation with its own
prompt, so its instructions stay focused and you can inspect its work
separately.

## Define the reviewers

All three reviewers get the same read-only tools for finding and reading code:

```{r define-agents}
library(deputy)

review_tools <- list(
  tool_read_file,
  tool_list_files,
  tool_glob_files,
  tool_grep_files
)

bug_hunter <- agent_definition(
  name = "bug_hunter",
  description = "Finds logic errors, unhandled edge cases and likely runtime
    failures in R code.",
  prompt = "You look for bugs in R code: wrong conditions, NULL/NA/empty
    inputs that aren't handled, off-by-one indexing, unchecked external calls,
    and surprising type coercion. Report only real bugs, each with a file path
    and line number. Ignore style.",
  tools = review_tools,
  permission_mode = "readonly"
)

style_reviewer <- agent_definition(
  name = "style_reviewer",
  description = "Checks R code against the tidyverse style guide.",
  prompt = "You review R code against the tidyverse style guide: naming,
    assignment with <-, function length and clear names. Flag what hurts
    readability; skip trivial nitpicks.",
  tools = review_tools,
  permission_mode = "readonly"
)

doc_checker <- agent_definition(
  name = "doc_checker",
  description = "Checks that roxygen2 documentation is complete and matches
    the code.",
  prompt = "You review roxygen2 documentation: missing @param, @return and
    @examples, and descriptions that don't match what the function does. Read
    the function body to check.",
  tools = review_tools,
  permission_mode = "readonly"
)
```

## Create the lead

The lead decides what to ask each reviewer and merges their answers:

```{r create-lead}
lead <- LeadAgent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  sub_agents = list(bug_hunter, style_reviewer, doc_checker),
  system_prompt = "You lead a code review. Give each specialist a clear,
    specific task, then merge their findings: remove duplicates, sort by
    severity (critical, warning, suggestion), and give a file path and a
    concrete fix for each finding.",
  permissions = Permissions(file_write = FALSE),
  usage_limits = UsageLimits(max_requests = 20)
)

lead$available_sub_agents()
#> [1] "bug_hunter"     "style_reviewer" "doc_checker"
```

The lead may delegate but not write files, and the reviewers run in read-only
mode, so nobody changes a file. The lead can't use `permissions_readonly()`
itself: read-only mode allows only reading tools, so it would deny the lead's
own `delegate_to_agent` calls.

To follow along while it runs, log each reviewer as it finishes, and each tool
call:

```{r monitoring-hooks}
lead$add_hook(HookMatcher(
  event = "SubagentStop",
  callback = function(agent_name, task, result, context) {
    cli::cli_alert_info("{agent_name} finished")
    NULL
  }
))
lead$add_hook(hook_log_tools())
```

## Run the review

```{r run-review}
result <- lead$run_sync(
  "Review the R files in R/. Ask each specialist to review them from their
   perspective, then combine the findings into one report."
)

cat(result$response)
```

## Get the findings as a data frame

For a report you can filter and count, ask for structured output. An array of
objects comes back as a data frame, and enums as factors:

```{r structured-review}
review_type <- ellmer::type_object(
  summary = ellmer::type_string("Two or three sentences on overall quality."),
  findings = ellmer::type_array(
    ellmer::type_object(
      severity = ellmer::type_enum(c("critical", "warning", "suggestion")),
      category = ellmer::type_enum(c("bug", "style", "documentation")),
      file = ellmer::type_string(),
      line = ellmer::type_integer(required = FALSE),
      issue = ellmer::type_string(),
      fix = ellmer::type_string(),
      reviewer = ellmer::type_string()
    )
  )
)

result <- lead$run_sync(
  "Review the R files in R/ and return every finding.",
  type = review_type
)

review <- result$structured_output
findings <- review$findings
```

From there it's ordinary R:

```{r process-results}
cli::cli_h1("Code review")
cli::cli_text(review$summary)

table(findings$category, findings$severity)

critical <- findings[findings$severity == "critical", ]
for (i in seq_len(nrow(critical))) {
  cli::cli_h2(critical$file[[i]])
  cli::cli_alert_danger(critical$issue[[i]])
  cli::cli_alert_info("Fix: {critical$fix[[i]]}")
}
```

## Cost and effort

The lead's result includes the reviewers' usage:

```{r costs}
result$usage
result$duration
length(result_tool_calls(result))
```

`result$usage$cost_usd` is `NA` when ellmer couldn't price every response,
for example with a model newer than your ellmer version. It's never a partial
total that looks complete. `lead$list_subagents()` shows each delegation,
with its status and stop reason, if you want to see where the requests went.

## Next steps

- [Subagents](multi-agent.html): definitions, budgets and parallel
  delegation.
- [Hooks](hooks.html): events and built-in hooks.
- [Structured output](structured-output.html): types and validation.
