---
title: "Example: Shiny Chat with shinychat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example: Shiny Chat with shinychat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Deputy Agents implement the ellmer chat methods used by
[shinychat](https://posit-dev.github.io/shinychat/). Pass
`agent$stream_async()` directly to `chat_append()`; there is no separate Shiny
bridge and no path that bypasses Deputy's governance.

```text
shinychat input -> Agent$stream_async() -> one governed run kernel -> ellmer
                                      \-> AgentResult + hooks + usage
```

The stream keeps ellmer's native chunks, so shinychat can render incremental
text, tool activity, and attachment-enabled input. After it finishes,
`agent$last_run()` exposes the corresponding `AgentResult`.

## Basic Setup

```{r}
library(shiny)
library(deputy)
library(shinychat)

workspace <- normalizePath(getwd(), mustWork = TRUE, winslash = "/")

ui <- bslib::page_fluid(
  chat_ui("chat", fill = TRUE, allow_attachments = TRUE)
)

server <- function(input, output, session) {
  chat <- ellmer::chat_openai(
    model = "gpt-4o-mini",
    system_prompt = "You are a concise data assistant."
  )

  agent <- Agent$new(
    chat = chat,
    tools = c(tools_file(), tools_data()),
    permissions = permissions_readonly(),
    usage_limits = UsageLimits(
      max_requests = 10,
      max_tool_calls = 12,
      max_cost_usd = 0.50
    ),
    context_policy = ContextPolicy(
      max_tokens = 32000,
      fallback = "error"
    ),
    working_dir = workspace
  )

  observeEvent(input$chat_user_input, {
    req(input$chat_user_input)
    chat_append(
      "chat",
      agent$stream_async(input$chat_user_input)
    )
  })
}

shinyApp(ui, server)
```

`input$chat_user_input` may be a string or shinychat's list of ellmer content
objects for attachments. Deputy forwards it unchanged.

## What the Agent Adds

Every public execution method uses the same kernel:

| Method | Native return | Typical host |
|---|---|---|
| `chat()` | final text | scripts and consoles |
| `chat_async()` | promise of final text | async applications |
| `stream()` | synchronous ellmer stream | terminal hosts |
| `stream_async()` | asynchronous ellmer stream | shinychat |
| `run_sync()` | `AgentResult` | inspected workflows |
| `run_async()` | promise of `AgentResult` | delegated workers |

All six enforce the Agent's permissions, hooks, `UsageLimits`, workspace path
resolution, file checkpoints, automatic context compaction, large-result
offloading, and run accounting. Native relative file paths resolve against the
Agent's immutable `working_dir`; Deputy never changes the process working
directory.

`ContextPolicy()` compacts before the estimated complete context crosses its
token threshold. The default fails closed if LLM summarization fails. Set
`fallback = "text"` only when a deterministic degraded summary is acceptable.
`agent$last_compaction()` reports which method was used and the summarization
usage.

Large tool results are stored as integrity-checked envelopes and replaced in
model context with a `deputy://tool-result/...` reference. Applications can
retrieve the original R value with `agent$resolve_tool_result(reference)`.

## Hooks and Cancellation

Hooks attach to the Agent as usual:

```{r}
agent$add_hook(hook_log_tools(verbose = TRUE))

agent$add_hook(HookMatcher$new(
  event = "PostCompact",
  callback = function(result, context) {
    message("Compacted with: ", result$method)
  }
))
```

`chat_append()` owns the completion promise and surfaces stream errors in the
chat UI. shinychat cancellation propagates through ellmer's stream controller;
Deputy still finalizes checkpoints and fires `Stop` and `SessionEnd`.

## Delegated Workers

`LeadAgent` delegation uses `run_async()` internally, so a child Agent does not
block the Shiny R process while the parent chat is streaming. A custom async
tool can use the same contract:

```{r}
tool_scout <- ellmer::tool(
  fun = coro::async(function(prompt) {
    result <- coro::await(scout$run_async(prompt))
    result$response
  }),
  name = "agent_scout",
  description = "Delegate a literature search.",
  arguments = list(prompt = ellmer::type_string("Self-contained task"))
)
agent$register_tool(tool_scout)
```

## Running the Example

```{r}
shiny::runApp(system.file("examples/shiny-chat", package = "deputy"))
```
