---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article builds an agent that explains an R package by reading its source.
You will run a task, check what the agent did, and then give a second agent
permission to write a file, and undo what it wrote.

## Setup

Install Deputy from GitHub:

```r
# install.packages("pak")
pak::pak("JamesHWade/deputy")
```

The examples use OpenAI's `gpt-6-luna`, so they need `OPENAI_API_KEY`. Put it
in your `.Renviron` file (`usethis::edit_r_environ()` opens it) and restart R.
Keep keys out of scripts and version control.

To use another provider, pass a different
[ellmer chat](https://ellmer.tidyverse.org/reference/index.html). Deputy uses
whatever model and settings that chat has.

## Create an agent

Start R in the package you want to explain, then run:

```{r first-agent, eval = FALSE}
library(deputy)

workspace <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)

agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  tools = tools_preset("minimal"),
  permissions = permissions_readonly(),
  usage_limits = UsageLimits(max_requests = 6, max_tool_calls = 8),
  working_dir = workspace
)
```

- `chat` is the ellmer chat that talks to the model.
- `tools` gives the model three tools: `read_file`, `read_markdown` and
  `list_files`.
- `permissions` allows reading and denies writing files, running code and
  using the web.
- `usage_limits` ends the run after six model requests or eight tool calls.
- `working_dir` is where relative paths start. To explain a package somewhere
  else, use its path instead of `getwd()`.

Read-only stops the agent from changing anything. It does not limit what it
reads: the agent can open any file your R session can open. Permissions decide
which tools may run; they are not a filesystem sandbox.

## Run a task

Ask a question that the files can answer:

```{r first-run, eval = FALSE}
result <- agent$run_sync(
  "Read DESCRIPTION and list R/. Explain what this package does and cite the files you used."
)

result$response
```

`run_sync()` blocks until the run ends and returns an `AgentResult`. Before you
use the response, check why the run ended:

```{r inspect-result, eval = FALSE}
result$stop_reason
result$usage
```

`"complete"` means the model finished. A limit stops the run with a reason such
as `"request_limit"` or `"tool_call_limit"`, and the response may be partial.
`result$usage` counts requests, tool calls and tokens, and the cost when ellmer
can price the model.

To see which files the agent read, list its tool calls and their results:

```{r inspect-tools, eval = FALSE}
result_tool_calls(result)
result_tool_results(result)
```

`?AgentResult` describes the other fields, including the full event log and
run identifiers.

## Other ways to run a task

`run_sync()` suits scripts. An agent also works like an ellmer chat, so pick the
method that fits your program:

| Method | Returns | Use it for |
|---|---|---|
| `chat()` | the final text | quick questions at the console |
| `run()` | a generator of events | printing progress as it happens |
| `run_async()` | a promise of an `AgentResult` | async code and background work |
| `stream_async()` | an ellmer stream | [Shiny apps](example-shiny-chat.html) |

Every method goes through the same permission checks, hooks and limits.

## Watch tool calls as they happen

Add a hook that prints each tool call:

```{r log-tools, eval = FALSE}
agent$add_hook(hook_log_tools())
result <- agent$run_sync("Read DESCRIPTION and name the package's imported dependencies.")
```

[Hooks](hooks.html) shows how to write your own.

## Let an agent write a file

An agent's permissions can be narrowed after it is created but never widened,
so writing needs a new agent. This one may write inside `workspace`. It still
can't run R or shell code, use the web or install packages.

```{r write-agent, eval = FALSE}
write_agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  tools = tools_file(),
  permissions = Permissions(file_write = workspace),
  usage_limits = UsageLimits(max_requests = 6, max_tool_calls = 8),
  enable_file_checkpointing = TRUE,
  working_dir = workspace
)

checkpoint_id <- write_agent$checkpoint("before package summary")

write_result <- write_agent$run_sync(
  "Read DESCRIPTION and write a short package summary to deputy-summary.md. Leave other files unchanged."
)

write_result$stop_reason
result_tool_calls(write_result)
```

Open `deputy-summary.md` and check the change in Git. Because checkpointing is
on, Deputy saved the old contents of every file its write tools touched after
the checkpoint. To undo the agent's changes:

```{r rewind, eval = FALSE}
write_agent$rewind_files(checkpoint_id)
```

Checkpoints cover Deputy's `write_file`, `edit_file` and `multi_edit` tools.
Changes made by R code, shell commands or other programs are not recorded.

## Troubleshooting

The provider can't authenticate
: Check the credential your ellmer provider expects. Try a plain ellmer chat
  before adding Deputy.

A tool call is denied
: Check that the tool is registered and that the permission it needs is on.
  Custom tools also need accurate annotations; see [Tools](tools.html).

The run stops early
: Look at `result$stop_reason` and `result$usage`, then give the agent a
  smaller task or raise the limit it hit. `"tool_loop"` means the model made
  the same tool call with the same result three times in a row.
  `"cost_unavailable"` means you set a cost limit but ellmer couldn't price the
  model, so Deputy stopped rather than guess. `?UsageLimits` lists every limit.

## Next steps

- [Tools](tools.html): choose built-in tools or turn your own R functions into
  tools.
- [Permissions and limits](permissions.html): decide what an agent may do and
  how much it may spend.
- [Structured output](structured-output.html): get lists and data frames back
  instead of text.
- [Shiny chat](example-shiny-chat.html): put an agent in an app.
- [Subagents](multi-agent.html): split a task across specialists.
