---
title: "Getting Started with Deputy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with Deputy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This guide uses Deputy to explain an R package from its source files. You will
run a task, inspect the answer and tool calls, then create an agent that can
write a summary file.

## Install and connect

Install Deputy from GitHub:

```r
# install.packages("pak")
pak::pak("JamesHWade/deputy")
```

You need credentials for an [ellmer provider](https://ellmer.tidyverse.org/reference/index.html).
These examples use OpenAI with `gpt-5.6-luna`. Set `OPENAI_API_KEY` in your R
environment before running them. You can edit your user `.Renviron` file with
[`usethis::edit_r_environ()`](https://usethis.r-lib.org/reference/edit.html) if you have usethis installed; restart R after
saving it. Keep credentials out of scripts and version control.

Use another ellmer chat constructor if you prefer a different provider. Deputy
keeps the model and parameters of the chat you supply.

## Create an agent

Open an R session in the package you want to inspect, then run:

```{r first-agent, eval = FALSE}
library(deputy)

workspace <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)

agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-5.6-luna"),
  tools = tools_preset("minimal"),
  permissions = permissions_readonly(),
  usage_limits = UsageLimits(max_requests = 6, max_tool_calls = 8),
  working_dir = workspace
)
```

Each setting has one job:

- `chat` connects to the model through ellmer.
- `tools` gives the model three tools: `read_file`, `read_markdown`, and
  `list_files`.
- `permissions` allows those reads and denies writes, code execution, and web
  requests.
- `usage_limits` stops the run after at most six model requests or eight tool
  calls.
- `working_dir = workspace` sets the base for relative file paths. To inspect
  a package elsewhere, replace `getwd()` in the `workspace <- normalizePath(...)`
  line with that package's path.

The read-only policy still allows access to files outside `working_dir` when
the R process can read them. It is not a filesystem sandbox.

## Run a task

Ask for an answer that the available files can support:

```{r first-run, eval = FALSE}
result <- agent$run_sync(
  "Read DESCRIPTION and list R/. Explain what this package does and cite the files you used."
)

result$response
```

`run_sync()` waits until the run stops, then returns an `AgentResult`.
The answer depends on the package and model. Check whether it finished before
using the response:

```{r inspect-result, eval = FALSE}
result$stop_reason
result$usage
```

`"complete"` means the task finished. Other reasons include `"request_limit"`
and `"tool_call_limit"`. A stopped run can still have a partial response.
`result$usage` reports requests, tokens, and any cost reported by the provider.

Inspect the tool calls to see which files the agent used:

```{r inspect-tools, eval = FALSE}
result$tool_calls()
result$tool_results()
```

The [AgentResult reference](../reference/AgentResult.html) describes the other
fields, including events and run identifiers.

## Show progress in the console

Add a logging hook before the next run to see tool activity as it happens:

```{r log-tools, eval = FALSE}
agent$add_hook(hook_log_tools())
result <- agent$run_sync("Read DESCRIPTION and name the package's imported dependencies.")
result$response
```

`hook_log_tools()` uses cli for console output. See [Hooks](hooks.html) to
write a callback or change what gets reported.

## Write a summary file

Writing needs a different permission policy. Create a new agent: permissions
on an existing agent can be narrowed, but cannot be expanded beyond those set
at construction.

This policy permits native file writes inside `workspace`. R execution, shell
commands, web requests, and package installation remain disabled by default.

```{r write-agent, eval = FALSE}
write_agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-5.6-luna"),
  tools = tools_file(),
  permissions = Permissions$new(file_write = workspace),
  usage_limits = UsageLimits(max_requests = 6, max_tool_calls = 8),
  enable_file_checkpointing = TRUE,
  working_dir = workspace
)

checkpoint_id <- write_agent$checkpoint("before package summary")

write_result <- write_agent$run_sync(
  "Read DESCRIPTION and write a short package summary to deputy-summary.md. Leave other files unchanged."
)

write_result$stop_reason
write_result$tool_calls()
```

Open `deputy-summary.md` and review the changes in Git. Deputy saved the
previous contents of files changed by its native write tools after the
checkpoint. If you want to undo those changes, run:

```{r rewind, eval = FALSE}
write_agent$rewind_files(checkpoint_id)
```

Checkpoints cover Deputy's native write, edit, and multi-edit tools. They do
not capture changes made by R code, shell commands, or other programs. See
[Permissions](permissions.html) for the limits of file policies and code
execution.

## Troubleshooting

### The provider cannot authenticate

Check the credential expected by your ellmer provider. Try a plain ellmer
chat to confirm the connection before adding tools or Deputy settings.

### A tool call is denied

Check that the tool is registered and that its required permission is enabled.
For custom tools, check their annotations and any allowlist as well. The
[Tools guide](tools.html) includes a complete custom-tool example.

### The run stops before finishing

Inspect `result$stop_reason` and `result$usage`. Try a smaller task, or raise the
limit that was reached. `"tool_loop"` means an identical tool call returned the
same result three times in a row. `"cost_unavailable"` means a cost limit was
set, but the provider did not report enough cost information to enforce it.
See `UsageLimits()` for all stop reasons.

## Next steps

- [Tools](tools.html): select a preset or register an R function.
- [Structured output](structured-output.html): extract data with ellmer types.
- [Shiny chat](example-shiny-chat.html): use the agent in an app.
- [Multiple agents](multi-agent.html): give separate tasks to other agents.
