---
title: "Getting Started with Deputy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with Deputy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Deputy turns an ellmer chat into a governed R Agent. In this guide you will
create a bounded, read-only Agent; inspect its result; then deliberately grant
one workspace-scoped write capability and protect it with a checkpoint.

## Before you begin

Before Deputy's first CRAN release, install the development version from
GitHub:

```r
# install.packages("pak")
pak::pak("JamesHWade/deputy")
```

After the release is available from CRAN, install the released package with:

```r
install.packages("deputy")
```

You also need credentials for one
[ellmer provider](https://ellmer.tidyverse.org/reference/index.html). The
examples use OpenAI, but the Deputy API is provider-independent.

## 1. Define the workspace and limits

Make the authority for the first run visible before creating a chat. The
`minimal` preset contains only `read_file`, `read_markdown`, and `list_files`.
The permission policy denies writes, code execution, shell commands, web
requests, and package installation. The run also has finite request and
tool-call budgets.

```{r first-policy}
library(deputy)

workspace <- normalizePath(getwd(), winslash = "/")
first_tools <- tools_preset("minimal")
first_permissions <- permissions_readonly()
first_limits <- UsageLimits(max_requests = 6, max_tool_calls = 8)

stopifnot(
  dir.exists(workspace),
  length(first_tools) == 3L,
  identical(first_permissions$mode, "readonly"),
  identical(first_limits$max_tool_calls, 8L)
)
```

Run this from the package you want to inspect, or replace `getwd()` with that
package's absolute path.

`working_dir` gives relative paths a stable base. It does not create a read
sandbox: a read permission can reach files that the R process can reach. Keep
the task narrow and run Deputy with operating-system isolation when the input
or generated code is untrusted.

## 2. Create the Agent

An `Agent` combines an ellmer chat with the runtime policy you just defined.
This chunk makes a provider request, so it is displayed but not run while the
vignette is built.

```{r first-agent, eval = FALSE}
chat <- ellmer::chat("openai")

agent <- Agent$new(
  chat = chat,
  tools = first_tools,
  permissions = first_permissions,
  usage_limits = first_limits,
  working_dir = workspace,
  agent_name = "package-scout"
)
```

The chat owns provider interaction. Deputy owns the tool loop around it:
permissions, hooks, limits, events, persistence, checkpoints, and delegation.

## 3. Run one useful task

Ask for an answer that the available tools can support. A read-only package
inspection is a better first task than an open-ended request to “improve the
code.”

```{r first-run, eval = FALSE}
result <- agent$run_sync(
  paste(
    "Read DESCRIPTION and list the R/ directory.",
    "Explain the package purpose and identify three good entry-point files.",
    "Cite the paths you used. Do not propose or make edits."
  ),
  run_context = list(stage = "package-review")
)
```

`run_sync()` blocks until the run stops and returns an `AgentResult`. The result
keeps the outcome and the evidence needed to understand how it happened:

```{r inspect-result, eval = FALSE}
cat(result$response)

result$stop_reason
result$usage
result$session_id
result$run_id
result$tool_calls()
result$tool_results()
```

Check `stop_reason` before treating the response as complete. A request,
tool-call, token, or cost limit produces a typed stop reason rather than a
silently truncated success. `"tool_loop"` means the model requested the same
tool with the same arguments and received the same result three times
consecutively; changing results count as progress. If a cost limit is set but
the provider omits any cost record, Deputy stops with
`"cost_unavailable"`; it never presents a known lower bound as the total.

## 4. Grant writes deliberately

When a task really needs to create a file, replace the read-only policy with an
explicit one. This policy permits Deputy's native file tools to write only
inside the canonical workspace. R execution, shell access, web access, and
package installation remain off.

```{r write-policy}
write_permissions <- Permissions$new(
  mode = "standard",
  file_read = TRUE,
  file_write = workspace,
  bash = FALSE,
  r_code = FALSE,
  web = FALSE,
  install_packages = FALSE
)

stopifnot(
  identical(write_permissions$file_write, workspace),
  identical(write_permissions$r_code, FALSE),
  identical(write_permissions$bash, FALSE)
)
```

Create a separate Agent with file checkpointing enabled. The checkpoint marks
the state before the task. Deputy records preimages when its native write,
edit, and multi-edit tools change files.

```{r write-agent, eval = FALSE}
write_agent <- Agent$new(
  chat = chat$clone()$set_turns(list()),
  tools = tools_file(),
  permissions = write_permissions,
  usage_limits = first_limits,
  enable_file_checkpointing = TRUE,
  working_dir = workspace,
  agent_name = "summary-writer"
)

checkpoint_id <- write_agent$checkpoint("before package summary")

write_result <- write_agent$run_sync(
  paste(
    "Read DESCRIPTION and write a concise package summary to deputy-summary.md.",
    "Do not modify any other file."
  )
)
```

Review the result and the actual diff. If the native file-tool changes are not
useful, rewind them:

```{r rewind, eval = FALSE}
write_result$stop_reason
write_agent$list_checkpoints()
write_agent$rewind_files(checkpoint_id)
```

Checkpoints do not capture changes made outside Deputy's native file tools, and
they do not replace Git or backups.

## Common failures

### The provider cannot authenticate

Configure the credential expected by your chosen ellmer provider, then verify a
plain ellmer chat before adding Deputy. Deputy does not manage provider keys.

### A tool call is denied

Treat the denial as useful policy feedback. Check that the tool belongs in this
task, that it is registered, and that the relevant capability is explicit. Do
not switch to `permissions_full()` merely to make the error disappear.

### The run stops at a limit

Inspect `result$stop_reason` and `result$usage`. Tighten the task first; increase
only the specific limit whose additional work is justified.

### Code execution sounds safer than it is

`run_r_code` starts a separate R process, but process separation is not an OS
sandbox. `run_bash` runs with the current user's privileges. Both are denied by
`permissions_standard()`. For untrusted model-generated R, configure mcp-repl
with an explicit OS policy and load it through `tools_mcp_repl()`; Deputy
refuses to load a server whose effective sandbox differs from the requested
one.

### A Shiny task reads the wrong directory

Shiny process working directories can differ from the app directory. Pass an
existing absolute `working_dir` and construct any directory-valued write grant
from that same path.

## Where next?

| If you want to... | Continue with... |
|---|---|
| Choose or build tools | `vignette("tools", package = "deputy")` |
| Design permission policy | `vignette("permissions", package = "deputy")` |
| Observe and intervene in runs | `vignette("hooks", package = "deputy")` |
| Stream an Agent into an app | `vignette("example-shiny-chat", package = "deputy")` |
| Delegate to specialist Agents | `vignette("multi-agent", package = "deputy")` |
| Require typed model output | `vignette("structured-output", package = "deputy")` |
| Build a concrete workflow | `vignette("example-data-analysis", package = "deputy")` |

The durable pattern is the same in every host: make the workspace, tools,
permissions, and limits explicit; run one bounded job; then inspect the typed
result before granting more authority.


## Standalone scripts

The package includes eight independent scripts for basic runs, tools,
permissions, hooks, delegation, structured output, session resume, and skills.
They are executed in CI with deterministic provider responses. To run one
against OpenAI, set `OPENAI_API_KEY` and source the installed file:

```{r, eval = FALSE}
source(system.file("examples", "standalone", "01-basic.R", package = "deputy"))
```

Find the other scripts and their setup guide with:

```{r, eval = FALSE}
list.files(system.file("examples", "standalone", package = "deputy"))
```
