---
title: "Getting Started with deputy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with deputy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

deputy is an agentic AI framework for R that builds on
[ellmer](https://ellmer.tidyverse.org/). It enables you to create AI agents
that use tools to accomplish multi-step tasks, with built-in support for
permissions, hooks, and streaming output. The core `Agent` and `LeadAgent`
APIs are provider-agnostic.

## Installation

```{r, eval = FALSE}
# Install from GitHub
pak::pak("JamesHWade/deputy")
```

## Quick Start

An agent wraps an ellmer chat object and gives it tools, permissions, and
lifecycle hooks:

```{r getting-started-quickstart}
library(deputy)

chat <- ellmer::chat_anthropic(model = "claude-sonnet-4-20250514")
agent <- Agent$new(chat = chat, tools = tools_file())

result <- agent$run_sync("List the files in the current directory")
cat(result$response)
```

The agent sends your task to the LLM, which may call tools, and loops until
the task is complete. `run_sync()` blocks until done and returns an
`AgentResult`.

## Tool Bundles

deputy organises built-in tools into bundles you can mix and match:

| Bundle | Tools | Purpose |
|--------|-------|---------|
| `tools_file()` | `read_file`, `read_markdown`, `write_file`, `list_files` | File operations |
| `tools_code()` | `run_r_code`, `run_bash` | Code execution |
| `tools_data()` | `read_csv`, `read_file`, `read_markdown` | Data reading |
| `tools_web()` | `web_fetch`, `web_search` | Web access |
| `tools_all()` | All of the above | Everything |

```{r, eval = FALSE}
# Combine bundles
agent <- Agent$new(
  chat = ellmer::chat_anthropic(),
  tools = c(tools_file(), tools_code())
)
```

There are also named presets available via `tools_preset()`:

```{r, eval = FALSE}
tools_preset("dev")
```

See `vignette("tools")` for custom tools, web tools, MCP integration, and
human-in-the-loop.

## Streaming

For real-time feedback, use `run()` which returns a generator:

```{r getting-started-streaming, eval = FALSE}
chat <- ellmer::chat_anthropic(model = "claude-sonnet-4-20250514")
agent <- Agent$new(chat = chat, tools = tools_file())

events <- agent$run("What is the name of this package?")
repeat {
  event <- events()
  if (coro::is_exhausted(event)) {
    break
  }
  switch(event$type,
    "text" = cat(event$text),
    "tool_start" = cli::cli_alert_info("Calling {event$tool_name}..."),
    "tool_end" = cli::cli_alert_success("Done"),
    "stop" = cli::cli_alert("Finished!")
  )
}
```

Events stream as they happen -- you see text tokens arrive, tool calls start
and finish, and a final stop event.

## Permissions

### Read-Only Permissions

```{r}
# Only allows reading files - no writes, no code execution
agent <- Agent$new(
  chat = ellmer::chat("openai"),
  tools = tools_file(),
  permissions = permissions_readonly()
)
```

### Full Permissions

```{r}
# Allows everything - use with caution!
agent <- Agent$new(
  chat = ellmer::chat("openai"),
  tools = tools_all(),
  permissions = permissions_full()
)

# Planning mode: read-only tools plus ask_user
agent <- Agent$new(
  chat = ellmer::chat("openai"),
  tools = tools_all(),
  permissions = permissions_plan()
)
```

### Custom Permissions

For fine-grained control:

```{r}
perms <- Permissions$new(
  file_read = TRUE,
  file_write = "/path/to/allowed/dir", # Restrict to specific directory

  bash = FALSE,
  r_code = TRUE,
  web = FALSE
)

agent <- Agent$new(
  chat = ellmer::chat("openai"),
  permissions = perms
)
```

### Custom Permission Callbacks

For complex permission logic:

```{r}
perms <- Permissions$new(
  can_use_tool = function(tool_name, tool_input, context) {
    # Block any file writes to sensitive directories
    if (tool_name == "write_file") {
      if (grepl("^\\.env|secrets|credentials", tool_input$path)) {
        return(PermissionResultDeny(
          reason = "Cannot write to sensitive files"
        ))
      }
    }
    PermissionResultAllow()
  }
)
```

### Tool Allow/Deny Lists

Use allow/deny lists when you want an explicit tool policy that is separate from
the broader mode flags:

```{r}
perms <- Permissions$new(
  mode = "standard",
  tool_allowlist = c("read_file", "list_files", "run_r_code"),
  tool_denylist = c("run_bash"),
  permission_prompt_tool_name = "ask_user"
)

agent <- Agent$new(
  chat = ellmer::chat("openai"),
  tools = tools_all(),
  permissions = perms
)
```

If both lists are present, `tool_denylist` wins. The
`permission_prompt_tool_name` tool can request explicit human approval and
appears in deny reasons when prompting is enabled.

## Hooks

Hooks let you intercept and customize agent behavior at key points:

### Available Hook Events

| Event | When it Fires | Can Modify |
|-------|---------------|------------|
| `PreToolUse` | Before a tool executes | Allow/deny, input |
| `PostToolUse` | After a tool executes | Continue flag |
| `Stop` | When agent stops | - |
| `UserPromptSubmit` | When user sends a message | - |
| `PreCompact` | Before conversation compaction | Summary |

### Example: Logging All Tool Calls

```{r}
agent$add_hook(HookMatcher$new(
  event = "PostToolUse",
  callback = function(tool_name, tool_result, tool_error, context) {
    cli::cli_alert_info("Tool {tool_name} completed")
    HookResultPostToolUse()
  }
))
```

### Example: Block Dangerous Commands

```{r}
agent$add_hook(HookMatcher$new(
  event = "PreToolUse",
  pattern = "^run_bash$", # Only match bash tool
  callback = function(tool_name, tool_input, context) {
    if (grepl("rm -rf|sudo|chmod 777", tool_input$command)) {
      HookResultPreToolUse(
        permission = "deny",
        reason = "Dangerous command pattern detected"
      )
    } else {
      HookResultPreToolUse(permission = "allow")
    }
  }
))
```

## Session Management

Every `Agent` has a stable `session_id` used to correlate its events and
results. Deputy generates one unless you provide it at construction:

```{r}
agent <- Agent$new(
  chat = ellmer::chat("openai"),
  tools = tools_file(),
  session_id = "session_analysis_001"
)

agent$session_id()
result <- agent$run_sync("Summarize this package")
result$session_id
```

The identifier does not save or locate state. Save and load are explicit,
path-based operations:

```{r}
agent$save_session("analysis-session.rds")

restored <- Agent$new(
  chat = ellmer::chat("openai"),
  tools = tools_file(),
  permissions = permissions_readonly(),
  session_id = "session_analysis_002"
)
restored$load_session("analysis-session.rds")

# Continue the conversation
result <- restored$run_sync("Continue where we left off...")
```

`Agent$load_session()` restores the conversation, prompt, run context, and
enabled file checkpoint state. Tools, permissions, hooks, the working
directory, and the receiving agent's `session_id` remain authoritative.

## Multi-Agent Systems

For complex tasks, you can create a lead agent that delegates to specialized sub-agents:

```{r}
# Define specialized sub-agents
code_agent <- agent_definition(
  name = "code_analyst",
  description = "Analyzes R code and suggests improvements",
  prompt = "You are an expert R programmer. Analyze code for best practices.",
  tools = tools_file()
)

data_agent <- agent_definition(
  name = "data_analyst",
  description = "Analyzes data files and provides statistical summaries",
  prompt = "You are a data analyst. Provide clear statistical insights.",
  tools = tools_data()
)

# Create a lead agent that can delegate
lead <- LeadAgent$new(
  chat = ellmer::chat("openai"),
  sub_agents = list(code_agent, data_agent),
  system_prompt = "You coordinate between specialized agents to complete tasks."
)

# The lead agent will automatically delegate to sub-agents as needed
result <- lead$run_sync(
  "Review the R code in src/ and analyze the data in data/"
)
```

## Working with Results

The `AgentResult` object contains useful information:

```{r}
result <- agent$run_sync("Analyze this project")

# The final response
cat(result$response)

# Cost information
result$cost
#> $input
#> [1] 1250
#> $output
#> [1] 450
#> $total
#> [1] 0.0045

# Execution duration
result$duration
#> [1] 3.45  # seconds

# Stop reason
result$stop_reason
#> [1] "complete"

# Stable session correlation
result$session_id

# All events (for detailed analysis)
length(result$events)
#> [1] 12
```

## Provider Support

deputy works with any LLM provider that ellmer supports:

```{r}
# OpenAI
agent <- Agent$new(chat = ellmer::chat("openai"))

# Anthropic
agent <- Agent$new(chat = ellmer::chat("anthropic/claude-sonnet-4-5-20250929"))

# Google
agent <- Agent$new(chat = ellmer::chat("google/gemini-1.5-pro"))

# Local models via Ollama
agent <- Agent$new(chat = ellmer::chat("ollama/llama3.1"))
```

## Best Practices

1. **Start with minimal permissions** - Use `permissions_readonly()` or
   `permissions_standard()` and only expand as needed.

2. **Use hooks for logging** - Add a `PostToolUse` hook to track what your agent does.
3. **Set run limits** - Use `UsageLimits()` to bound model requests, tool
   calls, tokens, or estimated cost.

4. **Save sessions** - For long-running tasks, save sessions periodically.

5. **Use streaming for UX** - The `run()` method provides real-time feedback.

```{r}
# Example combining best practices
agent <- Agent$new(
  chat = ellmer::chat("openai"),
  tools = tools_file(),
  permissions = permissions_standard(),
  usage_limits = UsageLimits(max_requests = 20, max_cost_usd = 1)
)
```

## Next Steps

- `vignette("tools")` -- Custom tools, web tools, MCP, and human-in-the-loop
- `vignette("permissions")` -- Permission presets, modes, and custom policies
- `vignette("hooks")` -- Lifecycle hooks for logging, blocking, and auditing
- `vignette("multi-agent")` -- Multi-agent delegation with LeadAgent
- `vignette("structured-output")` -- JSON schema output and validation
