---
title: "Hooks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hooks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

A hook is an R function that Deputy calls at a fixed point in a run: before a
tool runs, after it returns, when the run stops, and so on. Use hooks to log
what an agent does, keep an audit trail, block particular tool calls, or stop
a run early.

## Add a hook

Create a hook with `HookMatcher()`, naming the event and the callback, and add
it to an agent:

```{r}
library(deputy)

log_tool <- HookMatcher(
  event = "PostToolUse",
  callback = function(tool_name, tool_result, tool_error, context) {
    cli::cli_inform("{tool_name} finished")
    NULL
  }
)

agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  tools = tools_file(),
  permissions = permissions_readonly()
)
agent$add_hook(log_tool)
```

For tool events, `pattern` is a regular expression that limits the hook to
matching tool names. `hook_matches()` tells you whether a hook applies to a
tool:

```{r, eval = TRUE}
library(deputy)

watch_writes <- HookMatcher(
  event = "PreToolUse",
  pattern = "^(write_file|edit_file|multi_edit)$",
  callback = function(tool_name, tool_input, context) {
    cli::cli_inform("{tool_name}: {tool_input$path}")
    NULL
  }
)
hook_matches(watch_writes, "edit_file")
hook_matches(watch_writes, "read_file")
```

## Events

| Event | Fires | Callback arguments |
|---|---|---|
| `PreToolUse` | before a tool runs | `tool_name, tool_input, context` |
| `PostToolUse` | after a tool returns | `tool_name, tool_result, tool_error, context` |
| `PostToolUseFailure` | after a tool fails | `tool_name, tool_result, tool_error, context` |
| `PermissionRequest` | when the permission policy denies a call | `tool_name, tool_input, permission_result, context` |
| `SessionStart` | at the start of each run | `context` |
| `UserPromptSubmit` | at the start of each run, with the task | `prompt, context` |
| `Stop` | at the end of each run | `reason, context` |
| `SessionEnd` | at the end of each run, after `Stop` | `reason, context` |
| `SubagentStart` | when a subagent starts a delegated task | `agent_name, task, context` |
| `SubagentStop` | when a subagent finishes | `agent_name, task, result, context` |
| `PreCompact` | before older turns are summarised | `turns_to_compact, turns_to_keep, context` |
| `PostCompact` | after a summary is installed | `result, context` |
| `Notification` | when Deputy reports something, such as a denied call | `message, context` |
| `ConfigChange` | when `set_permission_mode()` changes the mode | `key, old_value, new_value, context` |

`context` is a named list with the working directory, the agent and run
identifiers, and fields specific to the event, such as the tool's annotations
for `PreToolUse` or the run's usage for `Stop`. `?HookEvent` lists them all.

## Block a tool call

A `PreToolUse` hook runs after the permission policy has allowed a call, and
can still deny it. Return `HookResultPreToolUse()`:

```{r}
no_secrets <- HookMatcher(
  event = "PreToolUse",
  pattern = "^write_file$",
  callback = function(tool_name, tool_input, context) {
    if (grepl("\\.env$|secrets", tool_input$path)) {
      return(HookResultPreToolUse("deny", reason = "Don't write to secret files."))
    }
    NULL
  }
)
agent$add_hook(no_secrets)
```

The model sees the reason and can try something else. Hooks can only take
permissions away: a hook that returns `"allow"` doesn't override a denial from
the policy.

## Stop a run from a hook

`PostToolUse` hooks see each tool's result or error. Return
`HookResultPostToolUse(continue = FALSE)` to end the run after the current
tool:

```{r}
stop_on_error <- HookMatcher(
  event = "PostToolUse",
  callback = function(tool_name, tool_result, tool_error, context) {
    if (!is.null(tool_error)) {
      cli::cli_alert_danger("{tool_name} failed: {tool_error}")
      return(HookResultPostToolUse(continue = FALSE, stop_reason = "tool_failed"))
    }
    NULL
  }
)
```

`HookResultPreToolUse()` also takes `continue` and `stop_reason`. Both take
`additional_context`, text that Deputy appends to the system prompt.

## Built-in hooks

`hook_log_tools()` prints a line after each tool call saying whether it
succeeded. `verbose = TRUE` adds a short preview of the result:

```{r}
agent$add_hook(hook_log_tools(verbose = TRUE))
```

`hook_block_dangerous_bash()` denies shell commands that match patterns such
as `rm -rf`, `sudo` and `chmod 777`. Add your own patterns with
`additional_patterns`:

```{r}
agent$add_hook(hook_block_dangerous_bash(
  additional_patterns = c("DROP\\s+TABLE", "TRUNCATE")
))
```

A pattern list catches accidents, not a determined adversary, so don't rely on
it to make `run_bash` safe.

`hook_limit_file_writes(dir)` denies writes and edits outside `dir`, applying
the same path rules as `Permissions(file_write = dir)`. Use it to add a second
check, or to confine writes more tightly for one agent than its policy does.

## Order and errors

Hooks for an event run in the order you added them, and the first hook that
returns something other than `NULL` decides. Later hooks for that event don't
run. So:

- return `NULL` from hooks that only observe, so the hooks after them still run;
- add hooks that can deny a call before hooks that might allow it;
- add the built-in hooks after your own hooks for the same event. They always
  return a result, even when they have no objection, so a hook added after
  `hook_block_dangerous_bash()` never sees `run_bash` calls.

If a `PreToolUse` hook throws an error, Deputy denies the tool call. An error
in any other hook is reported and the run carries on. `agent$hooks$last_errors()`
lists recent hook errors.

## Run a hook in a separate process

Hooks run in your R session by default, so they can read and update your
objects. Give `HookMatcher()` a positive `timeout` (in seconds) to run the
callback in a fresh R process that is killed if it takes too long. That
process can't see your session's objects, so refer to package functions with
`::`, for example `deputy::HookResultPreToolUse()`.

## Hooks or results?

Hooks are for reacting while a run is in progress. To analyse a run afterwards,
use the `AgentResult` instead: `result_tool_calls()`, `result_tool_results()`
and `result$events` hold the same information, without any hook state to
manage.

[Human input and approvals](approvals.html) shows hooks that ask a person to
approve a tool call before it runs.
