---
title: "Hooks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hooks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Hooks let you intercept agent behaviour at key points in the execution
lifecycle. Use them for logging, auditing, blocking dangerous actions, or
injecting custom logic.

## Hook Events

deputy fires hooks at these events:

| Event | Callback Signature | Purpose |
|-------|-------------------|---------|
| `PreToolUse` | `function(tool_name, tool_input, context)` | Before a tool runs (can allow/deny) |
| `PostToolUse` | `function(tool_name, tool_result, tool_error, context)` | After a tool completes |
| `Notification` | `function(message, context)` | Informational runtime notices |
| `Stop` | `function(reason, context)` | When the agent finishes |
| `SubagentStop` | `function(agent_name, task, result, context)` | When a sub-agent finishes |
| `UserPromptSubmit` | `function(prompt, context)` | When user submits input |
| `PreCompact` | `function(turns_to_compact, turns_to_keep, context)` | Before conversation compaction |
| `SessionStart` | `function(context)` | When a session starts |
| `SessionEnd` | `function(reason, context)` | When a session ends |

## Creating Hooks

Hooks are created with `HookMatcher$new()`:

```{r, eval = FALSE}
library(deputy)

hook <- HookMatcher$new(
  event = "PostToolUse",
  callback = function(tool_name, tool_result, tool_error, context) {
    cli::cli_alert_info("Tool {tool_name} completed")
    HookResultPostToolUse()
  }
)
```

Add hooks to an agent with `add_hook()`:

```{r, eval = FALSE}
agent$add_hook(hook)
```

Hooks run in the caller's R process by default, so callbacks can use ordinary
R state and side effects. Set a positive `timeout` only when you deliberately
want a clean `callr` subprocess with a hard deadline. An isolated callback
cannot rely on caller-process state; qualify package functions such as
`deputy::HookResultPostToolUse()` when using that mode.

### Filtering by Tool Name

The optional `pattern` argument is a regex that filters which tools the hook
applies to:

```{r, eval = FALSE}
# Only fires for bash commands
HookMatcher$new(
  event = "PreToolUse",
  pattern = "^run_bash$",
  callback = function(tool_name, tool_input, context) {
    cli::cli_alert_warning("Bash command: {tool_input$command}")
    HookResultPreToolUse(permission = "allow")
  }
)
```

## Pre-Built Hooks

deputy includes several ready-made hooks:

### Logging Tool Calls

`hook_log_tools()` logs every tool call using cli:

```{r hooks-log-demo}
library(deputy)

chat <- ellmer::chat_openai(model = "gpt-5.6-luna")
agent <- Agent$new(
  chat = chat,
  tools = tools_file(),
  permissions = permissions_readonly()
)
agent$add_hook(hook_log_tools(verbose = TRUE))

result <- agent$run_sync("What files are in the current directory?")
```

### Blocking Dangerous Bash Commands

`hook_block_dangerous_bash()` blocks patterns like `rm -rf`, `sudo`,
`chmod 777`, and more:

```{r, eval = FALSE}
agent$add_hook(hook_block_dangerous_bash())

# Optionally add your own patterns
agent$add_hook(hook_block_dangerous_bash(
  additional_patterns = c("DROP\\s+TABLE", "TRUNCATE")
))
```

### Limiting File Writes

Configure the Agent's `Permissions` when a directory is an authority boundary.
`hook_limit_file_writes()` is a defense-in-depth convenience hook that delegates
to the same canonical path policy for `write_file`, `edit_file`, and
`multi_edit`. The directory must already exist:

```{r, eval = FALSE}
output_dir <- file.path(getwd(), "output")
dir.create(output_dir, showWarnings = FALSE)

agent$add_hook(hook_limit_file_writes(output_dir))
```

## Custom PreToolUse Hooks

`PreToolUse` hooks can allow or deny tool calls. Return
`HookResultPreToolUse()` with `permission = "allow"` or `"deny"`:

```{r, eval = FALSE}
hook_no_secrets <- HookMatcher$new(
  event = "PreToolUse",
  pattern = "^write_file$",
  callback = function(tool_name, tool_input, context) {
    path <- tool_input$path
    if (is.null(path)) {
      path <- ""
    }
    if (grepl("\\.env$|secrets", path)) {
      HookResultPreToolUse(
        permission = "deny",
        reason = "Cannot write to secret files"
      )
    } else {
      HookResultPreToolUse(permission = "allow")
    }
  }
)
```

## Custom PostToolUse Hooks

`PostToolUse` hooks run after a tool completes. Use them for logging,
metrics, or conditional stopping:

```{r hooks-audit-demo}
hook_audit <- HookMatcher$new(
  event = "PostToolUse",
  callback = function(tool_name, tool_result, tool_error, context) {
    if (!is.null(tool_error)) {
      cli::cli_alert_danger("{tool_name} failed: {tool_error}")
    } else {
      cli::cli_alert_success("{tool_name} completed")
    }
    HookResultPostToolUse()
  }
)

chat <- ellmer::chat_openai(model = "gpt-5.6-luna")
agent <- Agent$new(
  chat = chat,
  tools = tools_file(),
  permissions = permissions_readonly()
)
agent$add_hook(hook_audit)

result <- agent$run_sync("What files are in the current directory?")

# For post-hoc analysis, use AgentResult instead of hook state:
result$tool_calls()
```

Set `continue = FALSE` to stop the agent after a tool call:

```{r, eval = FALSE}
HookResultPostToolUse(continue = FALSE)
```

## Session Lifecycle Hooks

Session hooks fire at the start and end of a session:

```{r, eval = FALSE}
HookMatcher$new(
  event = "SessionStart",
  callback = function(context) {
    message("Session started at ", Sys.time())
    NULL
  }
)
```

## Notification Hooks

`Notification` hooks are useful for informational events that should not alter
control flow, such as permission-denied guidance, session-load notices, or
compaction fallbacks:

```{r, eval = FALSE}
agent$add_hook(HookMatcher$new(
  event = "Notification",
  callback = function(message, context) {
    cli::cli_alert_info("[{context$code}] {message}")
    NULL
  }
))
```

## Error Handling in Hooks

If a `PreToolUse` hook throws an error, Deputy fails closed and denies the tool
call. Other hook errors are logged and the run continues. You can inspect
recent hook errors with:

```{r, eval = FALSE}
agent$hooks$last_errors()
```

## Approval before a tool call

The installed recipe defines `approval_gate()` and `approval_after_install()`:

```{r approval-recipe-load, eval = TRUE}
source(system.file("examples", "approval-gates.R", package = "deputy"))
```

`approval_gate()` calls the `ask_user` tool from `tools_interactive()` directly
inside a `PreToolUse` callback. The handler belongs to this hook instance; it
does not use the process-wide callback. The question includes the exact tool
name and arguments. Only the exact answer `"Approve"` permits execution;
missing answers, cancellation, and handler errors cannot approve a call.

Here is an executable dry run with a host that declines the request:

```{r approval-simple-demo, eval = TRUE}
decline <- function(questions, context) {
  setNames(list("Deny"), questions[[1]]$question)
}
gate <- approval_gate(callback = decline)
gate$callback(
  "write_file", list(path = "report.txt", content = "Draft"),
  context = list(run_id = "example-run")
)$permission
```

Register approval hooks **before all other matching hooks**, including audit
and Stop hooks. The registry returns the first non-NULL result, so an earlier
allow or audit result can bypass a later hook. These recipes return NULL after
approval, for unarmed gates, and after recording receipts so later hooks still
run and can deny the call.

Attach a fresh hook to an Agent whose permissions already allow the relevant
operation. Hooks can narrow permissions; approval cannot widen that ceiling.
In an interactive R terminal, omit `callback` to use the normal user prompt.
In a hosted application, supply your own synchronous human-input handler.
This recipe is not a durable pause/resume mechanism.

```{r, eval = FALSE}
agent$add_hook(approval_gate(callback = my_approval_handler))
```

## Approval after an earlier installation

Hook context supplies `run_id`, not a mutable history of earlier calls.
`approval_after_install()` accumulates successful installation receipts in its
own environment, keyed by that run ID. Its `PreToolUse` callback reads that
state before `push_changes`; `Stop` removes it. Create a fresh set per Agent.
The callbacks use `timeout = 0` so they share caller-process state.

This recipe assumes dedicated `install_dependency` and `push_changes` tools.
The install tool must return `list(installed = TRUE)` only after success.
A failed install does not arm the gate. Matching arbitrary shell text would
not reliably identify these operations.

```{r approval-stateful-demo, eval = TRUE}
hooks <- approval_after_install(callback = decline)
run <- list(run_id = "installation-run")
hooks[[1]]$callback(
  "install_dependency", list(installed = TRUE), NULL, run
)
hooks[[2]]$callback("push_changes", list(remote = "origin"), run)$permission
# A different run has no installation receipt.
hooks[[2]]$callback(
  "push_changes", list(remote = "origin"), list(run_id = "another-run")
)
hooks[[3]]$callback("end_turn", run)
```

Attach all three hooks together:

```{r, eval = FALSE}
for (hook in approval_after_install(callback = my_approval_handler)) {
  agent$add_hook(hook)
}
```

The example returns `NULL` before an installation because it demonstrates
that particular sequence policy. Use the unconditional gate if every push
needs approval. Neither recipe grants permission or performs an installation
or a push by itself.
