---
title: "MCP servers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MCP servers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

[Model Context Protocol](https://modelcontextprotocol.io/) (MCP) servers
provide tools, documents ("resources") and prompt templates to any client that
speaks the protocol. Deputy connects to them through the
[mcptools](https://posit-dev.github.io/mcptools/) package, and treats their
tools like any other: every call goes through the agent's permissions, hooks
and usage limits.

Configure servers in mcptools' configuration file,
`~/.config/mcptools/config.json` by default. The mcptools documentation
explains the format.

Deputy works with mcptools 1.0.2 and 1.0.3, and refuses other versions until
they have been tested. It relies on parts of mcptools that aren't yet public
API.

## Load tools from a server

`tools_mcp()` connects to the servers you name and returns their tools:

```{r}
library(deputy)

agent <- Agent$new(
  chat = ellmer::chat("anthropic/claude-sonnet-5"),
  tools = c(tools_file(), tools_mcp(servers = "github"))
)
```

Name servers exactly as they appear in the config file. If mcptools isn't
installed or the servers offer no tools, `tools_mcp()` warns and returns an
empty list. `agent$load_mcp()` does the same for an existing agent, and
`agent$mcp_status()` reports each load attempt.

Deputy keeps the annotations each server declares for its tools, so
permissions can judge them. Tools without annotations get the cautious
defaults described in [Tools](tools.html): with the default permissions, a
tool that might reach the network needs `web = TRUE`. A server's tool is never
confused with a built-in one: a remote tool called `read_file` is judged by its
own annotations, not given Deputy's file permissions, and file checkpoints
don't cover MCP tools. Deputy can't see what a server does with a call, so
connect only to servers you trust.

When a server restarts or reconnects, its old tools stop working. Reload them
with `agent$load_mcp(servers = "github", replace = TRUE)`, which also removes
tools the server no longer offers. Tools from other servers are left alone.

## A connection per conversation

`tools_mcp()` shares one connection per server across your R session. When
each conversation needs its own connection, for example in a Shiny app, or
when you want resources and prompts as well as tools, use `McpConnection`. It
belongs to one agent, and you choose exactly which tools, resources and
prompts it exposes:

```{r}
agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  permissions = Permissions(web = TRUE)
)

connection <- McpConnection$new(
  config = "~/.config/mcptools/config.json",
  server = "evidence",
  agent = agent,
  tools = "inspect_evidence",
  resources = "evidence://report/current",
  prompts = "summarize"
)

agent$register_tools(c(connection$tools(), connection$capability_tools()))
```

`$tools()` returns the allowed tools. `$capability_tools()` returns tools for
reading the allowed resources and prompts; with several connections on one
agent, give each a distinct `prefix` so their names don't clash. These tools
need `web = TRUE`, because the content comes from outside R.

The allow lists can't change after the connection is created. Your code can
browse what the server offers with `$discover("tools")` (or `"resources"`,
`"resource_templates"`, `"prompts"`), but discovering an item doesn't allow it.
`$read_resource(uri)` and `$get_prompt(name, arguments)` fetch allowed items
directly; a fetched prompt is returned to you, not added to any chat.

Calls return promises, and a connection handles one request at a time: a
second request while one is running fails with a busy error, while other
connections carry on. Call `$close()` when the conversation ends, for example
from `session$onSessionEnded()` in Shiny. `$cancel()` kills the connection and
discards the server's session state, as does a request timeout. Tools from a
closed connection stop working; create a new connection to continue.

### Slow servers

mcptools waits about four seconds for each reply from a server that runs as a
local process, and doesn't check that a reply belongs to the request it
answers. A late reply could therefore become the answer to the next call.
Deputy checks every reply's ID, and when a server doesn't answer in time or
answers out of order, the call fails with a `deputy_mcp_desynchronized` error
and the connection closes. The server's session state is lost, so create a new
connection. Tools loaded with `tools_mcp()` stop that server on the first lost
reply.

## Sandboxed interpreters

mcp-repl and MCP Console are MCP servers that run R (and, for MCP Console,
Python and SQL) inside an OS sandbox. Deputy has dedicated helpers for them
that check the sandbox is really on: see [Running R code](code-execution.html).
