---
title: "Multi-Agent Orchestration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Agent Orchestration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

For complex tasks, deputy supports a **lead agent** that delegates work to
specialised **sub-agents**. Each sub-agent has its own tools, prompt, and
(optionally) model. The lead agent decides when to delegate and synthesises
the results.

## When to Use Multi-Agent

Multi-agent orchestration is useful when:

- Different parts of a task need different tools (e.g., code analysis vs.
  data analysis)
- You want to isolate sub-tasks with their own system prompts
- Sub-agents should run with restricted permissions
- You need audit trails per sub-agent via `SubagentStop` hooks

For simpler tasks, a single `Agent` with the right tools is usually enough.

## Defining Sub-Agents

Use `agent_definition()` to describe a sub-agent:

```{r, eval = FALSE}
library(deputy)

code_reviewer <- agent_definition(
  name = "code_reviewer",
  description = "Reviews R code for best practices and potential issues",
  prompt = "You are an expert R developer. Review code for correctness,
    style, and potential bugs. Be specific and actionable.",
  tools = tools_file()
)

data_analyst <- agent_definition(
  name = "data_analyst",
  description = "Analyses data files and produces statistical summaries",
  prompt = "You are a data analyst. Read data files and provide clear,
    concise statistical summaries with key insights.",
  tools = tools_data()
)
```

Fields:

| Field | Description |
|-------|-------------|
| `name` | Unique identifier (used by the lead to delegate) |
| `description` | What this agent does (shown to the lead LLM) |
| `prompt` | System prompt for the sub-agent |
| `tools` | Tools available to the sub-agent |
| `model` | `"inherit"` (default) or a specific model name |
| `skills` | Optional list of skills to load |
| `disallowed_tools` | Additional tool names that the child must not use |
| `memory` | Optional context appended to the child's system prompt |
| `mcp_servers` | Optional MCP servers loaded for the child |
| `initial_prompt` | Optional text prepended to delegated tasks |
| `max_requests` | Optional per-child request limit |
| `permission_mode` | Optional mode that preserves or narrows the lead policy |

### Delegated Permission Boundaries

A sub-agent cannot gain authority that its lead does not have. Deputy
intersects the requested child mode with the lead's capabilities, write root,
tool gates, and permission callback. `disallowed_tools` adds further denials.

The allowed child-mode transitions use the same authority order as
`Agent$set_permission_mode()`:

| Lead mode | Allowed child modes |
|-----------|---------------------|
| `"readonly"` | `"readonly"` |
| `"standard"` | `"standard"`, `"readonly"` |
| `"plan"` | `"plan"`, `"readonly"` |
| `"full"` | `"full"`, `"standard"`, `"plan"`, `"readonly"` |

For example, a readonly reviewer can be declared beneath a standard lead
without gaining file reads, web access, or tools that the lead explicitly
denied:

```{r, eval = FALSE}
reviewer <- agent_definition(
  name = "reviewer",
  description = "Reviews files without changing them",
  prompt = "Review the requested files and report findings.",
  tools = tools_file(),
  permission_mode = "readonly",
  disallowed_tools = "read_csv",
  max_requests = 5
)
```

## Creating a LeadAgent

`LeadAgent` extends `Agent` with a built-in `delegate_to_agent` tool:

```{r, eval = FALSE}
lead <- LeadAgent$new(
  chat = ellmer::chat_anthropic(),
  sub_agents = list(code_reviewer, data_analyst),
  system_prompt = "You coordinate analysis tasks. Delegate to the
    appropriate specialist and synthesise their findings."
)

lead$available_sub_agents()
#> [1] "code_reviewer" "data_analyst"
```

## Running a Delegation Task

When the lead agent decides to delegate, it calls `delegate_to_agent`
internally. The sub-agent runs to completion and returns its result to the
lead:

```{r multi-agent-demo}
library(deputy)

code_reviewer <- agent_definition(
  name = "code_reviewer",
  description = "Reviews R code and suggests improvements",
  prompt = "You are an R code reviewer. Be concise.",
  tools = tools_file()
)

chat <- ellmer::chat_anthropic(model = "claude-sonnet-4-20250514")
lead <- LeadAgent$new(
  chat = chat,
  sub_agents = list(code_reviewer)
)

result <- lead$run_sync(
  "Ask the code reviewer to look at the DESCRIPTION file and
   summarise what this package does."
)
cat(result$response)
```

Direct synchronous delegation participates in the lead run's usage limits. Each
child inherits the lead's remaining `UsageLimits`, and its usage is aggregated
into `result$usage` and enforced by the lead after delegation. This accounting
does not yet cover parallel or background children, transitive child trees, or
cross-run global limits.

## Monitoring with SubagentStop Hooks

Use a `SubagentStop` hook to log or inspect sub-agent results:

```{r, eval = FALSE}
hook_monitor <- HookMatcher$new(
  event = "SubagentStop",
  callback = function(agent_name, task, result, context) {
    cli::cli_alert_info("Sub-agent {agent_name} finished")
    cli::cli_alert("Requests: {result$usage$requests}")
    NULL
  }
)

lead$add_hook(hook_monitor)
```
