---
title: "Subagents"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Subagents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

A `LeadAgent` can hand parts of a task to subagents: specialists with their
own prompt, tools, model and permissions. Each delegated task runs in a fresh
conversation, so a subagent sees only the brief the lead writes for it, and its
tool calls don't fill up the lead's context. The lead gets the subagent's
answer back and decides what to do next.

Subagents help when parts of a task need different tools or instructions, when
some work should run with tighter permissions, or when you want several
independent opinions. For a task one agent can handle with the right tools, one
agent is simpler and cheaper.

## Define subagents

`agent_definition()` describes a subagent. The lead's model reads the
`description` to decide when to use it:

```{r}
library(deputy)

code_reviewer <- agent_definition(
  name = "code_reviewer",
  description = "Reviews R code for bugs and unclear style.",
  prompt = "You review R code. Report specific problems with file and line.",
  tools = tools_file(),
  permission_mode = "readonly"
)

data_analyst <- agent_definition(
  name = "data_analyst",
  description = "Summarises data files.",
  prompt = "You read data files and summarise them in a few sentences.",
  tools = tools_data()
)
```

| Field | Meaning |
|---|---|
| `name` | routing name; converted to lower case, starts with a letter, then letters, digits, `_` or `-` |
| `description` | what the subagent is for, shown to the lead |
| `prompt` | the subagent's system prompt |
| `tools` | the subagent's tools; it doesn't inherit the lead's |
| `model` | which model to use (see below) |
| `skills` | skills to load into the subagent |
| `disallowed_tools` | tool names the subagent may never call |
| `memory` | text added to the subagent's system prompt |
| `mcp_servers` | MCP servers for the subagent (see [Briefing and inspecting subagents](delegation.html)) |
| `initial_prompt` | text put in front of every task the subagent receives |
| `max_requests` | model request limit for each delegated task |
| `permission_mode` | a mode at least as strict as the lead's |

Definitions are read-only values:

```{r, eval = TRUE}
library(deputy)

reviewer <- agent_definition(
  "Reviewer",
  "Reviews files without changing them.",
  "Review the files you are given and report problems.",
  max_requests = 5
)
reviewer$name
reviewer$max_requests
```

## Create a lead agent

`LeadAgent` is an `Agent` with one extra tool, `delegate_to_agent`, and a list
of subagents the model can choose from:

```{r}
lead <- LeadAgent$new(
  chat = ellmer::chat("anthropic/claude-sonnet-5"),
  sub_agents = list(code_reviewer, data_analyst),
  system_prompt = "You coordinate analysis tasks. Delegate to the right
    specialist, then combine their findings."
)

lead$available_sub_agents()
#> [1] "code_reviewer" "data_analyst"

result <- lead$run_sync(
  "Ask the code reviewer to read R/utils.R and summarise its problems."
)
result$response
```

Add a subagent later with `lead$register_sub_agent(definition)`, which also
updates the list of subagents in the lead's system prompt.

When the lead delegates, the subagent runs to completion and its answer comes
back as the result of the `delegate_to_agent` call. When the model delegates
several tasks in one turn, they can run concurrently.

### Choose each subagent's model

`model` accepts three forms:

- `"inherit"`, the default, uses a copy of the lead's chat: same provider,
  model and settings, without the lead's conversation or tools.
- A bare model id such as `"gpt-6-luna"` uses the lead's provider, endpoint
  and credentials with a different model, for example a cheaper one for simple
  work.
- `"provider/model"`, such as `"anthropic/claude-sonnet-5"`, creates a new
  chat with `ellmer::chat()`. Use this form for model ids that contain `/`.

## Permissions and budgets

A subagent can never do more than its lead. Deputy combines the subagent's
`permission_mode` with the lead's policy, keeping the lead's capabilities,
write directory, allow and deny lists, and callback, and adds the definition's
`disallowed_tools`. The allowed modes follow the same order as
`set_permission_mode()`:

| Lead mode | Subagent may use |
|---|---|
| `"readonly"` | `"readonly"` |
| `"standard"` | `"standard"`, `"readonly"` |
| `"plan"` | `"plan"`, `"readonly"` |
| `"full"` | any mode |

A lead in read-only or plan mode can still delegate, because its subagents
can't use a less strict mode. A lead created with `permissions_readonly()` and
all its subagents can read files but can't write them or run code.

The lead's `PreToolUse`, `PostToolUse` and `PostToolUseFailure` hooks also run
for subagent tool calls, and each subagent call is checked against the lead's
current permissions, so narrowing the lead mid-run narrows its subagents too.

Subagents spend from the lead run's budget. Each delegated task gets whatever
is left of the lead's `UsageLimits`, further capped by the definition's
`max_requests`, and its usage is added to the lead's `result$usage`. When the
model delegates several tasks at once, each reserves its share before it
starts, so together they can't overspend.

A few things subagents can't do. They can't use tools that run on the
provider's side, because Deputy can't check those calls. They can't pause for
a durable approval: a lead created with `approval_dir` refuses to delegate, and
a subagent permission callback that returns `PermissionResultPending()` denies
the call. And they share the lead's working directory and file checkpoints;
nothing about delegation isolates the file system.

## Ask several subagents at once

To get independent answers to the same question, call `parallel_delegate()`
yourself instead of waiting for the model to delegate. Each named definition
gets the task in a fresh conversation and answers with a single model request.
The lead makes no request of its own, and its conversation is unchanged.

```{r}
lead <- LeadAgent$new(
  ellmer::chat("openai/gpt-6-luna"),
  sub_agents = list(
    agent_definition("benefits", "Finds benefits", "Make the strongest case for."),
    agent_definition("risks", "Finds risks", "Make the strongest case against.")
  )
)

batch <- lead$parallel_delegate(
  c(
    benefits = "Should our team adopt a four-day week?",
    risks = "Should our team adopt a four-day week?"
  ),
  max_active = 2,
  usage_limits = UsageLimits(max_requests = 2)
)

batch$status
batch$results$benefits$response
batch$run$usage
```

The subagents in a batch can't have tools, skills or MCP servers. The result
keeps the order of the tasks:

| Field | Contents |
|---|---|
| `status` | `"completed"`, `"failed"`, `"stopped"` or `"not_started"` for each task |
| `results` | each subagent's `AgentResult`, or `NULL` |
| `outcomes` | each subagent's [compact outcome](delegation.html#what-a-delegation-returns) |
| `errors` | each failure's condition, or `NULL` |
| `run` | an `AgentResult` for the whole batch, with the combined usage |

One failure doesn't discard the other answers, so check `status` before using
them. `max_active` sets how many run at once; the batch works through the tasks
in waves of that size. Request limits are reserved before each wave starts, and
token and cost limits are split across it, so a wave can overshoot a token or
cost limit by up to one response per subagent. With the default
`on_exceed = "stop"`, a limit returns the answers so far.

In Shiny, use `parallel_delegate_async()`, which returns a promise.
`lead$interrupt()` stops tasks that haven't started and asks running ones to
stop.

For a complete example that compares two opposing answers and asks a separate
moderator to weigh them, run the bundled debate script with your OpenAI key
set:

```{r}
source(system.file("examples", "standalone", "09-debate.R", package = "deputy"))
```

## See what the subagents did

`lead$list_subagents()` returns a data frame with one row per delegated task,
including tasks still running: the subagent's name, its `status`, the exact
`stop_reason`, timestamps and identifiers.

| Status | Meaning |
|---|---|
| `queued`, `running` | not finished yet |
| `completed` | the subagent's run ended with `"complete"` |
| `stopped` | a limit or an interruption ended the run |
| `failed` | the subagent errored |
| `not_started` | the task never started |

`"completed"` means the run ended normally. It doesn't mean the subagent did
the task well; read its answer.

`lead$get_subagent_results()` returns each task's `AgentResult`, and
`lead$get_subagent_messages()` returns each subagent's conversation, including
what it managed before a failure. Filter both by `agent_name`, or by
`delegation_id` and `session_id` respectively. Reading them doesn't add
anything to the lead's context. These records live in the `LeadAgent` object
and disappear with it.

To react while delegations happen, add `SubagentStart` and `SubagentStop`
hooks to the lead:

```{r}
lead$add_hook(HookMatcher(
  event = "SubagentStop",
  callback = function(agent_name, task, result, context) {
    cli::cli_inform("{agent_name} finished: {context$status}")
    NULL
  }
))
```

These accessors are for your own code. Before you show subagent conversations
to users, read [Briefing and inspecting subagents](delegation.html), which
covers authorising and redacting what each user may see.

## Keep definitions in files

Definitions can live as YAML files, conventionally in `.deputy/agents/`, so
they can be reviewed and versioned like any other configuration:

```yaml
version: 1
name: reviewer
description: Reviews local text for gaps
prompt: |
  Read the supplied text using read_file.
  Report unsupported claims and missing evidence concisely.
tools: [read_file]
model: inherit
disallowed_tools: [write_file, run_bash]
initial_prompt: Keep the review grounded in the supplied text.
max_requests: 2
permission_mode: readonly
```

Only `version`, `name`, `description` and `prompt` are required; the other
fields default as in `agent_definition()`. A file can't contain R code. Tools
and skills are named by keys that your code maps to real objects, so a file can
only use the tools you offer it:

```{r definition-file-read, eval = TRUE}
tool_registry <- list(read_file = tool_read_file)
definition_dir <- system.file("examples", "agent-definitions", package = "deputy")

definitions <- agent_definitions(definition_dir, tools = tool_registry)
names(definitions)
```

`agent_definitions()` reads every `.yaml` and `.yml` file in a directory
(`.deputy/agents/` by default), in file name order. A missing directory gives
an empty list; an invalid file, an unknown key or a duplicate name is an
error. Reading files never runs code, loads skills, or connects to servers.

`agent_definition_read()` reads one file, and `agent_definition_write()` writes
a definition back, given the same registries:

```{r definition-file-roundtrip, eval = TRUE}
path <- tempfile(fileext = ".yaml")
agent_definition_write(definitions$reviewer, path, tools = tool_registry)
restored <- agent_definition_read(path, tools = tool_registry)
identical(restored, definitions$reviewer)
```

Unknown fields, unsupported versions and YAML `!expr` tags are errors. Quote
strings such as `"yes"` and `"123"` that YAML would otherwise convert. Files
are limited to 1 MiB, and a file can't give a subagent more permissions than
its lead.

## Going further

- [Briefing and inspecting subagents](delegation.html): pass structured briefs
  and evidence, give each subagent its own resources, and show subagent
  conversations to users safely.
- [Retained specialists and agent graphs](retained-agents.html): keep a
  specialist's conversation across several tasks, and let specialists delegate
  to each other.
- [Multi-agent code review](example-code-review.html): a complete example.
