---
title: "Multi-Agent Orchestration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Agent Orchestration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

For complex tasks, deputy supports a **lead agent** that delegates work to
specialised **sub-agents**. Each sub-agent has its own tools, prompt, and
(optionally) model. The lead agent decides when to delegate and synthesises
the results.

## When to Use Multi-Agent

Multi-agent orchestration is useful when:

- Different parts of a task need different tools (e.g., code analysis vs.
  data analysis)
- You want to isolate sub-tasks with their own system prompts
- Sub-agents should run with restricted permissions
- You need audit trails per sub-agent via `SubagentStop` hooks

For simpler tasks, a single `Agent` with the right tools is usually enough.

## Composing retained specialists

For existing, independently configured conversations, use an ordinary `Agent`
as the root and `retain_agent_graph()` to declare the permitted routes. The
following assumes `root`, `analyst` and `reviewer` are distinct Agents whose
permissions already allow the relevant tools:

```{r}
handles <- root$retain_agent_graph(
  agents = list(analyst = analyst, reviewer = reviewer),
  routes = list(
    root = list(analyze = list(
      target = "analyst", description = "Analyze the evidence",
      usage_limits = UsageLimits(max_requests = 4)
    )),
    analyst = list(review = list(
      target = "reviewer", description = "Check the analysis",
      usage_limits = UsageLimits(max_requests = 2)
    ))
  ),
  usage_limits = UsageLimits(max_requests = 12, max_tool_calls = 8),
  max_depth = 2, max_delegations = 8, max_concurrency = 2
)
result <- root$run_sync("Analyze the evidence and have it reviewed.")
root$delegation_graph_usage()
```

The root owns all retained conversations and descendant records. Each route
fixes its target and allocation; the model supplies only a new task brief.
Provider configuration, prompts and executable tools stay with each specialist.
Policies and tool hooks from every ancestor also govern descendant tool calls.
A child cannot grant itself more authority than its callers.

The graph budget is cumulative across root runs and explicit host follow-ups.
Depth counts edges from the root. Concurrency includes parents waiting for their
children, so this three-level route requires two child slots. At capacity calls
reject before dispatch. Cyclic routes can be configured, but a busy conversation
cannot be reentered. Token and cost ceilings use reported responses; requests
already in flight can overshoot them. A request ceiling bounds dispatch itself.

```{r}
# Host controls require application authorization independently of disclosure.
root$continue_agent(handles[["analyst"]], "Revisit the conclusion",
  UsageLimits(max_requests = 1))
rows <- root$list_subagents()
# With an explicitly configured DelegationDisclosure policy:
views <- root$inspect_subagents(requester, transcript = TRUE)
reader <- root$observe_subagents(requester)
reader$poll() # ordered root events with parent/depth metadata
reader$close()
root$interrupt_subagent(rows$delegation_id[[1]], "host_cancelled")
# Await outstanding continuations before releasing all members together.
root$release_agent_graph()
```

Observation and selected-child panels never consume a producer stream or inject
child transcripts into the root's model context. Cancellation cooperatively stops
a subtree; siblings remain independently inspectable. Release removes graph
routes and ownership, while borrowed tools and external resources remain the
host's responsibility. These handles do not survive an R restart.

The `examples/recursive-agents` Shiny app provides a deterministic local HTTP
producer and separate descendant transcript selection, with no paid model calls.
Each session owns a separate graph and its host controls. The example's requester
token scopes disclosure; shared applications must supply authenticated identities
and independent action authorization.

## Copying selected context into a specialist

Fresh delegation gives each child its own context. When a specialist needs prior
conversation material, the host can explicitly select and authorize a snapshot:

```{r}
# The application authenticates the user and selects a revision from its store.
selected <- host_selected_context()
fork <- ContextFork(
  owner_id = selected$owner_id,
  conversation_id = selected$conversation_id,
  branch_id = selected$branch_id,
  revision = selected$revision,
  fork_point = selected$fork_point,
  view = "context",
  turns = selected$turns,
  max_bytes = 1024 * 1024
)

# specialist is an independently configured Agent with an empty Chat.
handle <- fork_agent(
  root, specialist, fork,
  authorize = host_authorize_current_source,
  usage_limits = UsageLimits(max_requests = 4),
  max_runs = 2
)
root$continue_agent(handle, "Review the selected evidence.",
  UsageLimits(max_requests = 2))
```

The `host_*` functions above belong to the application. The authorization function
receives the selected snapshot, checks the current user's access and source
revision, and returns `owner_id`, `conversation_id`, `branch_id`, and `revision`.
Deputy requires an exact match and rechecks it before each continuation.

Choose `view = "transcript"` for selected retained history or `view = "context"`
for selected model input after compaction. The host supplies the corresponding
turns and any summary material it intends to include. Deputy never silently
replaces a bounded selection with the full transcript. An oversized selection
fails before dispatch.

ellmer's native turns preserve text, images, documents and settled tool evidence.
Copied tool bindings are removed; incomplete tool evidence becomes explicitly
labelled inert text. The specialist keeps its own configured prompt, tools and
permissions, subject to the current parent's restrictions. Parent and child
messages diverge after the snapshot, and continuing the child does not copy later
parent messages.

Selecting or creating a branch remains an operation on the host's conversation
store. `fork_agent()` copies context and retains an independent child; it does
not create a persistent branch. shinychat supplies its managed history UI and
rich-content rendering. A headless adapter requires its supported public branch
contract; native host-supplied forks need no private shinychat APIs.

## Defining Sub-Agents

Use `agent_definition()` to describe a sub-agent:

```{r, eval = FALSE}
library(deputy)

code_reviewer <- agent_definition(
  name = "code_reviewer",
  description = "Reviews R code for best practices and potential issues",
  prompt = "You are an expert R developer. Review code for correctness,
    style, and potential bugs. Be specific and actionable.",
  tools = tools_file()
)

data_analyst <- agent_definition(
  name = "data_analyst",
  description = "Analyses data files and produces statistical summaries",
  prompt = "You are a data analyst. Read data files and provide clear,
    concise statistical summaries with key insights.",
  tools = tools_data()
)
```

Fields:

| Field | Description |
|-------|-------------|
| `name` | Unique routing key: lowercase letter first, then letters, numbers, `_`, or `-` |
| `description` | What this agent does (shown to the lead LLM) |
| `prompt` | System prompt for the sub-agent |
| `tools` | Tools available to the sub-agent |
| `model` | `"inherit"` (default) or a specific model name |
| `skills` | Optional list of skills to load |
| `disallowed_tools` | Additional tool names that the child must not use |
| `memory` | Optional context appended to the child's system prompt |
| `mcp_servers` | Optional MCP servers loaded for the child |
| `initial_prompt` | Optional text prepended to delegated tasks |
| `max_requests` | Optional per-child request limit |
| `permission_mode` | Optional mode that preserves or narrows the lead policy |

### Delegated Permission Boundaries

A sub-agent cannot gain authority that its lead does not have. Deputy
intersects the requested child mode with the lead's capabilities, write root,
tool gates, and permission callback. `disallowed_tools` adds further denials.

The allowed child-mode transitions use the same authority order as
`Agent$set_permission_mode()`:

| Lead mode | Allowed child modes |
|-----------|---------------------|
| `"readonly"` | `"readonly"` |
| `"standard"` | `"standard"`, `"readonly"` |
| `"plan"` | `"plan"`, `"readonly"` |
| `"full"` | `"full"`, `"standard"`, `"plan"`, `"readonly"` |

For example, a readonly reviewer can be declared beneath a standard lead
without gaining file reads, web access, or tools that the lead explicitly
denied:

```{r, eval = FALSE}
reviewer <- agent_definition(
  name = "reviewer",
  description = "Reviews files without changing them",
  prompt = "Review the requested files and report findings.",
  tools = tools_file(),
  permission_mode = "readonly",
  disallowed_tools = "read_csv",
  max_requests = 5
)
```

## Creating a LeadAgent

`LeadAgent` extends `Agent` with a built-in `delegate_to_agent` tool:

```{r, eval = FALSE}
lead <- LeadAgent$new(
  chat = ellmer::chat_anthropic(),
  sub_agents = list(code_reviewer, data_analyst),
  system_prompt = "You coordinate analysis tasks. Delegate to the
    appropriate specialist and synthesise their findings."
)

lead$available_sub_agents()
#> [1] "code_reviewer" "data_analyst"
```

Names are trimmed and converted to lowercase by `agent_definition()`. A
`LeadAgent` rejects duplicate normalized names at construction and registration.
Its `sub_agent_defs` field returns a list of read-only S7 `AgentDefinition`
values; use `register_sub_agent()` so the registry and lead prompt stay
synchronized. `agent_definition()` and `AgentDefinition()` are the same S7
constructor. To revise a definition, edit a plain property record and construct
a new value:

```{r definition-value, eval = TRUE}
library(deputy)
reviewer <- agent_definition("Reviewer", "Reviews text", "Read carefully.")
fields <- S7::props(reviewer)
fields$name <- "limited-reviewer"
fields$max_requests <- 2L
limited_reviewer <- do.call(agent_definition, fields)
limited_reviewer$name
limited_reviewer$max_requests
# The original definition still has no request limit.
reviewer$max_requests
```

Read properties with `$`, `S7::prop()`, or `@`. Definitions compose original
ellmer tools and read-only Skill values. Executable tools nested in either
value retain their closures, services, and caller-owned state; frozen
configuration does not freeze tool state. `S7::props()` retains these objects, so use the YAML
helpers and explicit registries below when you need a portable definition.

## Running a Delegation Task

When the lead agent decides to delegate, it calls `delegate_to_agent`
internally. The sub-agent runs to completion and returns its result to the
lead:

```{r multi-agent-demo}
library(deputy)

code_reviewer <- agent_definition(
  name = "code_reviewer",
  description = "Reviews R code and suggests improvements",
  prompt = "You are an R code reviewer. Be concise.",
  tools = tools_file()
)

chat <- ellmer::chat_anthropic(model = "claude-sonnet-4-20250514")
lead <- LeadAgent$new(
  chat = chat,
  sub_agents = list(code_reviewer)
)

result <- lead$run_sync(
  "Ask the code reviewer to look at the DESCRIPTION file and
   summarise what this package does."
)
cat(result$response)
```

Direct delegation participates in the lead run's usage limits. Each
child inherits the lead's remaining `UsageLimits`, and its usage is aggregated
into `result$usage` and enforced by the lead after delegation. Concurrent direct
children reserve allocations before launch. Background children, transitive
child trees, and cross-run global limits remain outside this contract.

## Stateless fan-out

For independent perspectives on supplied text, the host can select responders
directly. Each definition gets a fresh conversation with its own prompt and
at most one model request. The lead makes no model request for orchestration.
Definitions with tools, skills, or MCP servers are rejected before dispatch.

```{r, eval = FALSE}
lead <- LeadAgent$new(
  ellmer::chat_openai(model = "gpt-5.6-luna"),
  sub_agents = list(
    agent_definition("benefits", "Find benefits", "Explain the strongest benefits."),
    agent_definition("risks", "Find risks", "Explain the strongest risks.")
  )
)
batch <- lead$parallel_delegate(
  c(benefits = "Review a four-day working week.",
    risks = "Review a four-day working week."),
  max_active = 2,
  usage_limits = UsageLimits(max_requests = 2),
  run_context = list(workflow = "decision-review")
)
batch$status
batch$results$benefits$response
batch$run$usage
lead$list_subagents()
```

Names are normalized and must select each registered definition at most once.
The result preserves input order and has five fields:

| Field | Meaning |
| --- | --- |
| `mode` | `"stateless"`; other modes are currently rejected |
| `results` | Named list of child `AgentResult` objects, or `NULL` |
| `errors` | Named list of captured conditions, or `NULL` |
| `status` | Named vector: `completed`, `failed`, `stopped`, or `not_started` |
| `run` | Aggregate `AgentResult` with usage, events, and batch stop reason |

A failed responder does not discard successful siblings. Check `status` before
synthesis; a stopped child may have partial text. Child records include unique
agent, session, run, and delegation identifiers linked to the batch run, and
fire the usual `SubagentStart` and `SubagentStop` hooks. `$last_run()` retains
the aggregate on normal return. The lead's conversation is unchanged; the
host explicitly supplies selected responses to any later synthesis run.

### Limits and cancellation

`max_active` bounds simultaneous responders. Work runs in waves: the next wave
starts after all current responders settle. Each wave reserves request slots
before dispatch and divides the remaining token and cost ceilings among its
responders. A definition can narrow its request allowance to zero. Failed
dispatch attempts count as requests; missing cost information stops queued
work when a cost ceiling is configured.

Request limits prevent excess dispatches. Token and cost ceilings use observed
usage, so a wave can exceed them by up to one response per active responder.
Unused allocations are released after settlement. All observed child usage is
included in the aggregate, including failed or interrupted runs. The default
`on_exceed = "stop"` returns partial outcomes. `on_exceed = "error"` raises the
usual typed limit condition after cleanup; completed child records remain
inspectable through `$list_subagents()` and `$get_subagent_results()`.

For Shiny or another asynchronous host, use `$parallel_delegate_async()` and
consume its promise. `$interrupt()` prevents queued work from starting and
asks active responders to stop at the next supported provider boundary. The
batch settles after its active responders drain. The lead rejects overlapping
runs until cleanup finishes. Tool-using background workers and persistent
agent scheduling are separate future work.

## Opposing perspectives and synthesis

For a complete opposing-perspective workflow, run the installed standalone
example after configuring your OpenAI credentials:

```{r, eval = FALSE}
source(system.file("examples", "standalone", "09-debate.R", package = "deputy"))
```

It prints a Markdown comparison, requires both heads to complete, and then
uses a separate one-request moderator. Its bundled prompt can also be reused
with any Agent:

```{r, eval = FALSE}
agent$load_skill(system.file("skills", "debate", package = "deputy"))
```

Loading this prompt does not start independent responders. The host supplies
arguments or orchestrates fan-out explicitly.

## Inspecting child conversations

`lead$list_subagents()` has one row per admitted delegation in admission order,
even while children are running. `delegation_id` remains stable as `status`
changes from `queued` to `running` and then a settled outcome:

| Status | Meaning |
| --- | --- |
| `completed` | The child returned an AgentResult with stop reason `complete` |
| `stopped` | A limit, interruption, or another non-complete stop |
| `failed` | Construction or execution failed |
| `not_started` | Admitted work never entered a child run |
| `suspended` | An approval suspension, where the child supports it |

`stop_reason` preserves the exact runtime reason. Completion does not verify the
assigned task or its evidence. `admitted_at` records acceptance, `started_at`
records lifecycle entry before the start hook, and `completed_at` records the
invocation settling, including suspension. Cancellation in a start hook can
therefore yield `not_started` with a start timestamp. Run IDs are unavailable
until the child starts a run. Callback failures follow the existing hook policy
and appear separately in `hook_error`; they do not turn a successful child
response into an execution failure.

`get_subagent_results(delegation_id = id)` returns a list of retained AgentResults,
with `NULL` entries where no result is available and an empty list when no
delegation matches. `get_subagent_messages(session_id = session)`
returns the child's available conversation turns, including partial history
after failure. Both accessors retain admission order. Live message snapshots
are not a token stream. Reading them does not copy them into the lead's context,
resume a child, or approve a pending action. The host must authorize and redact
any user-facing disclosure. These records belong to the live LeadAgent; they
are not persisted as a durable child job store.

For an asynchronous lead run, `lead$interrupt(reason = "cancelled_by_host")`
cooperatively interrupts both the lead and active children. It takes effect at
the next supported provider boundary. Closing an inspection panel should not
call this method unless the user explicitly requests cancellation.

## Explicit input and context inspection

A plain task string still starts a fresh child conversation. For structured work,
use `DelegationInput()` with constraints, evidence references, a deliverable and
stop conditions. Stop conditions are instructions; runtime limits and permissions
remain separate host configuration.

```{r explicit-delegation, eval=FALSE}
lead <- LeadAgent$new(
  chat = ellmer::chat_openai(),
  sub_agents = list(agent_definition("reviewer", "Review evidence", "Keep units explicit.")),
  delegation_scope = list(owner_id = "team-a", conversation_id = "review-17"),
  delegation_sources = list(list(
    source_id = "assay-17", revision = "r3",
    owner_id = "team-a", conversation_id = "review-17",
    text = "Reported concentration: 12 mg/L.",
    allowed_agents = "reviewer"
  ))
)
brief <- DelegationInput(
  "Assess the reported concentration",
  constraints = "Preserve the reported units",
  evidence = list(list(source_id = "assay-17", revision = "r3")),
  deliverable = "A conclusion citing the source revision",
  stop_conditions = "State uncertainty if method information is missing"
)
lead$parallel_delegate(list(reviewer = brief))
id <- lead$list_subagents()$delegation_id[[1]]
lead$get_subagent_contexts(id)                     # initial manifest
lead$get_subagent_contexts(id, view = "current")   # working context
lead$get_subagent_contexts(id, redact = TRUE)      # content-redacted metadata
lead$get_subagent_messages()                      # retained child chats
```

Model-driven `delegate_to_agent` calls support the same brief fields. Models can
select exact source revisions, but cannot supply source bodies, owner scope or
byte ceilings. Every child brief is encoded as one JSON object with separate
brief fields and a `resolved_evidence` array, even when that array is empty.
Text that imitates an evidence header remains inside its quoted brief field.
This preserves source-selection provenance in the representation; it does not
make source text trustworthy or guarantee how the model interprets it.
The host supplies authenticated, authorized text snapshots;
Deputy checks scope, definition eligibility and revision equality against that
snapshot. Refreshing external evidence requires a new snapshot and LeadAgent.
Missing or `NULL` `allowed_agents` permits all definitions; `character()` permits
none. Snapshots are text only, not live retrieval or multimodal storage.

The default admission ceiling is 64 KiB for both initial system/message text and
the complete serialized manifest. Inputs reject explicitly when missing, stale,
unauthorized, invalid or oversized; `list_subagents()$input_error` records why.
Known token estimates also obey `ContextPolicy$max_tokens`. Unknown estimates
remain explicit and rely on byte admission, without a token-bound guarantee.

The initial manifest records static definition memory separately from resolved
evidence and survives changes to the working context. It is a preparation receipt,
not the exact provider wire payload or an authority grant. A redacted view is
explicitly incomplete. Hosts control disclosure of both text and metadata.
Inspection does not add child history to the lead, resume a child, or approve a
tool. This provides the data for a child-chat inspector; nested live streaming
and durable child continuation remain separate work.

## Monitoring with SubagentStop Hooks

Use a `SubagentStop` hook to log or inspect sub-agent results:

```{r, eval = FALSE}
hook_monitor <- HookMatcher(
  event = "SubagentStop",
  callback = function(agent_name, task, result, context) {
    cli::cli_alert_info("Sub-agent {agent_name} finished")
    cli::cli_alert("Status: {context$status}")
    NULL
  }
)

lead$add_hook(hook_monitor)
```

## Portable definitions

AgentDefinitions can live as `.yaml` or `.yml` files in `.deputy/agents/`. A file
contains data, while the host chooses the actual tools and skills it can
reference. Reading files does not run R code, load a Skill, contact an MCP
server, or create an Agent.

Here is a complete version 1 definition:

```yaml
version: 1
name: reviewer
description: Reviews local text for gaps
prompt: |
  Read the supplied text using read_file.
  Report unsupported claims and missing evidence concisely.
tools: [read_file]
model: inherit
skills: []
disallowed_tools: [write_file, run_bash]
memory: []
mcp_servers: []
initial_prompt: Keep the review grounded in the supplied text.
max_requests: 2
permission_mode: readonly
```

Only `version`, `name`, `description`, and `prompt` are required. All other
fields use `agent_definition()` defaults when omitted. The format covers the
entire constructor surface:

| Field | Type and meaning |
| --- | --- |
| `version` | Required numeric format version, currently `1` |
| `name` | Required routing name; canonicalized to lowercase |
| `description` | Required nonempty description shown to the lead |
| `prompt` | Required nonempty prompt; `|` preserves multiline text |
| `tools` | Tool registry keys; defaults to `[]` |
| `model` | Model string, default `inherit` |
| `skills` | Skill registry keys; defaults to `[]` |
| `disallowed_tools` | Tool names denied to the subagent, or `null` |
| `memory` | Sequence of memory strings, or `null` |
| `mcp_servers` | MCP server names for the existing delegation lifecycle, or `null` |
| `initial_prompt` | Initial task prefix, or `null` |
| `max_requests` | Non-negative integer request limit, or `null` |
| `permission_mode` | `standard`, `readonly`, `plan`, `full`, or `null` |

The lead still controls the authority ceiling. A file cannot give a subagent
more permission than its lead, nor supply credentials, host settings, or
nested subagents. Tool and skill references are exact, case-sensitive keys
like `read_file` or `concise_review`; they do not name R expressions, package
exports, or filesystem paths. A host can map a key to an approved Skill path.
Loading that Skill happens when the subagent is instantiated.

Unknown fields, unsupported versions, missing references, duplicate keys,
and `!expr` tags are errors. Quote YAML strings like `"yes"` and `"123"` to
prevent YAML's automatic type conversion. Use `[]` for empty sequences and
`null` for absent optional fields. Single strings are also accepted for
one-element sequences. Files are limited to 1 MiB.

### Read and delegate

The package ships the YAML above as a working example. This chunk reads it
without model credentials:

```{r definition-file-read, eval = TRUE}
library(deputy)
tool_registry <- list(read_file = tool_read_file)
definition_dir <- system.file("examples", "agent-definitions", package = "deputy")
definitions <- agent_definitions(definition_dir, tools = tool_registry)
names(definitions)
definitions$reviewer$max_requests
```

For a project, use `agent_definitions(tools = tool_registry)` to discover
`.deputy/agents/`. Discovery is nonrecursive and ordered by filename. A missing
or empty directory returns an empty list. Duplicate canonical names and
invalid files stop discovery rather than silently dropping a definition.

```{r, eval = FALSE}
lead <- LeadAgent$new(
  chat = ellmer::chat_openai(model = "gpt-5.6-luna"),
  sub_agents = definitions,
  permissions = Permissions(mode = "standard", file_write = FALSE),
  usage_limits = UsageLimits(max_requests = 6)
)
result <- lead$run_sync("Delegate a review of input.txt to reviewer.")
lead$list_subagents()
```

### Write and round-trip

Pass the same registries when writing and reading. The writer finds registry
keys by exact object identity; it refuses unmapped tools/skills and ambiguous
aliases. Existing files require `overwrite = TRUE`. Comments, YAML formatting,
and names attached to R lists or character sequences are not preserved; object
order and registry identity are preserved.

The writer completes a temporary file in the destination directory before
installing it. With `overwrite = TRUE`, it renames the completed file into
place. Otherwise it creates a hard link, which fails atomically if another
writer has already created the destination. This requires filesystem support
for hard links. A failed write or installation leaves an existing definition
intact.

```{r definition-file-roundtrip, eval = TRUE}
path <- tempfile(fileext = ".yaml")
agent_definition_write(definitions$reviewer, path, tools = tool_registry)
restored <- agent_definition_read(path, tools = tool_registry)
identical(restored, definitions$reviewer)
unlink(path)
```

## Bind host policy and execution resources

A child starts a fresh conversation, but a tool closure can still capture shared
state. `DelegationPolicy()` makes that ownership visible in its initial manifest.
The default `resource_mode = "shared"` borrows definition and skill tools; Deputy
does not close them. Choose `"exclusive"` with a shared host `resource_key` to
reject overlapping delegations that use one resource in the same R process.
This does not lock direct host access or isolate the filesystem.

For an independent R worker per delegation, use the public resource factory:

```{r, eval=FALSE}
policy <- DelegationPolicy(
  resource_mode = "owned",
  resources = function(agent, definition, context) {
    worker <- RSession$new(agent)
    DelegationResources(worker$tools(), cleanup = worker$close)
  },
  observers = c("Notification", "Stop")
)
lead <- LeadAgent$new(
  chat = chat,
  permissions = permissions_full(), # Trusted R uses this account's access
  sub_agents = list(agent_definition("analyst", "Analyze data", "Check the data.")),
  delegation_policy = policy
)
```

The factory receives a fresh child before provider dispatch. It can bind an
`RSession` or `McpConnection` using public Agent identity and workspace APIs.
Return tools without running or mutating the child. A factory must clean up its
own partial construction if it throws; once it returns `DelegationResources`,
Deputy calls cleanup exactly once after settlement or subsequent setup failure.
Cancellation is cooperative. Cleanup errors appear separately in
`lead$list_subagents()$cleanup_error`.

Owned mode rejects tools captured in definitions or skills. MCP selections in a
definition require an owned factory, which resolves the host's explicit config
and server/tool allowances. The ordinary child permissions and definition denylist
apply to all registered tool origins. Stateless parallel responders never call a
resource factory or acquire tools, handlers or resource leases.

Tool governance hooks (`PreToolUse`, `PostToolUse`, `PostToolUseFailure`) forward
from the live lead registry. Other observers require explicit selection. Their
callbacks remain shared host code; Deputy does not clone captured state. Children
retain the initial permission ceiling and check current lead restrictions before
each tool operation. Provider-native tools are unsupported for delegated runs
because they bypass these execution gates.

For a hosted child that asks questions, select the interactive tool in its
definition (or return it from its owned factory) and bind a handler:

```{r, eval=FALSE}
policy <- DelegationPolicy(human_input = function(questions, context) {
  # The host authenticates the owner and supplies the UI routing function.
  request_answers(
    questions,
    owner = context$scope$owner_id,
    agent_id = context$agent_id,
    run_id = context$run_id,
    delegation_id = context$delegation_id
  )
})
lead <- LeadAgent$new(
  chat,
  sub_agents = list(agent_definition(
    "interviewer", "Clarify requirements", "Ask focused questions.",
    tools = tools_interactive()
  )),
  delegation_scope = list(owner_id = "authenticated-owner", conversation_id = "chat-1"),
  delegation_policy = policy
)
```

The current routing context includes child, parent and session identifiers. An
unbound `ask_user` fails preparation rather than falling back to a process-wide
handler. Scope values are host assertions, not authentication. The host still
controls access and redaction for manifests and retained transcripts.

Human answers and delegation briefs never approve tool effects. Delegated durable
approval is not yet supported: a lead configured with `approval_dir` rejects
delegation before dispatch, and a child permission callback returning
`PermissionResultPending` stops without executing the requested effect. Continue
to use standalone `Agent` approval APIs where durable suspension is needed.
Child continuation and durable job recovery are separate backlog work.

## Compact outcomes and child history

The `delegate_to_agent` tool returns a JSON `DelegationOutcome` projection. Its
`runtime` contains execution status, exact stop reason, and stable child, run,
parent tool-call and delegation identities. The `answer` and optional `claims`
are model-authored. A normal stop never verifies a claim or approves an artifact.
`missing_evidence` and `unresolved_work` are `NULL` when not supplied.
`parallel_delegate()` also returns named `outcomes` beside its existing results.

Answers are bounded to 8 KiB of UTF-8 text. Larger answers use the existing
ContextPolicy offload store; a scoped reference retains provenance. The model
projection includes at most eight references and reports omissions.
`runtime$claim_omissions` reports omitted entries and text truncation for each
structured claim, including in host inspection. Full child
history is separate from the lead's model context.

Bind disclosure to the authenticated host session before exposing inspection:

```{r, eval=FALSE}
owner_session <- new.env(parent = emptyenv()) # supplied by host authentication
scope <- list(owner_id = "owner-1", conversation_id = "chat-1")
disclosure <- DelegationDisclosure(
  authorize = function(requester, scope) identical(requester, owner_session),
  redact = function(view, requester) view # apply application-specific redaction
)
lead <- LeadAgent$new(
  chat, sub_agents = definitions,
  delegation_scope = scope, delegation_disclosure = disclosure
)
lead$parallel_delegate(c(analyst = "Analyze the evidence"))
views <- lead$inspect_subagents(owner_session, transcript = TRUE)
views[[1]]$outcome$runtime
views[[1]]$turns # native public ellmer turns for optional rendering

history <- lead$export_subagents(owner_session)
saveRDS(history, "child-history.rds") # host-owned private storage
restored <- delegation_history(
  readRDS("child-history.rds"), owner_session, disclosure,
  scope = history$scope # use the host's trusted ownership record on reload
)
```

Authorization runs before lookup, even for nonexistent IDs. The default policy
denies access. Redaction receives a child view, or a list with `kind = "artifact"`
and `result` for `read_subagent_result()`. Remove references as well as text when
an artifact must not be disclosed. Host callbacks are trusted application code;
requester strings and saved scope do not authenticate a user. The low-level
`get_subagent_*()` getters remain private-to-the-host inspection surfaces and do
not perform end-user authorization.

Snapshots include initial manifests, run and cumulative child usage, outcomes,
errors and optionally transcripts. Child cumulative usage currently equals its
single run; continuation is unsupported. Unknown usage/cost stays unknown.
Public ellmer record/replay strips executable tools, raw provider payloads,
hidden thinking and arbitrary display extras. General rich-widget persistence
is separate work. Text remains untrusted content for the renderer to sanitize.

Export accepts settled children only. Replay neither constructs an Agent nor
runs a model or tool, and never grants permission, cancels or resumes execution.
Restored artifact availability is explicitly unresolved until checked against
host storage. Live inspection checks retained artifacts and reports missing
ones. The default 16 MiB disclosure ceiling fails explicitly for larger views;
select one child or omit its transcript. Storage retention and durable active-run
recovery remain host responsibilities. Public content records carry a tool-value
kind receipt so ordinary application objects are never interpreted as content
constructors. Supported classed tool values (data frames, factors, dates, times,
and durations) use a portable JSON text projection; native typed attachments
retain their content representation. Unsupported application S7 values become
an explicit omission notice without invoking their methods or exposing private
fields. Other nonportable values still fail closed. The optional child-chat adapter builds on
these headless records; Deputy core does not require shinychat.

## Observe child activity

Observation reads a transient, bounded buffer. The runtime remains the only
consumer of each child stream. This supports concurrent subscribers without
running tools twice or letting a slow view block a child.

```{r, eval=FALSE}
reader <- lead$observe_subagents(owner_session)
initial <- reader$snapshot(transcript = FALSE)
# Start work through a separate host action; observation itself starts nothing.
promise <- lead$parallel_delegate_async(c(analyst = "Analyze the evidence"))
update <- reader$poll()
update$events
update$gaps # ranges lost to eviction; recover with reader$snapshot()
reader$close() # detach only; the child continues

reconnected <- lead$observe_subagents(owner_session, after = update$cursor)
reconnected$poll()
# A separately authorized host action may cancel one child:
lead$interrupt_subagent(delegation_id, reason = "user_cancelled")
```

Each cursor contains a transient stream ID and sequence number. Sequences increase
across the lead, while envelopes retain child conversation, delegation, run,
parent and tool-call IDs. A snapshot returns its boundary cursor; subsequent polls
return later events. Reconnecting with a previous cursor does not repeat events
already acknowledged. Foreign and future cursors fail. A filtered subscription
still reports gaps conservatively for the whole stream.

The defaults retain at most 256 envelopes and 1 MiB. Individual events above
64 KiB become an explicit `content_omitted` envelope, recoverable from retained
public history when available. These limits cover the observation buffer, not
the separately retained conversation. `DelegationObservation()` configures them.
There are no per-reader queues and no subscriber callbacks inside execution.

Every read reauthorizes the host request. Event redaction receives
`list(kind = "event", event = envelope)`; returning a list without `event` hides
it. Redaction errors leave the cursor unchanged and never fail the child. Host
redactors must remain synchronous and must not drive execution. Internal
observation errors appear separately as `observation_error`. Hidden thinking and
private transport objects are excluded. Tool start events are deduplicated
against the existing runtime tool-call identity; execution is unchanged.

`settled` is the delegation terminal envelope; a child's own `stop` event may
precede it. The former includes setup/unstarted outcomes and final governance
status. Closing, selecting and replaying a view do not cancel or resume a child.
`interrupt_subagent()` is a trusted host control API requiring the host's separate
action authorization. Streams are in-process only: process restart needs
host-owned saved history, and active-run recovery remains unsupported.

## Inspect children in Shiny

The optional adapter places an activity strip and selected child conversation
beside the host's lead chat. It requires shinychat 0.5.0, Shiny, bslib, commonmark
and xml2. The core runtime has no UI dependency. Assistant Markdown is rendered
then filtered to inert HTML so inline and fenced code retain their literal
characters. Native tool code values and errors remain plain text.

```{r, eval=FALSE}
ui <- bslib::page_fluid(
  bslib::layout_columns(lead_chat_ui, subagent_chat_ui("children"))
)
server <- function(input, output, session) {
  subagent_chat_server(
    "children", lead,
    requester = function() authenticated_owner_session,
    # Optional: authorize this action separately from disclosure.
    on_cancel = function(id, requester) {
      authorize_child_cancellation(requester, id)
      lead$interrupt_subagent(id, "user_cancelled")
    }
  )
}
```

Activity cards and the keyboard-accessible selector identify individual
delegations, including repeated calls to the same specialist. The selected view
shows execution status, initial context, usage, unresolved claims and native tool
cards with retained attachments. Completed execution does not establish task
success, approval or evidence verification. Tool results remain untrusted.

Selecting, closing and reopening a child only reads history. The panel has no
prompt handler; it cannot resume a child. Closing detaches its reader while
children continue. Cancellation appears only when the host supplies a separately
authorized callback. Polling rechecks authorization and redaction; denial clears
the displayed history. Buffer gaps recover from retained snapshots, while partial
live text is bounded. The read-only composer is hidden with scoped presentation
CSS tested against shinychat 0.5.0; no private runtime helpers are used.

For saved history, provide a reactive `history` plus the current host-owned
`disclosure` policy and trusted `scope`, obtained independently of the saved
payload. Replay restores no execution, and unresolved artifact references stay
unresolved. The module returns reactive selection, views, notice and closed
state for host composition.

Run `inst/examples/subagent-chats/app.R` for a deterministic local demo with
native tool output and a plot, concurrent specialists, repeated names, a partial
failure, optional cancellation, access revocation and saved-history replay. Its
visible model/tool counters make extra execution detectable. It uses a local
fixture service and needs no paid model credentials.


## A trusted scientific result

The runnable `inst/examples/trusted-mini-agent` example delegates proposal
writing to a specialist, then uses a separate host-owned Agent for durable
approval and a fixed descriptive computation. The reviewer sees the dataset
revision, groups, units and method, and can edit or deny the inputs. The
computed result panel reads only the designated tool's host receipt; deliberately
false model commentary cannot replace it. The example uses clearly synthetic
plant weights and a local deterministic provider, with no paid calls.

```r eval=FALSE
shiny::runApp(system.file("examples", "trusted-mini-agent", package = "deputy"))
```

The explicit handoff is necessary because delegated child approval continuation
is not yet supported. The receipt keeps the source child delegation separate
from the execution Agent. Reading a child remains observational. This example
demonstrates the execution and output contract, not scientific correctness or
model quality; trusted code, appropriate inputs and effective review are still
required.

### Trusted results across delegation

`LeadAgent$new(trusted_results = TrustedResults(...))` applies one
[trusted-results policy](permissions.html#trusted-results) to the whole
delegation tree. Every child inherits it, so each definition's tools must pass
the same no-bypass check: no code execution, no further delegation, and only
read-only, closed-world tools unless exempted. A designated tool can live in a
child, but it must be the same tool object wherever it appears. When a child
runs it, the result is recorded in the lead's run and sent to the lead's
`on_result`, carrying the child's `delegation_id`:

```r eval=FALSE
forecast <- get_forecast_tool()
lead <- LeadAgent$new(
  chat = ellmer::chat("openai/gpt-5.6-luna"),
  sub_agents = list(agent_definition(
    "forecaster",
    "Produce forecasts",
    "Call get_forecast.",
    tools = list(forecast)
  )),
  trusted_results = TrustedResults(
    forecast = "get_forecast",
    on_result = function(event) show_forecast(event$value)
  )
)
```

Definitions that break the rule are rejected by `register_sub_agent()` and at
construction. Children cannot suspend for durable approval, so input review
of a child's trusted tool still needs the separate executor pattern above.
