---
title: "Multi-Agent Orchestration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Agent Orchestration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

For complex tasks, deputy supports a **lead agent** that delegates work to
specialised **sub-agents**. Each sub-agent has its own tools, prompt, and
(optionally) model. The lead agent decides when to delegate and synthesises
the results.

## When to Use Multi-Agent

Multi-agent orchestration is useful when:

- Different parts of a task need different tools (e.g., code analysis vs.
  data analysis)
- You want to isolate sub-tasks with their own system prompts
- Sub-agents should run with restricted permissions
- You need audit trails per sub-agent via `SubagentStop` hooks

For simpler tasks, a single `Agent` with the right tools is usually enough.

## Defining Sub-Agents

Use `agent_definition()` to describe a sub-agent:

```{r, eval = FALSE}
library(deputy)

code_reviewer <- agent_definition(
  name = "code_reviewer",
  description = "Reviews R code for best practices and potential issues",
  prompt = "You are an expert R developer. Review code for correctness,
    style, and potential bugs. Be specific and actionable.",
  tools = tools_file()
)

data_analyst <- agent_definition(
  name = "data_analyst",
  description = "Analyses data files and produces statistical summaries",
  prompt = "You are a data analyst. Read data files and provide clear,
    concise statistical summaries with key insights.",
  tools = tools_data()
)
```

Fields:

| Field | Description |
|-------|-------------|
| `name` | Unique routing key: lowercase letter first, then letters, numbers, `_`, or `-` |
| `description` | What this agent does (shown to the lead LLM) |
| `prompt` | System prompt for the sub-agent |
| `tools` | Tools available to the sub-agent |
| `model` | `"inherit"` (default) or a specific model name |
| `skills` | Optional list of skills to load |
| `disallowed_tools` | Additional tool names that the child must not use |
| `memory` | Optional context appended to the child's system prompt |
| `mcp_servers` | Optional MCP servers loaded for the child |
| `initial_prompt` | Optional text prepended to delegated tasks |
| `max_requests` | Optional per-child request limit |
| `permission_mode` | Optional mode that preserves or narrows the lead policy |

### Delegated Permission Boundaries

A sub-agent cannot gain authority that its lead does not have. Deputy
intersects the requested child mode with the lead's capabilities, write root,
tool gates, and permission callback. `disallowed_tools` adds further denials.

The allowed child-mode transitions use the same authority order as
`Agent$set_permission_mode()`:

| Lead mode | Allowed child modes |
|-----------|---------------------|
| `"readonly"` | `"readonly"` |
| `"standard"` | `"standard"`, `"readonly"` |
| `"plan"` | `"plan"`, `"readonly"` |
| `"full"` | `"full"`, `"standard"`, `"plan"`, `"readonly"` |

For example, a readonly reviewer can be declared beneath a standard lead
without gaining file reads, web access, or tools that the lead explicitly
denied:

```{r, eval = FALSE}
reviewer <- agent_definition(
  name = "reviewer",
  description = "Reviews files without changing them",
  prompt = "Review the requested files and report findings.",
  tools = tools_file(),
  permission_mode = "readonly",
  disallowed_tools = "read_csv",
  max_requests = 5
)
```

## Creating a LeadAgent

`LeadAgent` extends `Agent` with a built-in `delegate_to_agent` tool:

```{r, eval = FALSE}
lead <- LeadAgent$new(
  chat = ellmer::chat_anthropic(),
  sub_agents = list(code_reviewer, data_analyst),
  system_prompt = "You coordinate analysis tasks. Delegate to the
    appropriate specialist and synthesise their findings."
)

lead$available_sub_agents()
#> [1] "code_reviewer" "data_analyst"
```

Names are trimmed and converted to lowercase by `agent_definition()`. A
`LeadAgent` rejects duplicate normalized names at construction and registration.
Its `sub_agent_defs` field returns a read-only snapshot; use
`register_sub_agent()` so the registry and lead prompt stay synchronized.

## Running a Delegation Task

When the lead agent decides to delegate, it calls `delegate_to_agent`
internally. The sub-agent runs to completion and returns its result to the
lead:

```{r multi-agent-demo}
library(deputy)

code_reviewer <- agent_definition(
  name = "code_reviewer",
  description = "Reviews R code and suggests improvements",
  prompt = "You are an R code reviewer. Be concise.",
  tools = tools_file()
)

chat <- ellmer::chat_anthropic(model = "claude-sonnet-4-20250514")
lead <- LeadAgent$new(
  chat = chat,
  sub_agents = list(code_reviewer)
)

result <- lead$run_sync(
  "Ask the code reviewer to look at the DESCRIPTION file and
   summarise what this package does."
)
cat(result$response)
```

Direct delegation participates in the lead run's usage limits. Each
child inherits the lead's remaining `UsageLimits`, and its usage is aggregated
into `result$usage` and enforced by the lead after delegation. Concurrent direct
children reserve allocations before launch. Background children, transitive
child trees, and cross-run global limits remain outside this contract.

## Stateless fan-out

For independent perspectives on supplied text, the host can select responders
directly. Each definition gets a fresh conversation with its own prompt and
at most one model request. The lead makes no model request for orchestration.
Definitions with tools, skills, or MCP servers are rejected before dispatch.

```{r, eval = FALSE}
lead <- LeadAgent$new(
  ellmer::chat_openai(),
  sub_agents = list(
    agent_definition("benefits", "Find benefits", "Explain the strongest benefits."),
    agent_definition("risks", "Find risks", "Explain the strongest risks.")
  )
)
batch <- lead$parallel_delegate(
  c(benefits = "Review a four-day working week.",
    risks = "Review a four-day working week."),
  max_active = 2,
  usage_limits = UsageLimits(max_requests = 2),
  run_context = list(workflow = "decision-review")
)
batch$status
batch$results$benefits$response
batch$run$usage
lead$list_subagents()
```

Names are normalized and must select each registered definition at most once.
The result preserves input order and has five fields:

| Field | Meaning |
| --- | --- |
| `mode` | `"stateless"`; other modes are currently rejected |
| `results` | Named list of child `AgentResult` objects, or `NULL` |
| `errors` | Named list of captured conditions, or `NULL` |
| `status` | Named vector: `completed`, `failed`, `stopped`, or `not_started` |
| `run` | Aggregate `AgentResult` with usage, events, and batch stop reason |

A failed responder does not discard successful siblings. Check `status` before
synthesis; a stopped child may have partial text. Child records include unique
agent, session, run, and delegation identifiers linked to the batch run, and
fire the usual `SubagentStart` and `SubagentStop` hooks. `$last_run()` retains
the aggregate on normal return. The lead's conversation is unchanged; the
host explicitly supplies selected responses to any later synthesis run.

### Limits and cancellation

`max_active` bounds simultaneous responders. Work runs in waves: the next wave
starts after all current responders settle. Each wave reserves request slots
before dispatch and divides the remaining token and cost ceilings among its
responders. A definition can narrow its request allowance to zero. Failed
dispatch attempts count as requests; missing cost information stops queued
work when a cost ceiling is configured.

Request limits prevent excess dispatches. Token and cost ceilings use observed
usage, so a wave can exceed them by up to one response per active responder.
Unused allocations are released after settlement. All observed child usage is
included in the aggregate, including failed or interrupted runs. The default
`on_exceed = "stop"` returns partial outcomes. `on_exceed = "error"` raises the
usual typed limit condition after cleanup; completed child records remain
inspectable through `$list_subagents()` and `$get_subagent_results()`.

For Shiny or another asynchronous host, use `$parallel_delegate_async()` and
consume its promise. `$interrupt()` prevents queued work from starting and
asks active responders to stop at the next supported provider boundary. The
batch settles after its active responders drain. The lead rejects overlapping
runs until cleanup finishes. Tool-using background workers and persistent
agent scheduling are separate future work.

## Opposing perspectives and synthesis

For a complete opposing-perspective workflow, run the installed standalone
example after configuring your OpenAI credentials:

```{r, eval = FALSE}
source(system.file("examples", "standalone", "09-debate.R", package = "deputy"))
```

It prints a Markdown comparison, requires both heads to complete, and then
uses a separate one-request moderator. Its bundled prompt can also be reused
with any Agent:

```{r, eval = FALSE}
agent$load_skill(system.file("skills", "debate", package = "deputy"))
```

Loading this prompt does not start independent responders. The host supplies
arguments or orchestrates fan-out explicitly.

## Monitoring with SubagentStop Hooks

Use a `SubagentStop` hook to log or inspect sub-agent results:

```{r, eval = FALSE}
hook_monitor <- HookMatcher$new(
  event = "SubagentStop",
  callback = function(agent_name, task, result, context) {
    cli::cli_alert_info("Sub-agent {agent_name} finished")
    cli::cli_alert("Status: {context$status}")
    NULL
  }
)

lead$add_hook(hook_monitor)
```

## Portable definitions

AgentDefinitions can live as `.yaml` or `.yml` files in `.deputy/agents/`. A file
contains data, while the host chooses the actual tools and skills it can
reference. Reading files does not run R code, load a Skill, contact an MCP
server, or create an Agent.

Here is a complete version 1 definition:

```yaml
version: 1
name: reviewer
description: Reviews local text for gaps
prompt: |
  Read the supplied text using read_file.
  Report unsupported claims and missing evidence concisely.
tools: [read_file]
model: inherit
skills: []
disallowed_tools: [write_file, run_bash]
memory: []
mcp_servers: []
initial_prompt: Keep the review grounded in the supplied text.
max_requests: 2
permission_mode: readonly
```

Only `version`, `name`, `description`, and `prompt` are required. All other
fields use `agent_definition()` defaults when omitted. The format covers the
entire constructor surface:

| Field | Type and meaning |
| --- | --- |
| `version` | Required numeric format version, currently `1` |
| `name` | Required routing name; canonicalized to lowercase |
| `description` | Required nonempty description shown to the lead |
| `prompt` | Required nonempty prompt; `|` preserves multiline text |
| `tools` | Tool registry keys; defaults to `[]` |
| `model` | Model string, default `inherit` |
| `skills` | Skill registry keys; defaults to `[]` |
| `disallowed_tools` | Tool names denied to the subagent, or `null` |
| `memory` | Sequence of memory strings, or `null` |
| `mcp_servers` | MCP server names for the existing delegation lifecycle, or `null` |
| `initial_prompt` | Initial task prefix, or `null` |
| `max_requests` | Non-negative integer request limit, or `null` |
| `permission_mode` | `standard`, `readonly`, `plan`, `full`, or `null` |

The lead still controls the authority ceiling. A file cannot give a subagent
more permission than its lead, nor supply credentials, host settings, or
nested subagents. Tool and skill references are exact, case-sensitive keys
like `read_file` or `concise_review`; they do not name R expressions, package
exports, or filesystem paths. A host can map a key to an approved Skill path.
Loading that Skill happens when the subagent is instantiated.

Unknown fields, unsupported versions, missing references, duplicate keys,
and `!expr` tags are errors. Quote YAML strings like `"yes"` and `"123"` to
prevent YAML's automatic type conversion. Use `[]` for empty sequences and
`null` for absent optional fields. Single strings are also accepted for
one-element sequences. Files are limited to 1 MiB.

### Read and delegate

The package ships the YAML above as a working example. This chunk reads it
without model credentials:

```{r definition-file-read, eval = TRUE}
library(deputy)
tool_registry <- list(read_file = tool_read_file)
definition_dir <- system.file("examples", "agent-definitions", package = "deputy")
definitions <- agent_definitions(definition_dir, tools = tool_registry)
names(definitions)
definitions$reviewer$max_requests
```

For a project, use `agent_definitions(tools = tool_registry)` to discover
`.deputy/agents/`. Discovery is nonrecursive and ordered by filename. A missing
or empty directory returns an empty list. Duplicate canonical names and
invalid files stop discovery rather than silently dropping a definition.

```{r, eval = FALSE}
lead <- LeadAgent$new(
  chat = ellmer::chat_openai(),
  sub_agents = definitions,
  permissions = Permissions$new(mode = "standard", file_write = FALSE),
  usage_limits = UsageLimits(max_requests = 6)
)
result <- lead$run_sync("Delegate a review of input.txt to reviewer.")
lead$list_subagents()
```

### Write and round-trip

Pass the same registries when writing and reading. The writer finds registry
keys by exact object identity; it refuses unmapped tools/skills and ambiguous
aliases. Existing files require `overwrite = TRUE`. Comments, YAML formatting,
and names attached to R lists or character sequences are not preserved; object
order and registry identity are preserved.

The writer completes a temporary file in the destination directory before
installing it. With `overwrite = TRUE`, it renames the completed file into
place. Otherwise it creates a hard link, which fails atomically if another
writer has already created the destination. This requires filesystem support
for hard links. A failed write or installation leaves an existing definition
intact.

```{r definition-file-roundtrip, eval = TRUE}
path <- tempfile(fileext = ".yaml")
agent_definition_write(definitions$reviewer, path, tools = tool_registry)
restored <- agent_definition_read(path, tools = tool_registry)
identical(restored, definitions$reviewer)
unlink(path)
```
