---
title: "Permissions and limits"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Permissions and limits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup, include = FALSE, eval = TRUE}
library(deputy)
```

Deputy checks every tool call against the agent's permission policy before the
tool runs, and counts every request and tool call against its usage limits.
Both are set when you create the agent. You can tighten them afterwards, but
you can't loosen them.

## Presets

Four functions cover the common cases:

| Preset | Reads files | Writes files | Runs code | Uses the web |
|---|---|---|---|---|
| `permissions_readonly()` | yes | no | no | no |
| `permissions_standard()` (the default) | yes | inside the working directory | no | no |
| `permissions_plan()` | yes | no | no | read-only web tools |
| `permissions_full()` | yes | yes | yes | yes |

Pass one to `Agent$new()`:

```{r}
agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  tools = tools_file(),
  permissions = permissions_readonly()
)
```

"Reads files" means any file the R process can read, wherever it is. The
working directory limits writes, not reads.

`permissions_plan()` is meant for planning before acting: it allows tools
annotated as read-only, including web fetches and searches, plus `ask_user` so
the model can check its plan with you. `permissions_full()` allows every tool
call.

## Build your own policy

`Permissions()` sets each capability directly:

```{r}
policy <- Permissions(
  file_read = TRUE,
  file_write = "/path/to/output",
  bash = FALSE,
  r_code = TRUE,
  web = FALSE,
  install_packages = FALSE
)
```

| Argument | Controls | Default |
|---|---|---|
| `file_read` | the built-in file reading tools | `TRUE` |
| `file_write` | `write_file`, `edit_file` and `multi_edit`: `TRUE`, `FALSE`, or a directory to confine them to | the current directory |
| `bash` | `run_bash` | `FALSE` |
| `r_code` | `run_r_code` | `FALSE` |
| `web` | web tools and any tool annotated as open world | `FALSE` |
| `install_packages` | package installation | `FALSE` |
| `tool_allowlist` | if set, only these tools may run | `NULL` |
| `tool_denylist` | these tools never run | `NULL` |
| `can_use_tool` | a function that decides each call (see below) | `NULL` |
| `permission_prompt_tool_name` | a tool the model may always call to ask for approval | `NULL` |
| `mode` | `"standard"`, `"readonly"`, `"plan"` or `"full"` | `"standard"` |

Deputy's own tools map to these capabilities by name. Other tools are judged
by their [annotations](tools.html#turn-an-r-function-into-a-tool): a tool
annotated as open world needs `web = TRUE`, and a destructive one needs file
writes or shell commands to be allowed. Missing annotations count as the risky
answer.

Setting `r_code = TRUE` or `bash = TRUE` lets the model run code with your
user account's access to files and the network. Only do that when you trust
the model, the prompt and every input it will see. [Running R
code](code-execution.html) shows how to run model-written code in an OS
sandbox instead.

To see how a policy treats a call without running anything, use
`permissions_check()`:

```{r, eval = TRUE}
policy <- Permissions(file_write = FALSE)
permissions_check(policy, "write_file", list(path = "notes.md"))
```

## Modes

The `mode` argument changes how the capabilities are applied:

| Mode | Allows |
|---|---|
| `"standard"` | any call the capabilities allow |
| `"readonly"` | Deputy's read tools, plus web tools if `web = TRUE` and custom tools you list in `tool_allowlist`; never writes, code or destructive tools |
| `"plan"` | tools annotated as read-only (open-world ones only if `web = TRUE`) |
| `"full"` | every call |

In every mode, `tool_denylist` and `tool_allowlist` are checked first, and the
tool named by `permission_prompt_tool_name` is always allowed.

## Narrow an agent's permissions

`set_permission_mode()` switches a running agent to a stricter mode. It can't
switch to a mode that allows more than the agent started with:

| Current mode | Can switch to |
|---|---|
| `"readonly"` | `"readonly"` |
| `"standard"` | `"standard"`, `"readonly"` |
| `"plan"` | `"plan"`, `"readonly"` |
| `"full"` | `"full"`, `"standard"`, `"plan"`, `"readonly"` |

```{r}
agent$set_permission_mode("readonly")
```

The new mode is combined with the original policy, so directory limits, allow
and deny lists and callbacks keep applying. If the new mode takes away web
access, Deputy also removes any provider-side web tools, because it can't
intercept their calls. To give an agent more permissions, create a new agent.

## Decide each call with a callback

A `can_use_tool` function receives the tool name, its arguments and a context
list, and returns `PermissionResultAllow()`, `PermissionResultDeny(reason)` or
`PermissionResultPending(reason)`. Any other return value, or an error, denies
the call.

In `"standard"` mode the callback has the final say: when it allows a call,
Deputy skips the capability checks. To add a rule on top of a policy rather
than replace it, deny what you want to block and hand everything else to
`permissions_check()` on a policy without the callback:

```{r}
base <- Permissions(file_write = getwd())

policy <- Permissions(
  file_write = getwd(),
  can_use_tool = function(tool_name, tool_input, context) {
    path <- if (is.character(tool_input$path)) tool_input$path else ""
    if (tool_name == "write_file" && grepl("^\\.env|secrets", path)) {
      return(PermissionResultDeny("Don't write to secret files."))
    }
    permissions_check(base, tool_name, tool_input, context)
  }
)
```

In `"readonly"` mode the callback can only deny calls that the mode would
otherwise allow. `"plan"` and `"full"` modes don't call it. The context list
includes the working directory, the tool's annotations and argument types, and
the run's identifiers. A [PreToolUse hook](hooks.html) is often simpler when
you only want to block a few calls.

Tools that run on the provider's servers, such as Anthropic's web search,
can't be checked per call. An agent whose policy has a callback refuses to
register them.

`PermissionResultPending()` pauses the run until a person approves the call.
[Human input and approvals](approvals.html) explains how to set that up.

## Usage limits

`UsageLimits()` caps what one run can spend. Set defaults on the agent, and
override them for a single run:

```{r}
agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  tools = tools_file(),
  usage_limits = UsageLimits(max_requests = 10, max_tool_calls = 20)
)

result <- agent$run_sync(
  "Summarise every R file in R/.",
  usage_limits = UsageLimits(max_tool_calls = 40)
)
```

Unset fields in a run's limits fall back to the agent's. The agent defaults to
25 model requests per run. Limits count only the current run, so a restored
conversation starts with a fresh budget.

| Limit | Stops the run with |
|---|---|
| `max_requests` | `"request_limit"` |
| `max_tool_calls` | `"tool_call_limit"` |
| `max_input_tokens` | `"input_token_limit"` |
| `max_output_tokens` | `"output_token_limit"` |
| `max_total_tokens` | `"total_token_limit"` |
| `max_cost_usd` | `"cost_limit"` |

Request and tool-call limits are checked before each request or call. Token
and cost limits can only be checked once a response reports its usage, so a
run can overshoot them by one response.

Cost comes from ellmer's price table. If you set `max_cost_usd` and ellmer
can't price the model, for example because the model is newer than your
version of ellmer, the run stops with `"cost_unavailable"` rather than guess.
Likewise, `result$cost$total` is `NA` whenever any response couldn't be
priced, and `result$cost$missing` counts those responses.

By default a limit returns an `AgentResult` with the reason in `stop_reason`.
With `UsageLimits(on_exceed = "error")`, it signals an error instead, and
`agent$last_run()` still holds the partial result.

Deputy also stops a run with `"tool_loop"` when the model makes the same tool
call three times in a row and gets the same result each time.

Subagents get the lead agent's remaining budget, further limited by their own
definition. See [Subagents](multi-agent.html).

## Permissions are not a sandbox

A permission decides whether the model may call a tool. It can't limit what a
tool does once it runs. `run_r_code` gets a separate process and a timeout, but
that process can read and write anything your user account can. For
model-written code that should only touch part of the system, use an OS sandbox
such as mcp-repl; see [Running R code](code-execution.html).

To make sure the results you show a user come from one trusted tool rather than
from model text, see [Trusted mini-agents](trusted-mini-agents.html).
