---
title: "Permissions and Safety"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Permissions and Safety}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Permissions control what an agent is allowed to do. They sit between the LLM
and tool execution, checking every tool call before it runs. deputy ships
with sensible presets and makes it easy to build custom policies.

## Permission Presets

```{r, eval = FALSE}
library(deputy)

# Read-only: known Deputy read tools are allowed
permissions_readonly()

# Plan: annotated read-only tools plus ask_user
permissions_plan()

# Standard (default): accessible reads and workspace-scoped writes;
# no R code, bash, or web
permissions_standard()

# Full: everything allowed (use with caution!)
permissions_full()
```

Each preset returns a `Permissions` object. Pass it to `Agent$new()`:

```{r, eval = FALSE}
agent <- Agent$new(
  chat = ellmer::chat_anthropic(),
  tools = tools_all(),
  permissions = permissions_readonly()
)
```

## Custom Permissions

For fine-grained control, create a `Permissions` object directly:

```{r, eval = FALSE}
perms <- Permissions$new(
  file_read = TRUE,
  file_write = "/path/to/allowed/dir",
  bash = FALSE,
  r_code = TRUE,
  web = FALSE,
  install_packages = FALSE
)
```

Setting `r_code = TRUE` or `bash = TRUE` authorizes Deputy's built-in
trusted-code tools. It does not confine the subprocess. Use this only when the
model, prompt, inputs, and current-user filesystem authority are all within the
same trust boundary. Direct `Permissions$new()` policies default `r_code` to
`FALSE`, including partial policies such as `Permissions$new(web = TRUE)`.

Fields:

| Field | Type | Description |
|-------|------|-------------|
| `file_read` | logical | Allow reading files |
| `file_write` | logical / path | Allow writing (optionally restricted to a directory) |
| `bash` | logical | Allow bash commands |
| `r_code` | logical | Allow R code execution |
| `web` | logical | Allow web access |
| `install_packages` | logical | Allow package installation |
| `can_use_tool` | function | Apply custom permission decisions |
| `tool_allowlist` | character | Deny tools not explicitly listed |
| `tool_denylist` | character | Always deny explicitly listed tools |
| `permission_prompt_tool_name` | character | Dedicated approval tool used in gating messages |

## Permission Modes

The `mode` field provides broad policy shortcuts:

| Mode | Behaviour |
|------|-----------|
| `"standard"` | Check each tool against the configured capabilities |
| `"readonly"` | Allow known Deputy read tools and explicit allowlist entries within configured capabilities |
| `"plan"` | Allow read-only annotated tools within configured capabilities and the human approval prompt tool |
| `"full"` | Allow every tool (dangerous!) |

```{r, eval = FALSE}
perms <- Permissions$new(mode = "readonly")
```

`permissions_plan()` creates a planning-oriented policy:

```{r, eval = FALSE}
agent <- Agent$new(
  chat = ellmer::chat_anthropic(),
  tools = tools_all(),
  permissions = permissions_plan()
)
```

## Changing an Agent's Mode

An agent's configured permissions are an authority ceiling. Calling
`set_permission_mode()` may preserve or narrow that authority, but it cannot
widen it or replace it with an incomparable policy:

| Current mode | Allowed target modes |
|--------------|----------------------|
| `"readonly"` | `"readonly"` |
| `"standard"` | `"standard"`, `"readonly"` |
| `"plan"` | `"plan"`, `"readonly"` |
| `"full"` | `"full"`, `"standard"`, `"plan"`, `"readonly"` |

Reapplying the current mode is an exact no-op. For an allowed narrowing,
Deputy intersects the target mode with the existing capabilities. Custom
restrictions, tool gates, callbacks, and directory-scoped write roots therefore
remain authoritative. If the target removes web access, Deputy also removes
registered provider-native web tools before activating the new policy. Ordinary
function tools stay registered because Deputy can still deny them when they are
requested; provider-side tools have no equivalent interception point.

```{r, eval = FALSE}
agent$set_permission_mode("readonly")
```

Create a newly configured `Agent` when broader or incomparable authority is
required.

## Tool Annotations

Tools carry annotations that describe their behaviour. The permission system
uses these annotations to make decisions:

```{r, eval = FALSE}
# A read-only tool
tool_safe <- ellmer::tool(
  fun = function(x) x,
  name = "safe_tool",
  description = "A safe, read-only tool",
  arguments = list(x = ellmer::type_string("Input")),
  annotations = ellmer::tool_annotations(
    read_only_hint = TRUE,
    destructive_hint = FALSE
  )
)
```

Annotations describe behavior; they do not independently grant authority.
In `"readonly"` mode, Deputy recognizes its built-in read tools and explicit
allowlist entries, while still denying known writes, destructive tools, and
open-world tools when web access is disabled. Unknown tools do not become
authorized merely by declaring `read_only_hint = TRUE`. In `"standard"` mode,
annotations participate in the configured capability checks.

## Custom Permission Callbacks

For complex logic, provide a `can_use_tool` callback:

```{r, eval = FALSE}
perms <- Permissions$new(
  can_use_tool = function(tool_name, tool_input, context) {
    # Block writes to sensitive files
    if (tool_name == "write_file") {
      if (grepl("^\\.env|secrets|credentials", tool_input$path)) {
        return(PermissionResultDeny(
          reason = "Cannot write to sensitive files"
        ))
      }
    }
    PermissionResultAllow()
  }
)
```

The callback receives:

- `tool_name` -- Name of the tool being called
- `tool_input` -- Named list of arguments
- `context` -- List with `working_dir` and `tool_annotations`

Custom callbacks govern function tools whose requests Deputy can intercept.
Provider-native tools are rejected when a callback is configured; use the
universal function tool when authorization depends on request arguments or run
context.

It must return `PermissionResultAllow()` or `PermissionResultDeny(reason)`.

## Example: Read-Only Agent

A read-only agent can explore files but cannot change anything:

```{r permissions-readonly-demo}
library(deputy)

chat <- ellmer::chat_anthropic(model = "claude-sonnet-4-20250514")
agent <- Agent$new(
  chat = chat,
  tools = tools_file(),
  permissions = permissions_readonly()
)

result <- agent$run_sync("What files are in the current directory?")
cat(result$response)
```

## Run Limits

Permissions decide whether a tool may run. Configure resource limits separately
with `UsageLimits()` on the agent or an individual run:

```{r, eval = FALSE}
agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-5.6-luna"),
  tools = tools_file(),
  permissions = permissions_standard(),
  usage_limits = UsageLimits(
    max_requests = 10,
    max_tool_calls = 20,
    max_cost_usd = 1
  )
)
```

When a limit is reached, the agent stops and the `AgentResult$stop_reason`
identifies the limit, such as `"request_limit"`, `"tool_call_limit"`, or
`"cost_limit"`. If `max_cost_usd` is configured and the provider omits a cost
record, Deputy fails closed with `"cost_unavailable"`. `Agent$cost()` and
`AgentResult$cost` expose `complete` and `missing`; an incomplete `total` is
`NA`, never an understated sum.

Deputy also stops a run with `"tool_loop"` after three consecutive requests for
the same tool and canonical arguments. Cosmetic changes in surrounding model
text do not reset that progress signal; changing the tool or its arguments
does.

## Trusted Code and OS Sandboxes

Permissions and sandboxes answer different questions. A permission decides
whether the model may call a tool. An OS sandbox constrains what already
authorized code can read, write, execute, or reach over the network.

`tool_run_r_code` and `tool_run_bash` are trusted-code tools. `callr` gives the
R tool a separate process and a timeout, but the subprocess retains the current
user's filesystem and network authority. Deputy therefore excludes R execution
from both the standard permission policy and the standard tool preset.

For model-generated R that needs confinement, use mcp-repl as the execution
backend. Its server entry must name the policy explicitly:

```json
{
  "mcpServers": {
    "r": {
      "command": "/absolute/path/to/mcp-repl",
      "args": [
        "--sandbox", "workspace-write",
        "--interpreter", "r"
      ]
    }
  }
}
```

```{r, eval = FALSE}
repl_tools <- tools_mcp_repl(
  server = "r",
  sandbox = "workspace-write"
)
```

`tools_mcp_repl()` checks the exact final `--sandbox` setting before starting
the server. It accepts only `"read-only"` and `"workspace-write"`. It rejects
implicit defaults, `danger-full-access`, `external-sandbox`, and
`inherit-codex`; the latter needs per-call metadata that mcptools does not
provide. mcp-repl owns the platform-specific enforcement and fails rather than
downgrading when it cannot establish the requested OS boundary. See mcp-repl's
[sandbox contract](https://github.com/posit-dev/mcp-repl/blob/main/docs/sandbox.md)
for its filesystem, network, and platform guarantees.
