---
title: "Permissions and Safety"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Permissions and Safety}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Permissions control which tool calls an agent can run. Use a preset for
common read or write tasks, or a `Permissions` object to allow specific
capabilities. Deputy checks the policy before running a tool.

## Permission Presets

```{r, eval = FALSE}
library(deputy)

# Read-only: known Deputy read tools are allowed
permissions_readonly()

# Plan: annotated read-only tools plus ask_user
permissions_plan()

# Standard (default): accessible reads and workspace-scoped writes;
# no R code, bash, or web
permissions_standard()

# Full: everything allowed (use with caution!)
permissions_full()
```

Each preset returns a `Permissions` object. Pass it to `Agent$new()`:

```{r, eval = FALSE}
agent <- Agent$new(
  chat = ellmer::chat_anthropic(),
  tools = tools_all(),
  permissions = permissions_readonly()
)
```

## S7 policy values

`Permissions(...)` creates a read-only S7 policy; presets return the same class.
Use `permissions_check(policy, tool_name, tool_input, context)` to inspect a
permission decision. Read fields with `S7::prop(policy, "mode")` or `policy$mode`.
Constructor flags must each be `TRUE` or `FALSE`, and a custom callback must be
a function. An empty tool allowlist denies all tools, while `NULL` disables that
gate.

```{r}
policy <- Permissions(file_write = FALSE)
permissions_check(policy, "write_file", list(path = "output.txt"))
S7::prop(policy, "file_write")
```

Create a new Agent to explicitly adopt broader authority. An existing Agent's
`set_permission_mode()` can only preserve or narrow its current policy; its
policy cannot be replaced. Serializing a policy retains configuration for the
caller to inspect, including its canonical directory grant. It does not grant
a restored record authority over an existing Agent. Callback closures remain
caller-owned executable state.

## Custom Permissions

For fine-grained control, create a `Permissions` object directly:

```{r, eval = FALSE}
perms <- Permissions(
  file_read = TRUE,
  file_write = "/path/to/allowed/dir",
  bash = FALSE,
  r_code = TRUE,
  web = FALSE,
  install_packages = FALSE
)
```

Setting `r_code = TRUE` or `bash = TRUE` authorizes Deputy's built-in
trusted-code tools. It does not confine the subprocess. Use this only when the
model, prompt, inputs, and current-user filesystem authority are all within the
same trust boundary. Direct `Permissions()` policies default `r_code` to
`FALSE`, including partial policies such as `Permissions(web = TRUE)`.

Fields:

| Field | Type | Description |
|-------|------|-------------|
| `file_read` | logical | Allow reading files |
| `file_write` | logical / path | Allow writing (optionally restricted to a directory) |
| `bash` | logical | Allow bash commands |
| `r_code` | logical | Allow R code execution |
| `web` | logical | Allow web access |
| `install_packages` | logical | Allow package installation |
| `can_use_tool` | function | Apply custom permission decisions |
| `tool_allowlist` | character | Deny tools not explicitly listed |
| `tool_denylist` | character | Always deny explicitly listed tools |
| `permission_prompt_tool_name` | character | Dedicated approval tool used in gating messages |

## Permission Modes

The `mode` field provides broad policy shortcuts:

| Mode | Behaviour |
|------|-----------|
| `"standard"` | Check each tool against the configured capabilities |
| `"readonly"` | Allow known Deputy read tools and explicit allowlist entries within configured capabilities |
| `"plan"` | Allow read-only annotated tools within configured capabilities and the human approval prompt tool |
| `"full"` | Allow every tool (dangerous!) |

```{r, eval = FALSE}
perms <- Permissions(mode = "readonly")
```

`permissions_plan()` creates a planning-oriented policy:

```{r, eval = FALSE}
agent <- Agent$new(
  chat = ellmer::chat_anthropic(),
  tools = tools_all(),
  permissions = permissions_plan()
)
```

## Changing an Agent's Mode

An agent's configured permissions are an authority ceiling. Calling
`set_permission_mode()` may preserve or narrow that authority, but it cannot
widen it or replace it with an incomparable policy:

| Current mode | Allowed target modes |
|--------------|----------------------|
| `"readonly"` | `"readonly"` |
| `"standard"` | `"standard"`, `"readonly"` |
| `"plan"` | `"plan"`, `"readonly"` |
| `"full"` | `"full"`, `"standard"`, `"plan"`, `"readonly"` |

Reapplying the current mode is an exact no-op. For an allowed narrowing,
Deputy intersects the target mode with the existing capabilities. Custom
restrictions, tool gates, callbacks, and directory-scoped write roots therefore
remain authoritative. If the target removes web access, Deputy also removes
registered provider-native web tools before activating the new policy. Ordinary
function tools stay registered because Deputy can still deny them when they are
requested; provider-side tools have no equivalent interception point.

```{r, eval = FALSE}
agent$set_permission_mode("readonly")
```

Create a newly configured `Agent` when broader or incomparable authority is
required.

## Tool Annotations

Tools carry annotations that describe their behaviour. The permission system
uses these annotations to make decisions:

```{r, eval = FALSE}
# A read-only tool
tool_safe <- ellmer::tool(
  fun = function(x) x,
  name = "safe_tool",
  description = "A safe, read-only tool",
  arguments = list(x = ellmer::type_string("Input")),
  annotations = ellmer::tool_annotations(
    read_only_hint = TRUE,
    destructive_hint = FALSE
  )
)
```

Annotations describe behavior; they do not independently grant authority.
In `"readonly"` mode, Deputy recognizes its built-in read tools and explicit
allowlist entries, while still denying known writes, destructive tools, and
open-world tools when web access is disabled. Unknown tools do not become
authorized merely by declaring `read_only_hint = TRUE`. In `"standard"` mode,
annotations participate in the configured capability checks.

## Custom Permission Callbacks

For complex logic, provide a `can_use_tool` callback:

```{r, eval = FALSE}
perms <- Permissions(
  can_use_tool = function(tool_name, tool_input, context) {
    # Block writes to sensitive files
    if (tool_name == "write_file") {
      if (grepl("^\\.env|secrets|credentials", tool_input$path)) {
        return(PermissionResultDeny(
          reason = "Cannot write to sensitive files"
        ))
      }
    }
    PermissionResultAllow()
  }
)
```

The callback receives:

- `tool_name` -- Name of the tool being called
- `tool_input` -- Named list of arguments
- `context` -- List with `working_dir` and `tool_annotations`

Custom callbacks govern function tools whose requests Deputy can intercept.
Provider-native tools are rejected when a callback is configured; use the
universal function tool when authorization depends on request arguments or run
context.

It must return `PermissionResultAllow()` or `PermissionResultDeny(reason)`.
These are read-only S7 values. Untyped lists and legacy S3 class tags do not
satisfy this callback contract and produce the existing invalid-result denial.
`interrupt` must be one non-missing logical value, and `reason` must be one
non-missing string. Construct a new result to revise a decision.

```{r permission-result-values, eval = TRUE}
library(deputy)
decision <- PermissionResultDeny("Host veto", interrupt = TRUE)
S7::S7_inherits(decision, PermissionResult)
decision$decision
S7::props(decision)
```

## Example: Read-Only Agent

A read-only agent can explore files but cannot change anything:

```{r permissions-readonly-demo}
library(deputy)

chat <- ellmer::chat_anthropic(model = "claude-sonnet-4-20250514")
agent <- Agent$new(
  chat = chat,
  tools = tools_file(),
  permissions = permissions_readonly()
)

result <- agent$run_sync("What files are in the current directory?")
cat(result$response)
```

## Run Limits

Permissions decide whether a tool may run. Configure resource limits separately
with `UsageLimits()` on the agent or an individual run:

```{r, eval = FALSE}
agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-5.6-luna"),
  tools = tools_file(),
  permissions = permissions_standard(),
  usage_limits = UsageLimits(
    max_requests = 10,
    max_tool_calls = 20,
    max_cost_usd = 1
  )
)
```

`UsageLimits()` and `AgentUsage()` create read-only S7 values. Their constructors
and `$` reads keep the same spelling; list indexing and field assignment are
replaced by property access and construction of a new value. Use `S7::props()`
to obtain a plain reporting record:

```{r}
limits <- UsageLimits(max_requests = 3, max_cost_usd = 0.25)
usage <- AgentUsage(requests = 1, input_tokens = 100, output_tokens = 20,
                    cached_tokens = 80, cost_usd = NA_real_)
limits$max_requests
usage$total_tokens # Cached input is already included in input_tokens
S7::props(usage)   # A plain list; editing it does not change usage
```

`NULL` leaves a limit unset; zero is an explicit limit. Run overrides inherit
unset fields from the Agent's defaults, while delegated agents receive the
lead's remaining allowance intersected with the child definition's limits.
Constructing or serializing a value does not change an active Agent's defaults.

When a limit is reached, the agent stops and the `AgentResult$stop_reason`
identifies the limit, such as `"request_limit"`, `"tool_call_limit"`, or
`"cost_limit"`. If `max_cost_usd` is configured and the provider omits a cost
record, Deputy fails closed with `"cost_unavailable"`. `Agent$cost()` and
`AgentResult$cost` expose `complete` and `missing`; an incomplete `total` is
`NA`, never an understated sum.

Deputy also stops a run with `"tool_loop"` after three consecutive requests for
the same tool and canonical arguments. Cosmetic changes in surrounding model
text do not reset that progress signal; changing the tool or its arguments
does.

## Trusted Code and OS Sandboxes

Permissions and sandboxes answer different questions. A permission decides
whether the model may call a tool. An OS sandbox constrains what already
authorized code can read, write, execute, or reach over the network.

`tool_run_r_code` and `tool_run_bash` are trusted-code tools. `callr` gives the
R tool a separate process and a timeout, but the subprocess retains the current
user's filesystem and network authority. Deputy therefore excludes R execution
from both the standard permission policy and the standard tool preset.

For model-generated R that needs confinement, use mcp-repl as the execution
backend. Its server entry must name the policy explicitly:

```json
{
  "mcpServers": {
    "r": {
      "command": "/absolute/path/to/mcp-repl",
      "args": [
        "--sandbox", "workspace-write",
        "--interpreter", "r"
      ]
    }
  }
}
```

```{r, eval = FALSE}
repl_tools <- tools_mcp_repl(
  server = "r",
  sandbox = "workspace-write"
)
```

`tools_mcp_repl()` checks the exact final `--sandbox` setting before starting
the server. It accepts only `"read-only"` and `"workspace-write"`. It rejects
implicit defaults, `danger-full-access`, `external-sandbox`, and
`inherit-codex`; the latter needs per-call metadata that mcptools does not
provide. mcp-repl owns the platform-specific enforcement and fails rather than
downgrading when it cannot establish the requested OS boundary. See mcp-repl's
[sandbox contract](https://github.com/posit-dev/mcp-repl/blob/main/docs/sandbox.md)
for its filesystem, network, and platform guarantees.

## Durable approval after process exit

For a decision that may outlive the R process, configure a private existing
`approval_dir` and return `PermissionResultPending()` from the permission
callback. Deputy suspends before that tool executes. It journals tools completed
before the pause and retains their results when a fresh Agent resumes.

```{r}
approval_dir <- file.path(tempdir(), "approvals")
dir.create(approval_dir, showWarnings = FALSE)

# Raw argument tools validate their own inputs. Keep convert = FALSE from the
# original registration; changing it on an existing tool changes its semantics.
export_report <- ellmer::tool(
  function(name) {
    if (!is.character(name) || length(name) != 1L || is.na(name)) {
      cli::cli_abort("name must be one string")
    }
    # A host would perform its effect here and return an external receipt.
    paste("Prepared", name)
  },
  name = "export_report",
  description = "Prepare a named report",
  arguments = list(name = ellmer::type_string()),
  convert = FALSE,
  annotations = ellmer::tool_annotations(
    read_only_hint = FALSE,
    destructive_hint = FALSE,
    open_world_hint = FALSE
  )
)

make_agent <- function(chat) {
  Agent$new(
    chat = chat,
    tools = list(export_report),
    permissions = Permissions(can_use_tool = function(...) {
      PermissionResultPending("Host review required for report preparation")
    }),
    approval_dir = approval_dir,
    working_dir = approval_dir,
    session_id = "report_session",
    agent_id = "report_agent",
    run_context = list(owner_id = "owner_1", conversation_id = "conversation_1")
  )
}

agent <- make_agent(ellmer::chat_openai())
result <- agent$run_sync("Prepare the annual report")
if (identical(result$stop_reason, "approval_pending")) {
  pending <- agent$pending_approval()
  approval_path <- pending$source$path
  pending$request
  pending$usage
}

# This inspection needs no Chat and executes no tools, including after restart.
pending <- approval_read(approval_path)

# The host authorizes its caller, restores its own conversation association,
# and reattaches the same tool definition, scope, and permission callback.
resumed <- make_agent(ellmer::chat_openai())
result <- resumed$resume_approval(
  approval_path,
  "approve",
  tool_input = list(name = "reviewed annual report")
)
# Alternatively: resumed$resume_approval(approval_path, "deny")
```

`ApprovalContinuation` is a read-only inspection value. It exposes the pending
request and tool-call ID, source correlation, decision, usage, budget and static
permission ceilings, and completed-effect journal. The approval event supplies
an ID and path that the host can associate with its owner, conversation, selected
branch and revision. These associations belong in the host's store; Deputy does
not provide a second conversation database. Do not treat correlation IDs as
credentials. Only expose a private record after authenticating its owner.

Only a `pending` record can resume. The receiving Agent's session ID, Agent ID,
working directory, context and delegation correlation must match. It checks
saved and current static permission ceilings before the current callback and
hooks. A host edit cannot authorize a forbidden path or tool. Rebinding checks
source body/formals, schema, conversion and metadata; the host remains
responsible for captured closure state and external resource identity.

When a batch contains three tool calls and the second needs approval, the first
result is retained and the third is marked unexecuted. Resume supplies one result
for every original ID. A later model request cannot replay an already executed
or denied operation in that continuation. Use a separate, explicitly authorized
workflow if repeating the same operation is intended.

A tool-boundary budget limit can also create a pending decision. Resume normally
keeps the suspended limits and all prior usage. To grant more allowance, pass
`usage_limits = UsageLimits(...)` explicitly; Deputy caps it by both the original
and current Agent ceilings. Inspect `pending$budget_ceiling` before presenting
choices. Model boundaries with no pending tool retain ordinary limit stops.

```{r}
resumed$resume_approval(
  approval_path,
  "approve",
  usage_limits = UsageLimits(max_requests = 10)
)
```

Consumption holds an OS lock and commits state before each effect. If a worker
exits while a tool is executing, the record remains `executing`; no automatic
retry is allowed. An interrupted `resuming` or `continuing` record, or a terminal
`indeterminate` record, also requires the host to inspect external receipts and
reconcile the outcome. Duplicate approval delivery is rejected. This protocol
does not claim exactly-once effects or power-loss durability.

The latest stored revision retains the complete effect journal. Superseded
revisions are removed after a successful commit, so repeated lifecycle updates
do not accumulate copies of the session. Each serialized revision is limited to
25 MiB within a 50 MiB store budget, leaving room for an atomic replacement.
Snapshots that cannot fit are rejected before publishing a pending approval;
later growth of the journal or session remains bounded by the same limit.

The stored transcript support uses ellmer's public content record/replay and the
same session payload as `save_session()` / `load_session()`. Loading a separate
session snapshot does not consume an approval. Tools, callbacks and provider
clients are never serialized. Offloaded data frames and serializable S3 checkpoint
metadata retain their values, classes and attributes. Runtime objects inside
metadata or attributes remain unsupported. The host owns access, retention and backups for
these trusted local R records. An optional shinychat adapter must use its
released public contract; it is not required for this headless workflow. See
[ADR-0016](https://github.com/JamesHWade/deputy/blob/main/dev/adr/0016-durable-approval-continuations.md)
for the control-state contract.
