---
title: "Retained specialists and agent graphs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retained specialists and agent graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

A [subagent](multi-agent.html) starts every task with an empty conversation.
Sometimes you want the opposite: a specialist that remembers what it did last
time, so a follow-up question can build on its earlier work. Deputy calls
these *retained* specialists. Any agent can own them, and you can let the
model delegate to them, follow up with them yourself, or wire several into a
graph where specialists delegate to each other.

## Retain an agent

`retain_agent()` hands a fully configured agent to an owner agent. The owner
then runs it, keeping its conversation between tasks:

```{r}
library(deputy)

owner <- Agent$new(ellmer::chat("openai/gpt-6-luna"))

analyst <- Agent$new(
  chat = ellmer::chat("anthropic/claude-sonnet-5"),
  tools = tools_data(),
  permissions = permissions_readonly(),
  system_prompt = "You analyse data files and explain your reasoning."
)

handle <- owner$retain_agent(
  analyst,
  usage_limits = UsageLimits(max_requests = 20)
)

first <- owner$continue_agent(
  handle,
  "Summarise sales.csv.",
  UsageLimits(max_requests = 4)
)
second <- owner$continue_agent(
  handle,
  "Which region drove the change you found?",
  UsageLimits(max_requests = 4)
)
```

Each call to `continue_agent()` sends a new task to the same conversation, with
its own allocation. The `usage_limits` given to `retain_agent()` cap the total
across all calls, and `max_runs` (32 by default) caps how many there can be.
`continue_agent_async()` returns a promise instead.

While retained, the specialist belongs to its owner. You can't run it directly
or change its prompt, model or tools until you release it, and it runs one task
at a time: a second call while one is running fails straight away.
`owner$cancel_agent(handle)` stops the current task and keeps the history so
far; `owner$release_agent(handle)` ends the arrangement. The specialist must be
a plain `Agent`, without an approval directory or fallback chats.

The specialist's tools and any connections they hold remain yours to manage.
Handles and retained conversations live in memory and don't survive an R
restart. To keep a record, export it with `owner$export_subagents()` (see
[Briefing and inspecting subagents](delegation.html)) before releasing.

## Let the model delegate to a specialist

`delegation_tool()` turns a handle into a tool for the owner's model. The model
supplies only the task; the specialist, its budget and everything else are
fixed by your code:

```{r}
owner$register_tool(delegation_tool(
  owner,
  handle,
  name = "ask_analyst",
  description = "Ask the data analyst a question about the sales files.",
  usage_limits = UsageLimits(max_requests = 4)
))

owner$run_sync("Find out what changed in sales this quarter and why.")
```

Every call continues the same conversation, whether the model or your code
makes it. The tool returns the same compact [outcome](delegation.html#what-a-delegation-returns)
as ordinary delegation.

### Start from an ellmer chat

If you already have configured ellmer chats, perhaps with their own tools and
history, `adopt_chat()` retains a copy of one without building an `Agent` by
hand:

```{r}
handle <- adopt_chat(
  analyst_chat,
  owner,
  permissions = Permissions(
    mode = "readonly",
    file_write = FALSE,
    tool_allowlist = names(analyst_chat$get_tools())
  ),
  usage_limits = UsageLimits(max_requests = 8),
  history = "retain",
  callbacks = "replace"
)
```

The copy keeps the chat's provider, model, system prompt and tools, and the
original chat is left alone. `history = "retain"` keeps its turns and
`"fresh"` starts empty. `callbacks = "replace"` acknowledges that Deputy
replaces the chat's tool callbacks with its own. The tools still share whatever
their closures hold with the original chat. Each call is also limited by the
owner's current permissions.

The `curated-chats` example app retains two specialists this way and lets you
follow up with one of them. It runs against a local test server, so it needs no
API key:

```{r}
shiny::runApp(system.file("examples", "curated-chats", package = "deputy"))
```

## A graph of specialists

`retain_agent_graph()` retains several agents at once and declares which may
delegate to which. Each route becomes a tool on the calling agent, with a fixed
target and allocation:

```{r}
handles <- root$retain_agent_graph(
  agents = list(analyst = analyst, reviewer = reviewer),
  routes = list(
    root = list(
      analyze = list(
        target = "analyst",
        description = "Analyse the evidence.",
        usage_limits = UsageLimits(max_requests = 4)
      )
    ),
    analyst = list(
      review = list(
        target = "reviewer",
        description = "Check the analysis.",
        usage_limits = UsageLimits(max_requests = 2)
      )
    )
  ),
  usage_limits = UsageLimits(max_requests = 12, max_tool_calls = 8),
  max_depth = 2,
  max_delegations = 8,
  max_concurrency = 2
)

result <- root$run_sync("Analyse the evidence and have the analysis reviewed.")
root$delegation_graph_usage()
```

Here the root may ask the analyst, and the analyst may ask the reviewer. The
graph's `usage_limits` cover every run in the graph until you release it,
including your own follow-ups with `continue_agent(handles$analyst, ...)`.
`max_depth` counts delegation steps from the root, `max_delegations` caps the
total number of delegations, and `max_concurrency` caps how many run at once.
An agent waiting for its own delegate counts toward concurrency, so this
three-level chain needs two slots. Routes can form a cycle, but a specialist
that is already busy can't be called again.

Token and cost limits are checked when responses arrive, so work already in
flight can overshoot them; request limits are checked before each request.

Permissions and tool hooks apply all the way down: a call made by the reviewer
must also pass the analyst's and the root's permissions and `PreToolUse`
hooks. Each specialist keeps its own provider, prompt and tools.

The root sees the whole graph through the same methods as a lead agent:
`list_subagents()`, `inspect_subagents()`, `observe_subagents()` and
`interrupt_subagent()`, described in [Briefing and inspecting
subagents](delegation.html). Interrupting one specialist also stops the
delegations it started. When the graph is idle, `root$release_agent_graph()`
removes the routes and handles.

The `recursive-agents` example app runs a three-level graph against a local
test server:

```{r}
shiny::runApp(system.file("examples", "recursive-agents", package = "deputy"))
```

## Start a specialist from part of a conversation

Sometimes a specialist needs to see part of an existing conversation, for
example the messages about one dataset. `ContextFork()` describes the turns to
copy and where they came from, and `fork_agent()` retains a new specialist
that starts with them:

```{r}
fork <- ContextFork(
  owner_id = "user-1",
  conversation_id = "chat-7",
  branch_id = "main",
  revision = "rev-42",
  fork_point = 12,
  view = "context",
  turns = selected_turns,
  max_bytes = 1024 * 1024
)

handle <- fork_agent(
  owner,
  specialist,
  fork,
  authorize = check_user_can_read,
  usage_limits = UsageLimits(max_requests = 4),
  max_runs = 2
)

owner$continue_agent(
  handle,
  "Review the selected evidence.",
  UsageLimits(max_requests = 2)
)
```

Your application chooses the turns, from the complete conversation
(`view = "transcript"`) or from what the model currently sees
(`view = "context"`), and describes where they came from. `specialist` must be
a new `Agent` with an empty conversation, and a selection larger than
`max_bytes` is an error. `authorize` is your function: it
receives the fork's description, checks that the current user may still read
that conversation, and returns its `owner_id`, `conversation_id`, `branch_id`
and `revision`. Deputy calls it again before every continuation and stops if
the answer changes.

The copied turns are history only. They keep text, images, documents and
completed tool calls, but not tool bindings or anything provider-specific, and
unfinished tool calls become plain text. The specialist uses its own prompt,
tools and permissions, and after the fork the two conversations are
independent. Deputy doesn't create or store branches; that's up to your
application.
