---
title: "Fallbacks, tracing and evaluation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fallbacks, tracing and evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(deputy)
```

Deputy leaves provider requests to ellmer: connections, retries, timeouts and
model parameters. It adds the things an agent needs on top of that: a second
provider to try when the first is down, a record of every request, and
OpenTelemetry spans that tie model calls to the run that made them.

## Fall back to another provider

`fallback_chats` lists chats to try, in order, when a request fails before the
model has sent anything back:

```{r fallback}
agent <- Agent$new(
  ellmer::chat("openai/gpt-6-luna"),
  fallback_chats = list(ellmer::chat("anthropic/claude-sonnet-5")),
  permissions = permissions_readonly(),
  usage_limits = UsageLimits(max_requests = 5)
)
result <- agent$run_sync("Summarise the supplied evidence.")
```

Deputy switches for connection errors and HTTP 408, 429, 500, 502, 503 and 504,
after ellmer's own retries. Authentication errors, errors in your callbacks and
validation failures stop the run instead. So does a failure after the model has
started streaming text or asked for a tool: switching then could repeat the
tool's effects, so Deputy keeps the partial result and reports the error.

Each fallback chat is copied, and the copy gets the agent's conversation,
system prompt and tools; its model, credentials and settings stay as you
configured them. Pass chats with no history or tools. Once Deputy switches, it
stays on the new chat for later runs, moving further down the list if that one
fails too. It never picks a provider or model you didn't list, and it doesn't
adapt content that a fallback provider can't accept.

Automatic compaction has its own fallback list,
`ContextPolicy(summary_fallback_chats = )`; see [Conversations and
context](conversations.html). A lead agent's fallbacks apply to the lead only;
subagents use the provider the lead is currently on.

## Retries and timeouts

ellmer retries failed HTTP requests itself, up to `getOption("ellmer_max_tries")`
attempts in total (3 by default), and gives up on a request after
`getOption("ellmer_timeout_s")` seconds (300 by default). Set these options once
when your application starts. Deputy doesn't add a retry loop of its own, so
three HTTP attempts count as one request toward `max_requests`, and a failed
request counts too.

The asynchronous structured-output path, `chat_structured_async()`, goes
through httr2's `req_perform_promise()`, which doesn't retry.

Each model request adds `"request_start"` and `"request_end"` events to the
run, or `"request_error"` with the original condition when it fails, and a
switch adds a `"fallback"` event. Find them in `result$events`. They may
contain provider error details, so treat them as private.

## Tracing with OpenTelemetry

ellmer already records OpenTelemetry spans for model calls, HTTP requests and
tool execution. Deputy adds a `deputy.run` span around each run, with
attributes such as `deputy.run.id` and `deputy.agent.id`, and events for
permission decisions, hooks, fallbacks, compaction, checkpoints, delegation
and the stop. ellmer's `gen_ai.conversation.id` is
set to the agent's session ID, and spans from subagents stay connected to the
run that delegated to them.

Configure an OpenTelemetry exporter before loading ellmer and Deputy. To try
it without an exporter, record spans in memory with otelsdk:

```{r tracing}
# In a fresh R session:
record <- otelsdk::with_otel_record({
  library(deputy)
  agent <- Agent$new(ellmer::chat("openai/gpt-6-luna"))
  agent$run_sync("Summarise this text in one sentence: ...")
})
record$value$run_id
record$traces
```

Deputy's spans carry identifiers and decisions only. They never include
prompts, tool arguments or results, validation feedback, error messages, file
paths or your `run_context`, even when ellmer's content capture is on. That
capture is controlled by ellmer: set
`OTEL_INSTRUMENTATION_GENAI_CAPTURE_MESSAGE_CONTENT=true` before loading ellmer
to include messages in ellmer's spans. ellmer's spans can also include endpoint
URLs and tool descriptions, so apply your exporter's privacy settings.

## Evaluate an agent

An evaluation runs a fixed set of cases, scores each answer, and records what
each run cost. `AgentResult` has what you need to record: `stop_reason`,
`usage`, `duration`, the run ID and the events. Put a case ID in `run_context`
to connect a result, and its trace, back to its case:

```{r}
result <- agent$run_sync(
  case$prompt,
  run_context = list(evaluation = list(case_id = case$id))
)
```

The bundled `10-evaluation.R` script is a small complete example: it runs a
two-case dataset, scores the answers with an exact-match rule, and returns a
data frame with the case, run and session IDs, status, requests, tokens, cost,
duration and any error class:

```{r evaluation}
source(system.file("examples", "standalone", "10-evaluation.R", package = "deputy"))
evaluation
```

Deputy doesn't have its own dataset format or scoring framework; replace the
cases and the scoring rule with your own.
