---
title: "Runtime integration and evaluation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Runtime integration and evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(deputy)
```

Deputy owns permissions, budgets, hooks, run identity, checkpoints, context
policy, and delegation. ellmer owns provider requests, schema transport, tool
execution, model parameters, and provider tracing. The integration requires
released ellmer 0.5.0 or later.

## Explicit fallback Chats

The caller chooses every destination by supplying configured `fallback_chats`.
Deputy clones these templates, preserves their credentials, base URL, model
parameters, and request callbacks, and installs the Agent's conversation,
system prompt, governed tools, and observers. Templates must have no history
or tools. Fallback neither discovers a provider nor silently selects a model.

```{r fallback}
agent <- Agent$new(
  ellmer::chat("openai/gpt-5.6-luna"),
  fallback_chats = list(ellmer::chat("openai/gpt-5.6-terra")),
  permissions = permissions_readonly(),
  usage_limits = UsageLimits(max_requests = 5)
)
result <- agent$run_sync("Summarize the supplied evidence.")
```

Fallback is eligible for a connection failure or HTTP 408, 429, 500, 502, 503,
or 504 before any response content or requested tool. Upstream serializers
check content/tool compatibility with the selected provider; Deputy does not
drop content or rewrite provider-specific tools to make a fallback succeed. Authentication errors,
application callbacks, and validation failures are terminal. A partially
streamed response or completed tool effect locks the run to its current Chat;
Deputy preserves the partial result and signals the failure. It never restarts
a completed tool loop. The selected Chat remains active; subsequent runs can
advance to later configured templates, without revisiting earlier ones.
LeadAgent's fallback policy applies to the lead. Child definitions inherit the
selected provider and do not implicitly gain additional destinations.

`request_start`, `request_end`, `request_error`, and `fallback` events retain
model selection and original conditions with run IDs. These local events may
contain provider error details. Failed dispatches count toward `max_requests`.
Reported token totals include only observed usage; unknown cost is `NA`, and a
cost cap prevents further requests when that cost cannot be enforced.

ellmer configures httr2 with `ellmer_max_tries` (default 3 total attempts) and
`ellmer_timeout_s` (default 300 seconds). Configure these once at application
startup if needed. Deputy does not change process-wide options during runs or
add another retry loop. HTTP retries belong to the upstream transport: our
real streaming fixture verifies three HTTP attempts are still one governed
model dispatch. The structured async path uses httr2 1.3.0's
[`req_perform_promise()`](https://httr2.r-lib.org/reference/req_perform_promise.html),
which explicitly does not perform retries. Our fixture confirms one HTTP
attempt for this path; callers must not assume identical retry behavior. Deputy does not estimate hidden
HTTP attempts from model-request callbacks or promise immediate cancellation
during an upstream backoff or in-flight structured request.

## Tracing governance

Configure an `otel` provider before loading ellmer and Deputy. Existing ellmer
spans describe invocation, model calls, HTTP work, and actual tool execution.
Deputy adds a `deputy.run` parent span and governance events, using
`deputy.run.id`, `deputy.agent.id`, parent/delegation IDs, and ellmer's
`gen_ai.conversation.id` mapped to the Deputy session ID. Async delegated runs
retain their trace parent and run correlation.

```{r tracing}
# In a fresh R session, this records locally without configuring an exporter.
record <- otelsdk::with_otel_record({
  library(deputy)
  agent <- Agent$new(ellmer::chat("openai/gpt-5.6-luna"))
  agent$run_sync("Summarize this supplied text.")
})
record$value$run_id
record$traces
```

A policy denial records a permission event and does not pretend a tool ran.
Deputy's trace adapter exports a small allowlist of identifiers and decisions;
it omits prompts, tool inputs/results, validation feedback, conditions, paths,
and product run context. Custom stop text is represented as `custom`. This
allowlist stays in place even when upstream content capture is enabled.

ellmer's message capture is separately opt-in via
`OTEL_INSTRUMENTATION_GENAI_CAPTURE_MESSAGE_CONTENT=true`, set before loading
ellmer. Deputy does not override it or create a second exporter. The caller's
upstream tracer may record endpoint URLs and tool descriptions; configure those
without secrets and apply the exporter's own privacy policy as appropriate.

## An external evaluation runner

The runnable `10-evaluation.R` example executes a fixed two-row dataset, applies
an application-owned exact-answer rule, and reports case/run/session IDs,
completion status, requests, reported tokens, cost, duration, and error class.
`run_context$evaluation$case_id` links local evidence to the case. The same run
ID appears on the trace, without enabling message capture.

```{r evaluation}
source(system.file("examples", "standalone", "10-evaluation.R", package = "deputy"))
evaluation
```

Replace that dataset and scoring rule in your external runner. Deputy provides
`AgentResult`, usage, stop reasons, events, and correlated traces; it does not
own a dataset format, scoring framework, exporter, or experiment database.
