---
title: "Structured output"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structured output}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(deputy)
```

When the next step of your analysis needs a list or a data frame rather than
prose, describe the shape you want with an ellmer type. Deputy asks the model
for that shape and gives you the converted R value.

## Extract from what's already there

`chat_structured()` works like ellmer's: it asks the model to answer in the
given type, without calling tools, and returns the R value.

```{r direct}
agent <- Agent$new(ellmer::chat("openai/gpt-6-luna"))

review_type <- ellmer::type_object(
  status = ellmer::type_enum(c("ok", "needs_review")),
  findings = ellmer::type_array(ellmer::type_string())
)

review <- agent$chat_structured(
  "The review found no problems.",
  type = review_type
)
review$status
```

`chat_structured_async()` returns a promise for the same value. The request
counts against the agent's usage limits like any other, and
`agent$last_run()$usage` shows what it cost.

If you already have a JSON Schema, `ellmer::type_from_schema()` converts it:

```{r schema}
status_type <- ellmer::type_from_schema(
  '{"type": "object",
    "properties": {"status": {"type": "string"}},
    "required": ["status"],
    "additionalProperties": false}'
)
```

## Do the work, then extract

Pass `type` to `run_sync()`, `run_async()` or `run()` and the agent works in
two steps: it completes the task with its tools as usual, then extracts the
typed value from the conversation. Both steps share one budget, and the
extraction step doesn't call tools, so no tool runs twice.

```{r task}
agent <- Agent$new(
  ellmer::chat("openai/gpt-6-luna"),
  tools = tools_file(),
  permissions = permissions_readonly(),
  usage_limits = UsageLimits(max_requests = 8)
)

result <- agent$run_sync("Review the README.", type = status_type)

if (!result_is_success(result)) {
  cli::cli_abort("The review stopped early: {result$stop_reason}.")
}
result$structured_output
```

`result$response` holds the model's text from the first step, and
`result$structured_output` holds the extracted value. Check
`result_is_success()` first: a run that hit a limit has no structured value.

## Check the value yourself

ellmer makes sure the value has the right shape. Rules about its content are
yours to check. Pass a `validate` function that returns `TRUE` when the value
is acceptable, and `FALSE` or a string explaining the problem when it isn't.
With `max_corrections` above zero, Deputy sends the explanation back to the
model and asks again:

```{r validation}
data <- agent$chat_structured(
  "Extract the review status.",
  type = status_type,
  validate = function(x) {
    if (x$status %in% c("ok", "needs_review")) {
      TRUE
    } else {
      "status must be 'ok' or 'needs_review'"
    }
  },
  max_corrections = 1
)
```

`max_corrections` defaults to zero, so a failed check is an error unless you
opt in. Each correction is another model request, without tools, and counts
against the run's limits. A reply that isn't valid JSON gets the same
treatment. When the corrections run out, Deputy signals a
`deputy_structured_output_invalid` error.

Some failures are never retried: the validator itself erroring or returning
`NA`, a conversion error ellmer can't explain, a provider error, or a response
cut off by the token limit. Every attempt is recorded as a
`"structured_attempt"` event on the result, including the rejected values, so
be careful where you log those events if the data is sensitive.

## Stream structured output

`stream(type = )` and `stream_async(type = )` use ellmer's structured
streaming: chunks arrive as raw JSON text while the model writes. Streaming
doesn't support `validate` or corrections. Some providers can only produce
structured output through a tool call; use `chat_structured()` with those.
