---
title: "Structured Output"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structured Output}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(deputy)

docs_init <- getFromNamespace("deputy_docs_init", "deputy")
docs <- docs_init("structured-output", "verified-vcr")
apply_output_format_instructions <- getFromNamespace(
  "apply_output_format_instructions",
  "deputy"
)
parse_structured_output <- getFromNamespace(
  "parse_structured_output",
  "deputy"
)

extract_openai_output_text <- function(stream_body) {
  lines <- strsplit(stream_body, "\n", fixed = TRUE)[[1]]
  data_lines <- lines[startsWith(lines, "data: ")]
  payloads <- sub("^data: ", "", data_lines)
  payloads <- payloads[nzchar(payloads) & payloads != "[DONE]"]

  deltas <- vapply(
    payloads,
    function(payload) {
      event <- jsonlite::fromJSON(payload, simplifyVector = FALSE)
      if (!identical(event$type, "response.output_text.delta")) {
        return("")
      }
      if (is.null(event$delta)) "" else event$delta
    },
    character(1)
  )

  paste(deltas, collapse = "")
}

run_structured_openai_demo <- function(task, output_format) {
  prompt <- apply_output_format_instructions(task, output_format)
  body <- docs$use_vcr("main-example", {
    httr2::request("https://api.openai.com/v1/responses") |>
      httr2::req_headers(
        Authorization = paste(
          "Bearer",
          Sys.getenv("OPENAI_API_KEY", unset = "docs-placeholder")
        )
      ) |>
      httr2::req_body_json(
        list(
          input = list(list(
            role = "user",
            content = list(list(type = "input_text", text = prompt))
          )),
          model = "gpt-4o-mini",
          stream = TRUE,
          store = FALSE,
          service_tier = "auto"
        ),
        auto_unbox = TRUE
      ) |>
      httr2::req_timeout(120) |>
      httr2::req_perform() |>
      httr2::resp_body_string()
  })

  text <- extract_openai_output_text(body)
  list(
    text = text,
    structured_output = parse_structured_output(text, output_format)
  )
}
```

```{r, echo = FALSE, results = 'asis'}
docs$status_block()
```

Sometimes you need the agent to return data in a specific format rather than
free-form text. deputy supports **structured output** via the `output_format`
parameter, which tells the LLM to return valid JSON matching a schema.

## JSON Object Format

The simplest form asks the LLM to return a JSON object:

```{r structured-json-object, eval = docs$should_eval, warning = FALSE, message = FALSE}
result <- run_structured_openai_demo(
  "List three popular R packages for data manipulation. Return a JSON object with a packages array where each element has name and description fields.",
  output_format = list(type = "json_object")
)

str(result$structured_output$parsed)
```

This recorded request uses the same `output_format` prompt shaping and JSON
parsing helpers that deputy applies inside `Agent$run_sync()`, but captures the
provider exchange directly so pkgdown can replay it deterministically.

## JSON Schema Format

For stricter control, provide a JSON Schema. The LLM will be constrained to
produce output matching the schema:

```{r structured-json-schema, eval = FALSE}
schema <- list(
  type = "object",
  properties = list(
    packages = list(
      type = "array",
      items = list(
        type = "object",
        properties = list(
          name = list(type = "string"),
          description = list(type = "string"),
          category = list(
            type = "string",
            enum = c("data", "visualization", "modeling", "infrastructure")
          )
        ),
        required = c("name", "description", "category")
      )
    )
  ),
  required = c("packages")
)

chat <- ellmer::chat_openai(model = "gpt-4o-mini")
agent <- Agent$new(chat = chat)

result <- agent$run_sync(
  "List three popular R packages and categorise them.",
  output_format = list(type = "json_schema", schema = schema)
)

parsed <- result$structured_output$parsed
str(parsed)
```

## Accessing Structured Output

The `structured_output` field on `AgentResult` contains:

| Field | Description |
|-------|-------------|
| `$parsed` | Parsed R list from the JSON response |
| `$raw` | Raw JSON string |
| `$valid` | `TRUE`, `FALSE`, or `NA` (if validation was skipped) |
| `$errors` | Validation error messages (if any) |
| `$format` | The `output_format` spec that was used |

```{r, eval = FALSE}
result$structured_output$valid
#> [1] TRUE

result$structured_output$parsed$packages[[1]]$name
#> [1] "dplyr"
```

## Schema Validation

When you provide a JSON Schema and the
[jsonvalidate](https://docs.ropensci.org/jsonvalidate/) package is installed,
deputy automatically validates the output:

```{r, eval = FALSE}
# If validation fails:
result$structured_output$valid
#> [1] FALSE

result$structured_output$errors
#> [1] "data/packages/0: must have required property 'category'"
```

Without jsonvalidate installed, `$valid` will be `NA` and `$errors` will be
empty. The parsed output is still available.

## Extraction Example

Structured output is especially useful for data extraction tasks:

```{r structured-extraction, eval = FALSE}
schema <- list(
  type = "object",
  properties = list(
    language = list(type = "string"),
    purpose = list(type = "string"),
    first_release_year = list(type = "integer")
  ),
  required = c("language", "purpose", "first_release_year")
)

chat <- ellmer::chat_openai(model = "gpt-4o-mini")
agent <- Agent$new(chat = chat)

result <- agent$run_sync(
  "Extract structured information about the R programming language.",
  output_format = list(type = "json_schema", schema = schema)
)

result$structured_output$parsed
```
