---
title: "Tools and Tool Bundles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tools and Tool Bundles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Tools are how agents interact with the world. deputy provides built-in tools
for common tasks and makes it easy to create your own.

## Built-in Tools

deputy includes the following tools out of the box:

| Tool | Description | Annotations |
|------|-------------|-------------|
| `tool_read_file` | Read a file's contents | read-only |
| `tool_read_markdown` | Convert a supported document to markdown | read-only |
| `tool_write_file` | Write or append to a file | destructive |
| `tool_edit_file` | Apply one targeted file edit | destructive |
| `tool_multi_edit` | Apply multiple file edits | destructive |
| `tool_list_files` | List files in a directory | read-only |
| `tool_glob_files` | Find files by glob pattern | read-only |
| `tool_grep_files` | Search file contents | read-only |
| `tool_run_r_code` | Execute trusted R code as the current user | destructive, open-world |
| `tool_run_bash` | Execute trusted shell code as the current user | destructive, open-world |
| `tool_read_csv` | Read and summarise a CSV | read-only |
| `tool_web_fetch` | Fetch a web page | read-only, open-world |
| `tool_web_search` | Search the web | read-only, open-world |
| `tool_ask_user` | Ask structured questions | read-only |

Each tool carries **annotations** that the permission system uses to decide
whether the agent is allowed to use it. See `vignette("permissions")` for
details.

## Tool Bundles

Use `ellmer::tool()` for local functions, package exports, and service call
wrappers, then register those objects through the same Agent API. Registration
does not load packages or infer a function's effects. Describe them explicitly:

```{r registration-contract, eval=TRUE}
library(deputy)
uppercase <- ellmer::tool(
  base::toupper,
  name = "uppercase",
  description = "Uppercase local text.",
  arguments = list(x = ellmer::type_string()),
  annotations = ellmer::tool_annotations(
    read_only_hint = TRUE, destructive_hint = FALSE,
    open_world_hint = FALSE, idempotent_hint = TRUE
  )
)
# Constructing this Chat does not make a model request.
chat <- ellmer::chat_openai(credentials = function() "example", model = "gpt-4o-mini")
agent <- Agent$new(chat = chat, tools = list(uppercase))
agent$get_tools()$uppercase("hello")
agent$register_tool(uppercase, replace = TRUE)
```

Existing names are errors unless `replace = TRUE` is explicit. Repeated names
inside one batch always fail, including with replacement enabled. List names
do not rename a tool. `set_tools()` replaces the whole user registry. Both
methods and construction validate and adapt the complete batch before changing
the backend registry; a malformed later tool leaves earlier tools untouched.
Skills use `load_skill(..., allow_conflicts = TRUE)` for explicit replacement.

Missing annotations stay absent on the source object. For custom tools,
permissions assume possible modification, destruction, and external access,
and no idempotence. An explicit read-only annotation makes an omitted
destructive annotation irrelevant. A local read tool therefore needs at least
`read_only_hint = TRUE` and `open_world_hint = FALSE` under standard permissions.
Readonly mode additionally requires an explicit allowlist entry for custom
tools. Annotations do not grant authority or verify a service's trustworthiness.
See `PermissionMode` for callback and mode behavior.

Inspect the supplied fields and defaults without executing the tool:

```{r tool-metadata, eval=TRUE}
tool_metadata(agent$get_tools()$uppercase)
```

For service tools, `tools_mcp(config, servers = "evidence")` selects an exact
configured server name before connecting. The qualified mcptools 1.0.2 bridge
retains annotations from the server and records `source$type = "mcp"`, its
server name, and tool name. It reports missing metadata rather than inventing
safe effects. Other mcptools versions fail explicitly pending qualification.
The caller must trust the configured service; metadata does not establish trust.

Pass these objects to `Agent$new(tools = ...)`, `register_tools()`, or an
explicit tool registry used by `agent_definition_read()`. The same metadata
survives YAML selection, Agent wrapping, cloning, and delegation. Permission
callbacks receive it as `context$tool_metadata`, alongside the supplied
`context$tool_annotations`. A service called `read_file` still uses its service
annotations and capabilities; its name does not confer native file access.
Local file checkpointing excludes MCP calls, so rewinding files never treats a
remote service write as a change to the local workspace.

Reconnecting a server invalidates tools from its old connection. Reload and
replace explicitly with `agent$load_mcp(config, servers = "evidence", replace = TRUE)`.
This refresh removes obsolete tools from the selected servers, including when
discovery succeeds with an empty tool set. Unrelated tools remain registered.
A load failure records a failed attempt in `mcp_status()`. Tools that still
have working connections remain registered. If mcptools already closed an
old connection, its invalidated handles are removed even when discovery or
registration fails. Malformed metadata never silently becomes an unannotated
executable.

Bundles group related tools together:

```{r, eval = FALSE}
library(deputy)

tools_file()  # read_file, read_markdown, write_file, list_files
tools_code()  # run_r_code, run_bash
tools_data()  # read_csv, read_file, read_markdown
tools_web()   # universal web_fetch and web_search
tools_all()   # everything
```

Registration is not authorization: `tools_code()` makes the two trusted-code
tools available to the model, while the Agent's `Permissions` still decides
whether either call may run. `permissions_standard()` denies both.

Combine bundles by concatenating:

```{r, eval = FALSE}
tools <- c(tools_file(), tools_code())
agent <- Agent$new(chat = ellmer::chat_openai(), tools = tools)
```

## Tool Presets

Presets are named collections with opinionated defaults:

```{r, eval = FALSE}
tools_preset("minimal")
tools_preset("standard")
tools_preset("dev")
tools_preset("data")
tools_preset("full")
```

| Preset | Included capabilities |
|---|---|
| `minimal` | file and document reads |
| `standard` | reads plus native file writes; no code execution |
| `data` | data reads plus trusted R execution |
| `dev` | native files plus trusted R and shell execution |
| `full` | every built-in tool |

## Creating Custom Tools

Use `ellmer::tool()` to define custom tools with type-safe arguments and
annotations:

```{r, eval = FALSE}
tool_lookup_user <- ellmer::tool(
  fun = function(user_id) {
    # your implementation
    list(name = "Alice", email = "alice@example.com")
  },
  name = "lookup_user",
  description = "Look up a user by their ID",
  arguments = list(
    user_id = ellmer::type_string("The user ID to look up")
  ),
  annotations = ellmer::tool_annotations(
    read_only_hint = TRUE,
    destructive_hint = FALSE,
    open_world_hint = FALSE,
    idempotent_hint = TRUE
  )
)
```

The `annotations` are optional but recommended. They tell the permission
system about the tool's behaviour:

- `read_only_hint` -- Tool does not change state
- `destructive_hint` -- Tool may delete or overwrite data
- `open_world_hint` -- Tool accesses external systems (network, APIs)
- `idempotent_hint` -- Safe to call multiple times with same input

### Using a Custom Tool

```{r tools-custom-demo}
library(deputy)

tool_dice <- ellmer::tool(
  fun = function(n = 1, sides = 6) {
    rolls <- sample(sides, n, replace = TRUE)
    paste("Rolled:", paste(rolls, collapse = ", "))
  },
  name = "roll_dice",
  description = "Roll one or more dice",
  arguments = list(
    n = ellmer::type_integer("Number of dice to roll"),
    sides = ellmer::type_integer("Number of sides per die")
  )
)

chat <- ellmer::chat_openai(model = "gpt-4o-mini")
agent <- Agent$new(chat = chat, tools = list(tool_dice))
result <- agent$run_sync("Roll 2d20 for me")
cat(result$response)
```

## Web Tools

deputy provides web tools that are **provider-aware**. Pass the chat to
`tools_web()` to select provider-native web tools when available. Calling it
without a chat returns Deputy's universal web tools:

```{r, eval = FALSE}
chat <- ellmer::chat_anthropic()
agent <- Agent$new(
  chat = chat,
  tools = tools_web(chat),
  permissions = Permissions$new(
    web = TRUE,
    tool_allowlist = c("web_search", "web_fetch")
  )
)
```

Provider-native tools execute outside R, so Deputy authorizes known native web
search and fetch tools before registering them. The policy must grant web
access and explicitly allow the tool; static denylists are also enforced at
registration. A custom `can_use_tool` callback rejects provider-native tools
because Deputy cannot supply request arguments or run context after provider
execution begins. Other provider-native tool types are rejected. Use the
universal function tools when authorization must inspect each request. If
`set_permission_mode()` later narrows away web access, Deputy removes registered
provider-native web tools before the new policy becomes active.

For providers without native support, deputy falls back to `tool_web_fetch`
(fetches a URL and extracts text) and `tool_web_search` (searches via
DuckDuckGo).

## MCP Integration

deputy can load tools from
[Model Context Protocol](https://modelcontextprotocol.io/) servers via
the [mcptools](https://github.com/posit-dev/mcptools) package:

```{r, eval = FALSE}
# Load tools from selected configured servers
agent <- Agent$new(
  chat = ellmer::chat_anthropic(),
  tools = c(tools_file(), tools_mcp(servers = "github"))
)
```

`tools_mcp()` returns an empty list with a warning when `mcptools` is not
installed or no tools are available. MCP servers are configured in
`~/.config/mcptools/config.json`; see the mcptools documentation for setup.

### Sandboxed R with mcp-repl

Use `tools_mcp_repl()` when model-generated R needs an OS sandbox rather than
the current user's authority:

```{r, eval = FALSE}
repl_tools <- tools_mcp_repl(
  config = "~/.config/mcptools/config.json",
  server = "r",
  sandbox = "workspace-write"
)

agent <- Agent$new(
  chat = ellmer::chat("openai"),
  tools = repl_tools,
  permissions = Permissions$new(web = FALSE)
)
```

The helper isolates the named server from the rest of the MCP configuration
and checks its exact final `--sandbox` argument. Missing policies and modes that
do not establish a Deputy-verifiable boundary are errors. mcp-repl then owns
OS-specific confinement, including fail-closed startup on unsupported hosts.
See `vignette("permissions")` for the complete trust model.

## Human-in-the-Loop

The `tool_ask_user` tool lets an agent ask the user a question during
execution. This is useful for confirmations, clarifications, or collecting
input:

```{r, eval = FALSE}
agent_id <- "agent-analysis"
session_id <- "session-analysis"

agent <- Agent$new(
  chat = ellmer::chat_openai(),
  tools = c(
    tools_file(),
    tools_interactive(
      callback = function(questions, context) {
        # Route a modal to this Agent's host session and collect its answers.
        collect_answers(questions, route = context$session_id)
      },
      context = list(agent_id = agent_id, session_id = session_id)
    )
  ),
  agent_id = agent_id,
  session_id = session_id
)
```

`tools_interactive()` creates a fresh `tool_ask_user` instance. Its handler and
routing context belong to that tool, so concurrent Agents cannot overwrite one
another. The context may also be a zero-argument function when routing values
must be resolved for each request. Omit the callback in an interactive console
to use `readline()`.

`set_ask_user_callback()` remains a process-wide compatibility fallback for
single-Agent scripts. Do not use it to route Shiny or other concurrent hosts.
