---
title: "Tools and Tool Bundles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tools and Tool Bundles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Tools are how agents interact with the world. deputy provides built-in tools
for common tasks and makes it easy to create your own.

## Built-in Tools

deputy includes the following tools out of the box:

| Tool | Description | Annotations |
|------|-------------|-------------|
| `tool_read_file` | Read a file's contents | read-only |
| `tool_read_markdown` | Convert a supported document to markdown | read-only |
| `tool_write_file` | Write or append to a file | destructive |
| `tool_edit_file` | Apply one targeted file edit | destructive |
| `tool_multi_edit` | Apply multiple file edits | destructive |
| `tool_list_files` | List files in a directory | read-only |
| `tool_glob_files` | Find files by glob pattern | read-only |
| `tool_grep_files` | Search file contents | read-only |
| `tool_run_r_code` | Execute R code | open-world |
| `tool_run_bash` | Execute a bash command | destructive, open-world |
| `tool_read_csv` | Read and summarise a CSV | read-only |
| `tool_web_fetch` | Fetch a web page | read-only, open-world |
| `tool_web_search` | Search the web | read-only, open-world |
| `tool_ask_user` | Ask structured questions | read-only |

Each tool carries **annotations** that the permission system uses to decide
whether the agent is allowed to use it. See `vignette("permissions")` for
details.

## Tool Bundles

Bundles group related tools together:

```{r, eval = FALSE}
library(deputy)

tools_file()  # read_file, read_markdown, write_file, list_files
tools_code()  # run_r_code, run_bash
tools_data()  # read_csv, read_file, read_markdown
tools_web()   # universal web_fetch and web_search
tools_all()   # everything
```

Combine bundles by concatenating:

```{r, eval = FALSE}
tools <- c(tools_file(), tools_code())
agent <- Agent$new(chat = ellmer::chat_openai(), tools = tools)
```

## Tool Presets

Presets are named collections with opinionated defaults:

```{r, eval = FALSE}
tools_preset("minimal")
tools_preset("standard")
tools_preset("dev")
tools_preset("data")
tools_preset("full")
```

## Creating Custom Tools

Use `ellmer::tool()` to define custom tools with type-safe arguments and
annotations:

```{r, eval = FALSE}
tool_lookup_user <- ellmer::tool(
  fun = function(user_id) {
    # your implementation
    list(name = "Alice", email = "alice@example.com")
  },
  name = "lookup_user",
  description = "Look up a user by their ID",
  arguments = list(
    user_id = ellmer::type_string("The user ID to look up")
  ),
  annotations = ellmer::tool_annotations(
    read_only_hint = TRUE,
    destructive_hint = FALSE,
    open_world_hint = FALSE,
    idempotent_hint = TRUE
  )
)
```

The `annotations` are optional but recommended. They tell the permission
system about the tool's behaviour:

- `read_only_hint` -- Tool does not change state
- `destructive_hint` -- Tool may delete or overwrite data
- `open_world_hint` -- Tool accesses external systems (network, APIs)
- `idempotent_hint` -- Safe to call multiple times with same input

### Using a Custom Tool

```{r tools-custom-demo}
library(deputy)

tool_dice <- ellmer::tool(
  fun = function(n = 1, sides = 6) {
    rolls <- sample(sides, n, replace = TRUE)
    paste("Rolled:", paste(rolls, collapse = ", "))
  },
  name = "roll_dice",
  description = "Roll one or more dice",
  arguments = list(
    n = ellmer::type_integer("Number of dice to roll"),
    sides = ellmer::type_integer("Number of sides per die")
  )
)

chat <- ellmer::chat_openai(model = "gpt-4o-mini")
agent <- Agent$new(chat = chat, tools = list(tool_dice))
result <- agent$run_sync("Roll 2d20 for me")
cat(result$response)
```

## Web Tools

deputy provides web tools that are **provider-aware**. Pass the chat to
`tools_web()` to select provider-native web tools when available. Calling it
without a chat returns Deputy's universal web tools:

```{r, eval = FALSE}
chat <- ellmer::chat_anthropic()
agent <- Agent$new(
  chat = chat,
  tools = tools_web(chat)
)
```

For providers without native support, deputy falls back to `tool_web_fetch`
(fetches a URL and extracts text) and `tool_web_search` (searches via
DuckDuckGo).

## MCP Integration

deputy can load tools from
[Model Context Protocol](https://modelcontextprotocol.io/) servers via
the [mcptools](https://github.com/posit-dev/mcptools) package:

```{r, eval = FALSE}
# Load tools from selected configured servers
agent <- Agent$new(
  chat = ellmer::chat_anthropic(),
  tools = c(tools_file(), tools_mcp(servers = "github"))
)
```

`tools_mcp()` returns an empty list with a warning when `mcptools` is not
installed or no tools are available. MCP servers are configured in
`~/.config/mcptools/config.json`; see the mcptools documentation for setup.

## Human-in-the-Loop

The `tool_ask_user` tool lets an agent ask the user a question during
execution. This is useful for confirmations, clarifications, or collecting
input:

```{r, eval = FALSE}
agent <- Agent$new(
  chat = ellmer::chat_openai(),
  tools = c(tools_file(), tools_interactive())
)

# In interactive sessions, a readline prompt appears.
# For non-interactive use, set a custom callback:
set_ask_user_callback(function(questions) {
  answers <- list()
  for (question in questions) {
    answers[[question$question]] <- question$options[[1]]$label
  }
  answers
})
```

The `tools_interactive()` function returns a list containing `tool_ask_user`.
