---
title: "Tools and Tool Bundles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tools and Tool Bundles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Tools let a model call R functions: read a file, fetch a page, or run a
calculation. Choose built-in tools or register functions with `ellmer::tool()`.
Permissions determine which calls can run.

## Reusable Skill Configuration

A `Skill` bundles instructions, tools, and declared requirements in a read-only
S7 value. Construct it with `Skill(...)` or `skill_create(...)`, and inspect
requirements with `skill_check_requirements(skill)`. `Skill()` defaults to
version `"0.0.0"`; `skill_create()` retains its `"1.0.0"` default.

```{r skill-values, eval = TRUE}
library(deputy)
concise <- skill_create(
  "concise",
  prompt = "Answer in one short paragraph.",
  requires = list(packages = "base")
)
skill_check_requirements(concise)$ok
fields <- S7::props(concise)
fields$prompt <- "Answer in one sentence."
shorter <- do.call(Skill, fields)
shorter$prompt
concise$prompt
```

Read properties with `$`, `@`, or `S7::prop()`. Configuration is frozen,
including the declared requirements and initially NULL fields. Tools keep their
original closures, clients, and caller-owned state when a Skill is constructed,
revised, or loaded into an Agent.

`agent$load_skill(skill)` checks package and provider requirements, registers
tools under the Agent's existing permissions, and appends the prompt. Its
`agent$skills()` getter returns a list of read-only Skill values. Checking
requirements does not install packages or execute tools. `skill_load(path)`
remains the explicit file-loading entry point and can source tool files declared
by the skill. AgentDefinition YAML attaches approved skills through a host
registry; it does not source those files.

## Built-in Tools

deputy includes the following tools out of the box:

| Tool | Description | Annotations |
|------|-------------|-------------|
| `tool_read_file` | Read a file's contents | read-only |
| `tool_read_markdown` | Convert a supported document to markdown | read-only |
| `tool_write_file` | Write or append to a file | destructive |
| `tool_edit_file` | Apply one targeted file edit | destructive |
| `tool_multi_edit` | Apply multiple file edits | destructive |
| `tool_list_files` | List files in a directory | read-only |
| `tool_glob_files` | Find files by glob pattern | read-only |
| `tool_grep_files` | Search file contents | read-only |
| `tool_run_r_code` | Execute trusted R code as the current user | destructive, open-world |
| `tool_run_bash` | Execute trusted shell code as the current user | destructive, open-world |
| `tool_read_csv` | Read and summarise a CSV | read-only |
| `tool_web_fetch` | Fetch a web page | read-only, open-world |
| `tool_web_search` | Search the web | read-only, open-world |
| `tool_ask_user` | Ask structured questions | read-only |

Each tool carries **annotations** that the permission system uses to decide
whether the agent is allowed to use it. See `vignette("permissions")` for
details.

## Tool Bundles

Bundles group related tools together:

```{r, eval = FALSE}
library(deputy)

tools_file()  # read_file, read_markdown, write_file, list_files
tools_code()  # run_r_code, run_bash
tools_data()  # read_csv, read_file, read_markdown
tools_web()   # universal web_fetch and web_search
tools_all()   # everything
```

Registration is not authorization: `tools_code()` makes the two trusted-code
tools available to the model, while the Agent's `Permissions` still decides
whether either call may run. `permissions_standard()` denies both.

Combine bundles by concatenating:

```{r, eval = FALSE}
tools <- c(tools_file(), tools_code())
agent <- Agent$new(chat = ellmer::chat_openai(model = "gpt-5.6-luna"), tools = tools)
```

## Tool Presets

Presets are named collections with opinionated defaults:

```{r, eval = FALSE}
tools_preset("minimal")
tools_preset("standard")
tools_preset("dev")
tools_preset("data")
tools_preset("full")
```

| Preset | Included capabilities |
|---|---|
| `minimal` | file and document reads |
| `standard` | reads plus native file writes; no code execution |
| `data` | data reads plus trusted R execution |
| `dev` | native files plus trusted R and shell execution |
| `full` | every built-in tool |

## Create a tool from an R function

A tool wraps an R function with a name, a description, and argument types that
the model can use. Start with a function you can call directly. This example
wraps `toupper()`; a package function or your own function works the same way.

```{r custom-tool, eval = TRUE}
library(deputy)
uppercase <- ellmer::tool(
  base::toupper,
  name = "uppercase",
  description = "Uppercase local text.",
  arguments = list(x = ellmer::type_string()),
  annotations = ellmer::tool_annotations(
    read_only_hint = TRUE,
    destructive_hint = FALSE,
    open_world_hint = FALSE,
    idempotent_hint = TRUE
  )
)
uppercase("hello")
```

Register the tool when constructing the agent:

```{r custom-agent, eval = FALSE}
agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-5.6-luna"),
  tools = list(uppercase),
  permissions = Permissions(file_write = FALSE)
)
result <- agent$run_sync("Use uppercase to convert 'hello' to capitals.")
result$response
```

The annotations describe the function's effects. Here the tool changes no
files, uses no network, and returns the same answer for the same input. Deputy
uses these declarations when checking permissions; it does not inspect the
function body to verify them.

| Annotation | Meaning when `TRUE` |
|---|---|
| `read_only_hint` | Does not change state |
| `destructive_hint` | May delete or overwrite data |
| `open_world_hint` | Accesses external systems |
| `idempotent_hint` | Repeated calls have the same effect |

### Replace or inspect registered tools

Use `agent$register_tool(tool)` to add a tool to an existing agent, or
`agent$register_tool(tool, replace = TRUE)` to replace a tool with the same name.

Existing names are errors unless `replace = TRUE` is explicit. Repeated names
inside one batch always fail, including with replacement enabled. List names
do not rename a tool. `set_tools()` replaces the whole user registry. Both
methods and construction validate and adapt the complete batch before changing
the backend registry; a malformed later tool leaves earlier tools untouched.
Skills use `load_skill(..., allow_conflicts = TRUE)` for explicit replacement.

Missing annotations stay absent on the source object. For custom tools,
permissions assume possible modification, destruction, and external access,
and no idempotence. An explicit read-only annotation makes an omitted
destructive annotation irrelevant. A local read tool therefore needs at least
`read_only_hint = TRUE` and `open_world_hint = FALSE` under standard permissions.
Readonly mode additionally requires an explicit allowlist entry for custom
tools. Annotations do not grant authority or verify a service's trustworthiness.
See `PermissionMode` for callback and mode behavior.

Inspect the supplied fields and defaults without executing the tool:

```{r tool-metadata, eval = TRUE}
tool_metadata(uppercase)
```

## Web Tools

deputy provides web tools that are **provider-aware**. Pass the chat to
`tools_web()` to select provider-native web tools when available. Calling it
without a chat returns Deputy's universal web tools:

```{r, eval = FALSE}
chat <- ellmer::chat_anthropic()
agent <- Agent$new(
  chat = chat,
  tools = tools_web(chat),
  permissions = Permissions(
    web = TRUE,
    tool_allowlist = c("web_search", "web_fetch")
  )
)
```

Provider-native tools execute outside R, so Deputy authorizes known native web
search and fetch tools before registering them. The policy must grant web
access and explicitly allow the tool; static denylists are also enforced at
registration. A custom `can_use_tool` callback rejects provider-native tools
because Deputy cannot supply request arguments or run context after provider
execution begins. Other provider-native tool types are rejected. Use the
universal function tools when authorization must inspect each request. If
`set_permission_mode()` later narrows away web access, Deputy removes registered
provider-native web tools before the new policy becomes active.

For providers without native support, deputy falls back to `tool_web_fetch`
(fetches a URL and extracts text) and `tool_web_search` (searches via
DuckDuckGo).

## MCP Integration

deputy can load tools from
[Model Context Protocol](https://modelcontextprotocol.io/) servers via
the [mcptools](https://github.com/posit-dev/mcptools) package:

```{r, eval = FALSE}
# Load tools from selected configured servers
agent <- Agent$new(
  chat = ellmer::chat_anthropic(),
  tools = c(tools_file(), tools_mcp(servers = "github"))
)
```

`tools_mcp()` returns an empty list with a warning when `mcptools` is not
installed or no tools are available. MCP servers are configured in
`~/.config/mcptools/config.json`; see the mcptools documentation for setup.

### Server metadata and reconnection

For service tools, `tools_mcp(config, servers = "evidence")` selects an exact
configured server name before connecting. The qualified mcptools 1.0.2 bridge
retains annotations from the server and records `source$type = "mcp"`, its
server name, and tool name. It reports missing metadata rather than inventing
safe effects. Other mcptools versions fail explicitly pending qualification.
The caller must trust the configured service; metadata does not establish trust.

Pass these objects to `Agent$new(tools = ...)`, `register_tools()`, or an
explicit tool registry used by `agent_definition_read()`. The same metadata
survives YAML selection, Agent wrapping, cloning, and delegation. Permission
callbacks receive it as `context$tool_metadata`, alongside the supplied
`context$tool_annotations`. A service called `read_file` still uses its service
annotations and capabilities; its name does not confer native file access.
Local file checkpointing excludes MCP calls, so rewinding files never treats a
remote service write as a change to the local workspace.

Reconnecting a server invalidates tools from its old connection. Reload and
replace explicitly with `agent$load_mcp(config, servers = "evidence", replace = TRUE)`.
This refresh removes obsolete tools from the selected servers, including when
discovery succeeds with an empty tool set. Unrelated tools remain registered.
A load failure records a failed attempt in `mcp_status()`. Tools that still
have working connections remain registered. If mcptools already closed an
old connection, its invalidated handles are removed even when discovery or
registration fails. Malformed metadata never silently becomes an unannotated
executable.

### Connections owned by a conversation

Use `McpConnection` when a host needs independent connections, asynchronous
calls, or resource and prompt access. Create the Agent first, then allow exact
items from one configured server:

```{r, eval = FALSE}
agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-5.6-luna"),
  permissions = Permissions(web = TRUE)
)
connection <- McpConnection$new(
  config = "~/.config/mcptools/config.json",
  server = "evidence",
  agent = agent,
  tools = "inspect_evidence",
  resources = "evidence://report/current",
  prompts = "summarize"
)
agent$register_tools(c(connection$tools(), connection$capability_tools()))
```

When one Agent uses multiple connections, choose distinct capability prefixes,
such as `first$capability_tools(prefix = "papers")` and
`second$capability_tools(prefix = "notes")`. These produce names such as
`papers_read_resource` and `notes_read_resource` without changing the allowlists.
Connections with no allowed tools can use resource-only or prompt-only servers;
they do not request a tool catalogue during initialization.

The allowlists are fixed. `$discover("resources")` returns a promise for one
catalogue page; its result may include a cursor for the next page. Discovery
does not fetch resource contents, retrieve prompts, register tools or expand
the Agent's permissions. Resource and prompt tools still require the Agent's
web capability. The host can use `$get_prompt(name, arguments)` for prompts
that require arguments; retrieving a prompt never inserts it into a Chat.

Each connection has its own mcptools registry. Two conversations can therefore
use the same configured server name without replacing each other's connection.
Tools carry the connection ID into permission callbacks and hooks. They reject
registration with a different Agent, session or run context. The host assigns
those identities and remains responsible for authentication.

Calls return promises and permit one active request per connection. Overlapping
requests fail with a busy error; other connections and the host event loop can
continue. Construction waits for startup. Call `$close()` when the conversation
ends, for example from a Shiny session's end callback. `$cancel()` terminates
the connection and discards its server session state. A timeout does the same;
old tool handles never start a replacement connection implicitly.

This is a temporary, version-checked mcptools 1.0.2 adapter. It uses upstream
transport, authentication, tool conversion and shutdown with a small internal
request bridge. Public replacements are requested in
[mcptools #129](https://github.com/posit-dev/mcptools/issues/129),
[#130](https://github.com/posit-dev/mcptools/issues/130), and
[#109](https://github.com/posit-dev/mcptools/issues/109).

### Sandboxed R with mcp-repl

Use `tools_mcp_repl()` when model-generated R needs an OS sandbox rather than
the current user's authority:

```{r, eval = FALSE}
repl_tools <- tools_mcp_repl(
  config = "~/.config/mcptools/config.json",
  server = "r",
  sandbox = "workspace-write"
)

agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-5.6-luna"),
  tools = repl_tools,
  permissions = Permissions(web = FALSE)
)
```

The helper isolates the named server from the rest of the MCP configuration
and checks its exact final `--sandbox` argument. Missing policies and modes that
do not establish a Deputy-verifiable boundary are errors. mcp-repl then owns
OS-specific confinement, including fail-closed startup on unsupported hosts.
See `vignette("permissions")` for the complete trust model.

## Human-in-the-Loop

The `tool_ask_user` tool lets an agent ask the user a question during
execution. This is useful for confirmations, clarifications, or collecting
input:

```{r, eval = FALSE}
agent_id <- "agent-analysis"
session_id <- "session-analysis"

agent <- Agent$new(
  chat = ellmer::chat_openai(model = "gpt-5.6-luna"),
  tools = c(
    tools_file(),
    tools_interactive(
      callback = function(questions, context) {
        # Route a modal to this Agent's host session and collect its answers.
        collect_answers(questions, route = context$session_id)
      },
      context = list(agent_id = agent_id, session_id = session_id)
    )
  ),
  agent_id = agent_id,
  session_id = session_id
)
```

`tools_interactive()` creates a fresh `tool_ask_user` instance. Its handler and
routing context belong to that tool, so concurrent Agents cannot overwrite one
another. The context may also be a zero-argument function when routing values
must be resolved for each request. Omit the callback in an interactive console
to use `readline()`.

`set_ask_user_callback()` remains a process-wide compatibility fallback for
single-Agent scripts. Do not use it to route Shiny or other concurrent hosts.
