---
title: "Tools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

A tool is an R function the model can ask to call. Deputy ships tools for
files, data, the web and code, and any R function becomes a tool once you
describe it with `ellmer::tool()`. Registering a tool only makes it available:
the agent's [permissions](permissions.html) decide whether each call runs.

## Built-in tools

| Tool | What it does | Effects |
|---|---|---|
| `tool_read_file` | Read a text file, or pages of a PDF | read-only |
| `tool_read_markdown` | Convert a PDF, Word, PowerPoint or HTML file to Markdown (needs the Python `markitdown` module) | read-only |
| `tool_list_files` | List a directory | read-only |
| `tool_glob_files` | Find files by glob pattern | read-only |
| `tool_grep_files` | Search file contents | read-only |
| `tool_read_csv` | Read a CSV file and summarise it | read-only |
| `tool_write_file` | Write or append to a file | destructive |
| `tool_edit_file` | Replace text in a file | destructive |
| `tool_multi_edit` | Make several replacements in one file | destructive |
| `tool_run_r_code` | Run R code in a new R process | destructive, open world |
| `tool_run_bash` | Run a shell command | destructive, open world |
| `tool_web_fetch` | Fetch a web page as text | read-only, open world |
| `tool_web_search` | Search the web with DuckDuckGo | read-only, open world |
| `tool_ask_user` | Ask the person running the agent a question | read-only |

`run_r_code` and `run_bash` are trusted-code tools: the code runs with your
user account's access to files and the network. The default permissions deny
both. [Running R code](code-execution.html) covers when to enable them and
how to sandbox model-written code instead.

## Bundles and presets

Bundles group related tools:

```{r}
library(deputy)

tools_file() # read_file, read_markdown, write_file, list_files
tools_data() # read_csv, read_file, read_markdown
tools_code() # run_r_code, run_bash
tools_web() # web_fetch, web_search
tools_all() # every built-in tool except ask_user
```

Combine them with `c()`:

```{r}
agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  tools = c(tools_file(), tools_web()),
  permissions = Permissions(web = TRUE)
)
```

Presets are named combinations for common jobs:

| Preset | Tools |
|---|---|
| `tools_preset("minimal")` | `read_file`, `read_markdown`, `list_files` |
| `tools_preset("standard")` | `minimal` plus `write_file` |
| `tools_preset("data")` | `minimal` plus `read_csv` and `run_r_code` |
| `tools_preset("dev")` | `standard` plus `run_r_code` and `run_bash` |
| `tools_preset("full")` | `tools_all()` |

The `data` and `dev` presets include trusted-code tools, so they only do
something useful with permissions that allow code (`r_code = TRUE`,
`bash = TRUE`).

## Turn an R function into a tool

A tool needs a function, a name, a description the model reads, and the types
of its arguments. This one wraps `toupper()`:

```{r custom-tool, eval = TRUE}
library(deputy)

uppercase <- ellmer::tool(
  base::toupper,
  name = "uppercase",
  description = "Convert text to upper case.",
  arguments = list(x = ellmer::type_string()),
  annotations = ellmer::tool_annotations(
    read_only_hint = TRUE,
    destructive_hint = FALSE,
    open_world_hint = FALSE,
    idempotent_hint = TRUE
  )
)
uppercase("hello")
```

Pass it to the agent like any other tool:

```{r custom-agent}
agent <- Agent$new(
  chat = ellmer::chat("openai/gpt-6-luna"),
  tools = list(uppercase)
)
agent$run_sync("Use the uppercase tool to shout 'hello'.")$response
```

The annotations tell Deputy what the function does. Permissions are decided
from them:

| Annotation | `TRUE` means |
|---|---|
| `read_only_hint` | the tool changes nothing |
| `destructive_hint` | the tool may delete or overwrite data |
| `open_world_hint` | the tool reaches outside the machine, for example the network |
| `idempotent_hint` | calling it twice has the same effect as calling it once |

Deputy takes annotations at their word; it doesn't inspect the function to
check them. When you leave one out, Deputy assumes the risky answer: the tool
might destroy data, might reach the network, and isn't safe to retry. A tool
that might reach the network then needs a policy with `web = TRUE`, and one
that might destroy data needs a policy that allows file writes or shell
commands. So declare all four.
In `"readonly"` mode a custom tool must also be listed in the policy's
`tool_allowlist`.

## Add, replace and inspect tools

Add tools to an existing agent with `$register_tool()` or `$register_tools()`.
Registering a name that already exists is an error unless you pass
`replace = TRUE`. `$set_tools()` replaces all tools at once. Deputy checks the
whole batch before changing anything, so one bad tool leaves the agent's tools
as they were.

```{r}
agent$register_tool(tool_list_files)
agent$register_tool(uppercase, replace = TRUE)
```

`tool_metadata()` shows where a tool came from, which annotations it declares
and which defaults Deputy will assume, without running it:

```{r tool-metadata, eval = TRUE}
tool_metadata(uppercase)
```

## Web tools

`tools_web()` returns Deputy's own `web_fetch` and `web_search`, which work with
any provider. Pass it your chat to use the provider's built-in web tools
instead, where there are any (Anthropic, Google, and OpenAI search):

```{r}
chat <- ellmer::chat("anthropic/claude-sonnet-5")
agent <- Agent$new(
  chat = chat,
  tools = tools_web(chat),
  permissions = Permissions(
    web = TRUE,
    tool_allowlist = c("web_search", "web_fetch")
  )
)
```

Provider tools run on the provider's servers, so Deputy can't check each call.
It checks them once, when they're registered: the policy must allow web access
and list them in `tool_allowlist`. A policy with a `can_use_tool` callback
can't use them at all, because the callback would never see their arguments.
Use `tools_web(chat, use_native = FALSE)` when you need per-call checks.

## Skills

A skill packages instructions, tools and the packages they need, so you can
give the same abilities to several agents. Create one in code:

```{r skill-values, eval = TRUE}
concise <- skill_create(
  "concise",
  prompt = "Answer in one short paragraph.",
  requires = list(packages = "base")
)
skill_check_requirements(concise)$ok
```

`agent$load_skill(concise)` appends the prompt to the agent's system prompt and
registers the skill's tools under the agent's current permissions. Missing
packages produce a warning, and a tool name that clashes with an existing tool
is an error unless you pass `allow_conflicts = TRUE`. `agent$skills()` lists
what's loaded.

Skills can also live in a directory with a `SKILL.md` prompt and a
`SKILL.yaml` that lists tools and requirements. Deputy ships two:
`data_analysis`, with exploratory analysis tools, and `debate`, a prompt for
weighing opposing arguments.

```{r}
agent$load_skill(system.file("skills", "data_analysis", package = "deputy"))
```

Loading a skill directory sources the R files its tools are defined in, so
load only skills you trust. `skills_list()` lists the skills in a directory.

## More tools

- [Running R code](code-execution.html): trusted R execution, persistent R
  sessions, and sandboxed interpreters through mcp-repl and MCP Console.
- [MCP servers](mcp.html): tools, resources and prompts from Model Context
  Protocol servers.
- [Human input and approvals](approvals.html): `tools_interactive()` lets the
  model ask a person questions.
